---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ13:** Question: Analyze how nonlinear RL mechanism in the BM
  algorithm affect learning dynamics, route adaptation,
  and convergence towards DUE states.

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

**The plots to answer this question will be:**

1. Mechanism plot
2. R-gap evolution across $\tau$
3. Flow evolution over episodes

For more details, check the research questions document.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(glue)
library(ggtext)
library(patchwork)
library(ggrepel)
library(forcats)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())
```

```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ13","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
data <- list.files(data_dir, pattern="\\.parquet$", full.names = TRUE) |> 
  (\(x) set_names(x, tools::file_path_sans_ext(basename(x))))() |> 
  map(read_parquet)
```

```{r}
#| label: get_warm_up_period(exploration)

# It's not perfectly precise — the true per-agent start is episode > 24 and all routes visited, so some agents start a few episodes later than 24. But it's the honest guaranteed floor ("no agent has started learning before this point"), it's already computed for every run via config.warm_up, and given the plot already treats each seed's trajectory independently without forcing alignment (per the existing code comment about seeds finishing at very different episode counts), a single shared floor line is consistent with that treatment rather than adding seed-specific precision the rest of the plot doesn't have either.

exploration_phase_floor <- as.numeric(unique(data$bm_rgap$warm_up))
```

## Plot 1 — mechanism

```{r}
#| label: tau_values

tau_values <- data$bm_rgap |> 
  mutate(
    stimulus_tau = as.numeric(stimulus_tau)
  ) |> 
  filter(
    !(stimulus_tau %in% c(0,0.1))
  ) |> 
  arrange(stimulus_tau) |> 
  select(stimulus_tau) |> 
  pull(stimulus_tau) |> 
  unique()

tau_labels <- glue("τ = {sprintf('%.2f',tau_values)}")
```

```{r}
#| label: deadzone_function

# Mirrors BMAgent._apply_stimulus_threshold (src/agents/agent.py) — the
# dead-zone + quadratic-ramp transform applied to each route's relative
# margin x_r = (AC - PC_r)/AC before max/min normalisation.
# g(x) = sign(x) * ( max(0, |x| - tau) / (1 - tau + epsilon) )^2
deadzone <- function(x, tau, epsilon = 1e-9) {
  magnitude <- pmax(0, abs(x) - tau) / (1 - tau + epsilon)
  sign(x) * magnitude^2
}
```

```{r}
#| label: data_plot1_lines

# Relative margins (x_r)
x_grid <- seq(-1, 1, length.out = 401)

# map_dfr: Applies the function to each element of a vector
# and combine results into one big df
plot1_lines_df <- map_dfr(tau_values, function(tau) {
  tibble(
    x = x_grid,
    g = deadzone(x_grid, tau),
    tau = tau
  )
}) |>
  mutate(
    tau_fct = factor(tau, levels = tau_values, labels = tau_labels)
  )
```

```{r}
#| label: label_info_plot1

label_info <- tibble(
  x = 0.5,
  g = 0.5,
  label = "Classical BM"
)
```


```{r}
#| label: plot_1

ggplot(
  data = plot1_lines_df,
  mapping = aes(x = x, y = g)
) +
  annotate(
  "segment",
  x = -1, y = -1,
  xend = 1, yend = 1,
  linetype = "dashed",
  linewidth = 0.5,
  color = "grey60"
) +
  geom_text_repel(
    data = label_info, 
    mapping = aes(x = x, y = g, label = label),
    size = 3.5,
    box.padding = 0.75,
    seed = 123,
    nudge_x = -0.2,
    nudge_y = 0.05,
    colour = "grey40",
    segment.size = 0.15
  ) +
  geom_line(linewidth = 0.5, mapping = aes(color = tau_fct)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey80") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey80") +
  scale_color_grey(
  start = 0.7,
  end = 0.1,
  name = expression(tau)
) +
  labs(
    title = "The dead-zone suppresses small margins and emphasizes large ones",
    subtitle = "Within the threshold, the response is zero; beyond it, the response grows quadratically.",
    x = expression("Relative cost margin " * x[r] * " = (AC - PC"[r] * ")/AC"),
    y = expression("Transformed margin " * g(x[r]))
  )

ggsave(here("r","RQ13","figures","plot1_nonlinear_mechanism.png"), width = 12, height = 6)
```

## Plot 2 — R-gap evolution across $\tau$

```{r}
#| label: representative_seeds
representative_seeds <- sample(unique(data$bm_rgap$seed), size = 3)
labels_seeds <- c("S1", "S2", "S3")
```

```{r}
#| label: data_plot2

representative_stimulus_tau_levels <- c(0.05, 0.2, 0.4)
representative_stimulus_tau_levels_labels <- map_chr(
  representative_stimulus_tau_levels,
  \(x) paste0("τ = ", x)
)
representative_memory_levels <- c(1.0,0.6,0.1)
representative_memory_levels_labels <- map_chr(representative_memory_levels, \(x) paste0("Memory = ", x))

rgap_df <- data$bm_rgap |> 
  mutate(
    stimulus_tau = as.numeric(stimulus_tau),
    memory_level = as.numeric(memory_level)
  ) |> 
  filter(
    stimulus_tau %in% representative_stimulus_tau_levels,
    memory_level %in% representative_memory_levels,
  ) |> 
  select(-c(run_id, warm_up)) |> 
  mutate(
    stimulus_tau_fct = 
      factor(stimulus_tau,
             levels = representative_stimulus_tau_levels,
             labels = representative_stimulus_tau_levels_labels
      ),
    memory_level_fct = fct_rev(factor(memory_level, levels = representative_memory_levels, labels = representative_memory_levels_labels))
  ) |> 
  select(-c(stimulus_tau, memory_level))

# Median and quantiles across seeds (here I always run the same episodes, 
# so i can compute summaries on seeds, instead of having to plot individual trajectories)
rgap_df_summary <- rgap_df |> 
  group_by(episode, stimulus_tau_fct, memory_level_fct) |> 
  summarise(
    median_rgap = median(rgap),
    q25_rgap    = quantile(rgap, 0.25),
    q75_rgap    = quantile(rgap, 0.75)
  ) |> 
  ungroup()
```

```{r}
#| label: breaks_plot2

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
step <- floor(max_episode / 30) * 10
breaks <- seq(0, max_episode, by = step)

# Remove breaks that are within step of the learning start
breaks <- breaks[abs(breaks - exploration_phase_floor) >= (step/2.25)]
# Add the learning start break
breaks <- sort(c(breaks, exploration_phase_floor))
labels <- as.character(breaks)
labels[breaks==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```

```{r}
#| label: plot_2

ggplot(
) +
  geom_line(
    data = rgap_df_summary,
    mapping = aes(x = episode, y = median_rgap),
  ) + 
  geom_ribbon(
    data = rgap_df_summary,
    mapping = aes(x = episode, ymin = q25_rgap, ymax = q75_rgap),
    alpha = 0.1
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.text.x = element_text(size = 9),
    axis.text.y = element_text(size = 9),
    strip.text = element_text(size = 10)
) +
  labs(
    x = "Episode",
    y = "R-gap (%)",
    title = "Nonlinear learning affects R-gap convergence",
    subtitle = "Median R-gap across seeds; ribbons show variability"
  ) +
  scale_x_continuous(breaks = breaks, labels = labels) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = "red",
    alpha = 0.08
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.5
  ) +
  geom_hline(yintercept = 0, alpha = 0.6, linewidth = 0.7, color = "#7D7D7D") +
  geom_hline(
    yintercept = 5,
    colour = "grey40",
    linetype = "longdash",
    linewidth = 0.5
  ) +
  # Automatically chosen breaks and add one extra tick
  scale_y_continuous(
    breaks = function(x) sort(unique(c(scales::breaks_pretty()(x), 5)))
) +
  facet_grid(memory_level_fct ~ stimulus_tau_fct)

ggsave(here("r","RQ13","figures","plot2_rgap_evolution.png"), width = 12, height = 6)
```

########
# PLOT 3
########

```{r}
#| label: random_seed
# Randomly select 1 seed 
set.seed(23)
random_seed <- sample(unique(data$bm_rgap$seed), size = 1)
```

```{r}
#| label: representative_OD

# Choose an OD whose shortest route is not the most stable (hence, there exist alternative routes that are longer but more stable).
# Is gonna be chosen taking into account equally weighted means (without memory mechanism) and with theta=0 (under normal drivers scenario). 
# We do it this way because we merely wanna inspect the network/demand structure to choose our OD and isolate it from the effect of other variables

# We wamt that the shortest route is at least 5 seconds faster than the most stable route
# (this way we could see a difference in the r-gap when implemented risk sensitivity model)

rewards_df <- data$rewards |> 
  mutate(
    stimulus_tau = as.numeric(stimulus_tau),
    memory_level = as.numeric(memory_level)
  ) |> 
  filter(
    seed == random_seed,
    stimulus_tau == 0,
    memory_level == 1
  ) |> 
  select(-c(run_id, seed, warm_up, stimulus_tau, memory_level))

actions_df <- data$actions |>
  mutate(
    stimulus_tau = as.numeric(stimulus_tau),
    memory_level = as.numeric(memory_level)
  ) |> 
  filter(
    seed == random_seed,
    stimulus_tau == 0,
    memory_level == 1
  ) |> 
  select(-c(run_id, seed, warm_up, stimulus_tau, memory_level))

agents_od_df <- data$agents_od |> 
  mutate(
    stimulus_tau = as.numeric(stimulus_tau),
    memory_level = as.numeric(memory_level)
  ) |> 
  filter(
    seed == random_seed,
    stimulus_tau == 0,
    memory_level == 1
  ) |> 
  select(-c(run_id, seed, warm_up, stimulus_tau, memory_level, post_warm_up, departure_time)) |> 
  rename(agent_id = id)

representative_od_df <- rewards_df |> 
  left_join(actions_df, by = c("episode", "agent_id")) |>
  left_join(agents_od_df, by = "agent_id")

representative_od_df_summary <- representative_od_df |> 
  group_by(origin, destination, action) |> 
  summarise(
    mean_tt = mean(reward),
    sd_tt = sd(reward),
    count = n()
  ) |> 
  ungroup() |> 
  filter(
    count > 15
  )

representative_od_decision_df <- 
  representative_od_df_summary |> 
  group_by(
    origin, destination
  ) |> 
  mutate(
    min_mean_tt = round(min(mean_tt),2),
    distance_to_min = round((mean_tt - min_mean_tt) / min_mean_tt,2)
  ) 

representative_od_tmp <- representative_od_decision_df |> 
  summarise(
    below_10 = sum(distance_to_min < 0.10 & distance_to_min > 0.0),
    between_10_20 = sum(distance_to_min < 0.20 & distance_to_min > 0.1),
    above_40 = sum(distance_to_min > 0.4)
  ) |> 
  filter(
    below_10 > 0,
    between_10_20 > 0,
    above_40 > 0
  ) |> 
  ungroup() 

representative_od <- representative_od_tmp |>
  slice_tail(n = 1) |> 
  select(origin, destination)

representative_od_routes <- representative_od_df_summary |> 
  filter(
    origin == representative_od$origin,
    destination == representative_od$destination
  ) |> 
  arrange(mean_tt)

agents_representative_od <- agents_od_df |> 
  filter(
    origin == representative_od$origin,
    destination == representative_od$destination
  )
```

```{r}
#| label: data_plot3_qualitative
# Things to take into account:
# 1. OD matrix is different across seeds (keeping everything else constant)
# 2. OD matrix is the same across stimulus tau values (keeping everything else constant)
# 3. OD matrix is the same across memory levels (keeping everything else constant)

# When analyzing the impact of stimulus tau on the route flow evolution, we just want to analyze one OD from a FIXED OD MATRIX!!!
###########################
# We already choose our representative OD before
# #########################
# 1. representative_seed
representative_seed <- random_seed
# 2. representative_stimulus_tau
representative_stimulus_tau_levels
# 3. representative_od
representative_OD <- data$demand_odt |> filter(
  seed == representative_seed,
  origin == representative_od$origin,
  destination == representative_od$destination
) |> 
  mutate(
    stimulus_tau = as.numeric(stimulus_tau),
    memory_level = as.numeric(memory_level)
  ) |> 
  filter(
    stimulus_tau == min(representative_stimulus_tau_levels),
    memory_level == min(representative_memory_levels),
  ) |> 
  select(-c(run_id, seed, warm_up, stimulus_tau, memory_level)) |> 
  group_by(origin, destination) |> 
  summarise(
    demand = sum(count)
  ) |> 
  ungroup() |> 
  rename(
    dest = destination
  )
```

```{r}
#| label: data2_plot3_qualitative

# Now that we have the scenario (defined by the seed) 
# and we also have the representative OD whose paths we want to analyze, 
# we need to do the following:

# 1. Compute the flow on each route of the representative OD per episode
# Efficiently dealing with large dataframes

# 1. Partition parquet file
pq_path_flow_paths <- here("r/RQ13/data/flow_paths_partitioned")

if (dir.exists(pq_path_flow_paths)) {
  unlink(pq_path_flow_paths, recursive = TRUE, force = TRUE)
}

dir.create(pq_path_flow_paths, recursive = TRUE)

data$flow_paths |> 
  group_by(path) |> 
  write_dataset(path = pq_path_flow_paths, format = "parquet")

# 2. List partitioned parquet files
tibble(
  files = list.files(pq_path_flow_paths, recursive = TRUE),
  size_MB = file.size(file.path(pq_path_flow_paths, files)) / 1024^2
)

# 3. Use dplyr with arrow
flow_paths_pq <- open_dataset(pq_path_flow_paths)
query <- flow_paths_pq |> 
  mutate(
    stimulus_tau = as.numeric(stimulus_tau),
    memory_level = as.numeric(memory_level)
  ) |> 
  filter(
    # Filter by the particular scenario
    seed == representative_seed, 
    stimulus_tau %in% representative_stimulus_tau_levels,
    memory_level %in% representative_memory_levels,
    
    # Filter by the particular OD
    origin == representative_OD$origin,
    destination == representative_OD$dest,
  ) |> 
  group_by(
    episode, path, stimulus_tau, memory_level
  ) |> 
  summarise(
    flow = sum(count)
  ) |> 
  ungroup()

flow_paths_df_query <- query |> collect()

# Check: The sum of flows across paths for the particular OD for one episode should be equal to the demand on that OD
check_demand_od_df <- flow_paths_df_query |> 
  group_by(episode, stimulus_tau, memory_level) |> 
  summarise(
    demand = sum(flow)
  ) |> ungroup()

demand_od_sum_paths <- unique(check_demand_od_df$demand)
demand_od <- representative_OD$demand

if (demand_od != demand_od_sum_paths) {
  stop("Something is wrong in the computation of path flows for the particular OD and scenario.")
}
```

```{r}
#| label: data_plot_3_part_3

# Now we still have to manipulate a bit the flow paths on the particular OD and scenario, so that is properly visualized in the plot

# 1. Handling some things to convert path variable to factor
path_levels <- as.character(sort(unique(flow_paths_df_query$path)))
path_levels_labels <- as.character(sort(unique(flow_paths_df_query$path + 1)))

flow_paths_df <- flow_paths_df_query |> 
  mutate(
    # Convert to factors
    stimulus_tau_fct = factor(
      stimulus_tau,
      levels = representative_stimulus_tau_levels,
      labels = representative_stimulus_tau_levels_labels
    ),
    memory_level_fct = factor(
      memory_level,
      levels = representative_memory_levels,
      labels = representative_memory_levels_labels
    ),
    path = factor(
      path, 
      levels = path_levels,
      labels = path_levels_labels
    ),
    share = flow / representative_OD$demand,
    memory_level_fct = fct_rev(memory_level_fct)
  ) 
```

```{r}
#| label: breaks_plot3
#| eval: False

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
step_plot4 <- floor(max_episode / 50) * 10
breaks_plot4 <- seq(0, max_episode, by = step_plot4)

# Remove breaks that are within step of the learning start
breaks_plot4 <- breaks[abs(breaks - exploration_phase_floor) >= step_plot4 / 2]
# Add the learning start break
breaks_plot4 <- sort(c(breaks_plot4, exploration_phase_floor))
labels_plot4 <- as.character(breaks_plot4)
labels_plot4[breaks_plot4==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```


```{r}
#| label: sidebar_table_n_eff

# Median effective number of routes (n_eff = exp(entropy of p)) per agent
# on the representative OD, at the final episode, for each (tau, memory)
# combo — annotated per facet on plot3 below. Restated here from bm_results
# (already exported: p per agent/episode/route), not a new data pull.
bm_results_df <- data$bm_results |>
  mutate(
    stimulus_tau = as.numeric(stimulus_tau),
    memory_level = as.numeric(memory_level)
  ) |>
  filter(
    episode == max(episode),
    seed == representative_seed,
    stimulus_tau %in% representative_stimulus_tau_levels,
    memory_level %in% representative_memory_levels,
    agent_id %in% agents_representative_od$agent_id
  )

n_eff_routes_df <- bm_results_df |>
  select(agent_id, memory_level, stimulus_tau, route_id, p) |>
  mutate(
    term_i = ifelse(p > 0, p * log(p), 0)
  ) |>
  group_by(agent_id, memory_level, stimulus_tau) |>
  summarise(
    entropy = -sum(term_i),
    .groups = "drop"
  ) |>
  mutate(
    n_eff_routes = exp(entropy)
  )

n_eff_annotation_df <- n_eff_routes_df |>
  group_by(memory_level, stimulus_tau) |>
  summarise(
    mean_n_eff = mean(n_eff_routes),
    .groups = "drop"
  ) 
```

```{r}
#| label: plot3

# This plots refers to a particular OD, for a particular seed, and facet for stimulus_tau and memory_levels
# Uniform, equal-weight color treatment (same 5-color legend, unchanged across every panel) is the right choice here specifically because the plot's job is to show a hierarchy dissolving, not to guide attention within a hierarchy that's assumed stable throughout.

ggplot(
  data = flow_paths_df,
  mapping = aes(x = episode, y = share, group = path, color = path)
) +
  geom_line() +
  facet_grid(stimulus_tau_fct ~ memory_level_fct) +
  labs(
    color = "Route",
    y = "Route flow",
    x = "Episode",
  ) +
  # Shaded region
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = "red",
    alpha = 0.06
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.4
  ) +
  # Change color scale
  scale_color_brewer(palette = "Dark2", name = "Path") +
  # Tweak miscellaneous stuff
  theme(
    legend.text  = element_text(size = 10),
    legend.key.size = unit(0.6, "lines"),
    plot.subtitle = element_text(size = 11),
    strip.text.y = element_text(angle = 0, size = 11),
    axis.text.x = element_text(size = 9),
    axis.text.y = element_text(size = 9),
) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(breaks = breaks, labels = labels) +
  labs(
    y = "Route share (%)",
    title = "As τ grows, agents become less sensitive to small route differences",
    subtitle = "Route share over episodes for a representative OD, faceted by τ and memory level"
  ) 

ggsave(here("r","RQ13","figures","plot3_qualitative_route_flow_evolution.png"), width = 12, height = 6)
```

```{r}
#| label: sidebar_latex_table

# QUANTITATIVE TABLE
representative_od_decision_df |> 
  filter(
    origin == representative_od$origin,
    destination == representative_od$destination
  ) |> 
  mutate(
    route_id = action + 1
  ) |> 
  select(origin, destination, route_id, mean_tt, distance_to_min)
```

#####
APPENDIX
#####

```{r}
#| label: plot2_linear

representative_stimulus_tau_levels <- 0
representative_stimulus_tau_levels_labels <- "Classic BM - Linear"
```

```{r}
#| label: data_plot2

rgap_df <- data$bm_rgap |> 
  mutate(
    stimulus_tau = as.numeric(stimulus_tau),
    memory_level = as.numeric(memory_level)
  ) |> 
  filter(
    stimulus_tau %in% representative_stimulus_tau_levels,
    memory_level %in% representative_memory_levels,
  ) |> 
  select(-c(run_id, warm_up)) |> 
  mutate(
    stimulus_tau_fct = 
      factor(stimulus_tau,
             levels = representative_stimulus_tau_levels,
             labels = representative_stimulus_tau_levels_labels
      ),
    memory_level_fct = fct_rev(factor(memory_level, levels = representative_memory_levels, labels = representative_memory_levels_labels))
  ) |> 
  select(-c(stimulus_tau, memory_level))

# Median and quantiles across seeds (here I always run the same episodes, 
# so i can compute summaries on seeds, instead of having to plot individual trajectories)
rgap_df_summary <- rgap_df |> 
  group_by(episode, stimulus_tau_fct, memory_level_fct) |> 
  summarise(
    median_rgap = median(rgap),
    q25_rgap    = quantile(rgap, 0.25),
    q75_rgap    = quantile(rgap, 0.75)
  ) |> 
  ungroup()
```

```{r}
#| label: breaks_plot2

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
step <- floor(max_episode / 30) * 10
breaks <- seq(0, max_episode, by = step)

# Remove breaks that are within step of the learning start
breaks <- breaks[abs(breaks - exploration_phase_floor) >= (step/2.25)]
# Add the learning start break
breaks <- sort(c(breaks, exploration_phase_floor))
labels <- as.character(breaks)
labels[breaks==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```

```{r}
#| label: plot_2

ggplot(
) +
  geom_line(
    data = rgap_df_summary,
    mapping = aes(x = episode, y = median_rgap),
  ) + 
  geom_ribbon(
    data = rgap_df_summary,
    mapping = aes(x = episode, ymin = q25_rgap, ymax = q75_rgap),
    alpha = 0.1
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.text.x = element_text(size = 9),
    axis.text.y = element_text(size = 9),
    strip.text = element_text(size = 10)
) +
  labs(
    x = "Episode",
    y = "R-gap (%)",
    title = "R-gap evolution - Classical Linear BM",
    subtitle = "Median R-gap across seeds; ribbons show variability"
  ) +
  scale_x_continuous(breaks = breaks, labels = labels) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = "red",
    alpha = 0.08
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.5
  ) +
  geom_hline(yintercept = 0, alpha = 0.6, linewidth = 0.7, color = "#7D7D7D") +
  geom_hline(
    yintercept = 5,
    colour = "grey40",
    linetype = "longdash",
    linewidth = 0.5
  ) +
  # Automatically chosen breaks and add one extra tick
  scale_y_continuous(
    breaks = function(x) sort(unique(c(scales::breaks_pretty()(x), 5)))
) +
  facet_grid(memory_level_fct ~ stimulus_tau_fct)

ggsave(here("r","RQ13","figures","appendix_plot2_rgap_evolution.png"), width = 12, height = 6)
```

# PLOT 3

```{r}
#| label: data_plot3_qualitative
# Things to take into account:
# 1. OD matrix is different across seeds (keeping everything else constant)
# 2. OD matrix is the same across stimulus tau values (keeping everything else constant)
# 3. OD matrix is the same across memory levels (keeping everything else constant)

# When analyzing the impact of stimulus tau on the route flow evolution, we just want to analyze one OD from a FIXED OD MATRIX!!!
###########################
# We already choose our representative OD before
# #########################
# 1. representative_seed
representative_seed <- random_seed
# 2. representative_stimulus_tau
representative_stimulus_tau_levels
# 3. representative_od
representative_OD <- data$demand_odt |> filter(
  seed == representative_seed,
  origin == representative_od$origin,
  destination == representative_od$destination
) |> 
  mutate(
    stimulus_tau = as.numeric(stimulus_tau),
    memory_level = as.numeric(memory_level)
  ) |> 
  filter(
    stimulus_tau == min(representative_stimulus_tau_levels),
    memory_level == min(representative_memory_levels),
  ) |> 
  select(-c(run_id, seed, warm_up, stimulus_tau, memory_level)) |> 
  group_by(origin, destination) |> 
  summarise(
    demand = sum(count)
  ) |> 
  ungroup() |> 
  rename(
    dest = destination
  )
```

```{r}
#| label: data2_plot3_qualitative

# Now that we have the scenario (defined by the seed) 
# and we also have the representative OD whose paths we want to analyze, 
# we need to do the following:

# 1. Compute the flow on each route of the representative OD per episode
# Efficiently dealing with large dataframes

# 3. Use dplyr with arrow
flow_paths_pq <- open_dataset(pq_path_flow_paths)
query <- flow_paths_pq |> 
  mutate(
    stimulus_tau = as.numeric(stimulus_tau),
    memory_level = as.numeric(memory_level)
  ) |> 
  filter(
    # Filter by the particular scenario
    seed == representative_seed, 
    stimulus_tau %in% representative_stimulus_tau_levels,
    memory_level %in% representative_memory_levels,
    
    # Filter by the particular OD
    origin == representative_OD$origin,
    destination == representative_OD$dest,
  ) |> 
  group_by(
    episode, path, stimulus_tau, memory_level
  ) |> 
  summarise(
    flow = sum(count)
  ) |> 
  ungroup()

flow_paths_df_query <- query |> collect()

# Check: The sum of flows across paths for the particular OD for one episode should be equal to the demand on that OD
check_demand_od_df <- flow_paths_df_query |> 
  group_by(episode, stimulus_tau, memory_level) |> 
  summarise(
    demand = sum(flow)
  ) |> ungroup()

demand_od_sum_paths <- unique(check_demand_od_df$demand)
demand_od <- representative_OD$demand

if (demand_od != demand_od_sum_paths) {
  stop("Something is wrong in the computation of path flows for the particular OD and scenario.")
}
```

```{r}
#| label: data_plot_3_part_3

# Now we still have to manipulate a bit the flow paths on the particular OD and scenario, so that is properly visualized in the plot

# 1. Handling some things to convert path variable to factor
path_levels <- as.character(sort(unique(flow_paths_df_query$path)))
path_levels_labels <- as.character(sort(unique(flow_paths_df_query$path + 1)))

flow_paths_df <- flow_paths_df_query |> 
  mutate(
    # Convert to factors
    stimulus_tau_fct = factor(
      stimulus_tau,
      levels = representative_stimulus_tau_levels,
      labels = representative_stimulus_tau_levels_labels
    ),
    memory_level_fct = factor(
      memory_level,
      levels = representative_memory_levels,
      labels = representative_memory_levels_labels
    ),
    path = factor(
      path, 
      levels = path_levels,
      labels = path_levels_labels
    ),
    share = flow / representative_OD$demand,
    memory_level_fct = fct_rev(memory_level_fct)
  ) 
```

```{r}
#| label: breaks_plot3
#| eval: False

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
step_plot4 <- floor(max_episode / 50) * 10
breaks_plot4 <- seq(0, max_episode, by = step_plot4)

# Remove breaks that are within step of the learning start
breaks_plot4 <- breaks[abs(breaks - exploration_phase_floor) >= step_plot4 / 2]
# Add the learning start break
breaks_plot4 <- sort(c(breaks_plot4, exploration_phase_floor))
labels_plot4 <- as.character(breaks_plot4)
labels_plot4[breaks_plot4==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```


```{r}
#| label: sidebar_table_n_eff

# Median effective number of routes (n_eff = exp(entropy of p)) per agent
# on the representative OD, at the final episode, for each (tau, memory)
# combo — annotated per facet on plot3 below. Restated here from bm_results
# (already exported: p per agent/episode/route), not a new data pull.
bm_results_df <- data$bm_results |>
  mutate(
    stimulus_tau = as.numeric(stimulus_tau),
    memory_level = as.numeric(memory_level)
  ) |>
  filter(
    # episode == max(episode),
    seed == representative_seed,
    stimulus_tau %in% representative_stimulus_tau_levels,
    memory_level %in% representative_memory_levels,
    agent_id %in% agents_representative_od$agent_id
  )

n_eff_routes_df <- bm_results_df |>
  select(agent_id, memory_level, stimulus_tau, route_id, p) |>
  mutate(
    term_i = ifelse(p > 0, p * log(p), 0)
  ) |>
  group_by(agent_id, memory_level, stimulus_tau) |>
  summarise(
    entropy = -sum(term_i),
    .groups = "drop"
  ) |>
  mutate(
    n_eff_routes = exp(entropy)
  )

n_eff_annotation_df <- n_eff_routes_df |>
  group_by(memory_level, stimulus_tau) |>
  summarise(
    mean_n_eff = mean(n_eff_routes),
    .groups = "drop"
  ) 
```

```{r}
#| label: plot3

# This plots refers to a particular OD, for a particular seed, and facet for stimulus_tau and memory_levels
# Uniform, equal-weight color treatment (same 5-color legend, unchanged across every panel) is the right choice here specifically because the plot's job is to show a hierarchy dissolving, not to guide attention within a hierarchy that's assumed stable throughout.

ggplot(
  data = flow_paths_df,
  mapping = aes(x = episode, y = share, group = path, color = path)
) +
  geom_line() +
  facet_grid(. ~ memory_level_fct) +
  labs(
    color = "Route",
    y = "Route flow",
    x = "Episode",
  ) +
  # Shaded region
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = "red",
    alpha = 0.06
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.4
  ) +
  # Change color scale
  scale_color_brewer(palette = "Dark2", name = "Path") +
  # Tweak miscellaneous stuff
  theme(
    legend.text  = element_text(size = 10),
    legend.key.size = unit(0.6, "lines"),
    plot.subtitle = element_text(size = 11),
    strip.text.y = element_text(angle = 0, size = 11),
    axis.text.x = element_text(size = 9),
    axis.text.y = element_text(size = 9),
) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(breaks = breaks, labels = labels) +
  labs(
    y = "Route share (%)",
    title = "Route flow evolution - Classical Linear BM",
    subtitle = "Route share over episodes for a representative OD across memory levels"
  ) 

ggsave(here("r","RQ13","figures","appendix_plot3_qualitative_route_flow_evolution.png"), width = 12, height = 6)
```



