---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ1:** Does Bush-Mosteller multi-agent RL converge to a Dynamic User Equilibrium
(DUE) in a microscopic SUMO traffic simulation?

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

**The plots to answer this question will be:**

1. Aggregate R-gap evolution (main convergence evidence).
2. Heatmap R-gap by departure interval (all episodes)
3. Worst-case (90th percentile) OD-specific R-gap by seed

For more details, check the research questions document.



```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
congestion_regime_tiers <- c("Near free-flow", "Moderate", "High")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```


```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ1","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
data <- list.files(data_dir, pattern="\\.parquet$", full.names = TRUE) |> 
  (\(x) set_names(x, tools::file_path_sans_ext(basename(x))))() |> 
  map(read_parquet)
```

```{r}
#| label: representative_seeds_spaghetti_plots

# Randomly select 3 seeds (seeded so the same 3 - and their colors - are
# picked every time this re-renders)
set.seed(123)
representative_seeds <- sample(unique(data$bm_rgap$seed), size = 3)

labels_seeds <- c("S1", "S2", "S3")

color_lines <- c(
  "S1" = "#0072B2",  # blue
  "S2" = "#E69F00",  # orange
  "S3" = "#009E73"   # bluish green
  )  
```

```{r}
#| label: get_warm_up_period(exploration)

# It's not perfectly precise — the true per-agent start is episode > 24 and all routes visited, so some agents start a few episodes later than 24. But it's the honest guaranteed floor ("no agent has started learning before this point"), it's already computed for every run via config.warm_up, and given the plot already treats each seed's trajectory independently without forcing alignment (per the existing code comment about seeds finishing at very different episode counts), a single shared floor line is consistent with that treatment rather than adding seed-specific precision the rest of the plot doesn't have either.

exploration_phase_floor <- as.numeric(unique(data$bm_rgap$warm_up))

```



```{r}
#| label: data_plot1

# Not aggregating mean +/- SD across seeds here: seeds finish training at very
# different episode counts (e.g. High regime: some stop by episode ~22, one
# runs to 129), so for a large share of each panel's episode range the "mean"
# would silently collapse to a single seed's raw value (SD undefined). Each
# seed's own trajectory is plotted instead - no cross-seed averaging, so no
# misleading shrinking-N problem.

rgap_df <- data$bm_rgap |>
  mutate(
    congestion_regime = make_congestion_regime(n_agents),
    ) |>
  filter(
    seed %in% representative_seeds
  ) |> 
  mutate(
    seed_fct = factor(seed, levels = representative_seeds, labels = labels_seeds)
  )

plot1_annotation <- rgap_df |> 
  group_by(seed, n_agents) |> 
  slice_max(episode, n = 1, with_ties = FALSE)
```

```{r}
#| label: plot1

ggplot(
  data = rgap_df,
  mapping = aes(x = episode, y = rgap, color = seed_fct, group = seed_fct)
) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = "red",
    alpha = 0.08
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.5
  ) +
  annotate(
    "text",
    x = (exploration_phase_floor - 2) / 2,
    y = Inf,
    label = "pre-learning \nphase",
    fontface = "bold",
    colour = "firebrick",
    vjust = 2.5,
    size = 3.5
  ) +
  geom_line() +
  geom_point(data = plot1_annotation, size = 1.25) +
  geom_hline(yintercept = 0, alpha = 0.6, linewidth = 0.7, color = "#7D7D7D") +
  facet_grid(. ~ congestion_regime, scales = "free_x", space = "free_x") +
  scale_color_manual(name = "Seed", values = color_lines) +
  labs(
    title = "R-gap evolution across episodes",
    subtitle = "Training trajectories for three random seeds (BM)",
    x = "Episode",
    y = "R-gap (%)"
  ) +
  facetted_pos_scales(
    x = list(
      scale_x_continuous(breaks = scales::breaks_width(10)),
      scale_x_continuous(breaks = scales::breaks_width(10)),
      scale_x_continuous(breaks = scales::breaks_width(10))
    )
  ) +
  scale_y_continuous(breaks = scales::breaks_width(5))

ggsave(here("r","RQ1","figures","plot1_rgap_evolution.png"), width = 12, height = 6)
```

## Interpretation plot 1: (R-gap evolution across episodes)
It should be noted, that the first episodes (for this particular case the first 24 episodes) 
are a warm-up in which the agents only explore and do not learn. 
Learning phase begins at episode 25.
- **Convergence speed:**
  - Under Near free flow (1300), all seeds converge quickly (around episodes 30-40)
  - Under Moderate (1550), convergence is slightly slower.
  - Under High (2000), convergence is slower, around episodes 40-50, and much noisier (it has pretty large spikes)
- **Stability:**
  - Post-convergence stability: The trajectories show post-convergence stability nicely (once it converges to DUE, it remains in DUE for the rest of episodes).
  - Convergence dynamics (rate of convergence once convergence begins): In the free-flow case it reaches DUE almost immediately. Moderate proceeds more gradually. High congestion also proceeds more gradually, and it has more fluctuations before settling.
- **Robustness across seeds**
  - Free-flow: Almost identical trajectories.
  - Moderate: Pretty similar, but they show more variability. (the learning trajectories).
  - High congestion: The scenario in which the trajectories show the biggest differences. One of the seeds have large spikes.
  So the simulation is not very sensitive to the seed, in either the near-free flow or moderate scenario, but it shows more sensitivity to the sensitivity on the high congestion scenario.
- **Effect of congestion**:
  - It can be easily seen that higher congestion makes the problem harder, resulting in higher initial r-gaps, slower convergence, and more variability across seeds.

```{r}
#| label: data_plot2

# Not aggregating (mean) across seeds here, same reasoning as plot1: seeds
# finish training at very different episode counts, so a per-episode mean
# across seeds would silently collapse to a shrinking number of contributing
# seeds. Each of the 3 representative seeds is plotted as its own heatmap
# instead (rows), with congestion regime as columns - full seed set is still
# used for any statistical summaries elsewhere, just not for this plot.
#
# duaIterate is excluded (as before): its refined R-gap was only ever
# computed for the final iteration, not per-episode, so it has nothing to
# show along an episode x-axis.

refined_rgap_df <- data$bm_refined_rgap |>
  mutate(
    congestion_regime = make_congestion_regime(n_agents)
  ) |>
  filter(
    seed %in% representative_seeds
  ) |>
  mutate(
    seed_fct = factor(seed, levels = representative_seeds, labels = labels_seeds)
  )

cocient <- min(data$bm_refined_rgap$time_interval)

refined_rgap_df_agg <- refined_rgap_df |> 
  mutate(
    interval_group = time_interval %/% cocient
  ) |> 
  mutate(
    interval_group = factor(
      interval_group,
      levels = 1:max(interval_group),
      labels = paste0(
        seq(10, by = 10, length.out = max(interval_group)),
        "—",
        seq(19, by = 10, length.out = max(interval_group))
      )
      )
  ) |> 
  group_by(interval_group, seed_fct, congestion_regime, episode) |> 
  summarise(median_refined_rgap=median(refined_rgap)) |> 
  ungroup()

refined_rgap_df_agg
```

```{r}
#| label: plot2

upper_limit_color_scale <- quantile(refined_rgap_df_agg$median_refined_rgap, probs = c(0.95))

ggplot(
  data = refined_rgap_df_agg,
  mapping = aes(x = episode, y = interval_group, fill = median_refined_rgap)
) +
  geom_tile() +
  scale_y_discrete(limits = rev) +
  scale_fill_viridis_c(
    limits = c(0, upper_limit_color_scale),
    oob = scales::squish,
    option = "C"
  ) +
  facet_grid(seed_fct ~ congestion_regime, scales = "free_x", space = "free_x") +
  labs(
    title = "Refined R-gap by departure interval across episodes",
    subtitle = "Departure intervals were aggregated into 10-interval groups \n(median refined R-gap per group).",
    x = "Episode",
    y = "Departure interval (group)",
    fill = "Refined\nR-gap (%)"
  ) + 
  geom_vline(xintercept = exploration_phase_floor, colour = "black", linewidth = 1.2)   +
  geom_vline(xintercept = exploration_phase_floor, colour = "white", linewidth = 0.5,  linetype = "dotdash") 

ggsave(here("r","RQ1","figures","plot2_refined_rgap_heatmap.png"), width = 12, height = 10)
```

## Interpretation plot 2:
It should be noted, that the first episodes (for this particular case the first 24 episodes) 
are a warm-up in which the agents only explore and do not learn. 
Learning phase begins at episode 25.
**Which intervals converge first?**
Initially we could expect to see a strong diagonal pattern, indicating that later intervals take longer to converge, and that early intervals are the first ones to converge. 
But, that is not the case with our plot, suggesting that most intervals improve at roughly the same time.
That is the case for the 3 congestion regimes.

**Effect of congestion regimes**
It can be seen that higher congestion makes the problem harder, resulting in intervals with higher r-gap, represented with brighter bands.

**Some intervals become unstable after initially converging?**
In the moderate case, we can see on the seed 3, that some intervals become unstable after initially converging.
In the high congestion case, it happens more times. For example, a few intervals on the seed 2 and on the seed 3 become unstable after converging. On the seed 1, it is harder to visualize, but there are also some intervals that become unstable after converging.

**Late intervals consistently have large errors?**
It can be easily visualized on the plot, that is not our case.

**All intervals converge to the same r-gap?**
In order to answer this question, we have to look at the colors of the band for the last episodes.

In the near-free flow case most intervals achieve a very low r-gap, and hence it can be said that all intervals converge to the same r-gap.

In the moderate case, most intervals achieve a very low r-gap. Except for the seed 3, where some intervals remain with higher r-gaps.

In the high congestion case the r-gap of the final episode shows greater variability among intervals.

The greater the congestion, the greater the variability of the final r-gap among intervals.




```{r}
#| label: data_plot3

# Plot3: worst-case (90th percentile) OD-specific r-gap, one value per
# algorithm x seed, computed across that seed's own final episode. 

rgap_od_df <- bind_rows(
  mutate(data$bm_rgap_by_od, algorithm = "BM"),
  mutate(data$dua_rgap_by_od, algorithm = "duaIterate")
) |>
  mutate(congestion_regime = make_congestion_regime(n_agents))

rgap_od_seed_summary <- rgap_od_df |>
  group_by(algorithm, congestion_regime, n_agents, seed) |>
  filter(episode == max(episode)) |>
  summarise(
    p90_rgap = quantile(rgap, 0.9),
    .groups = "drop"
  )
```

```{r}
#| label: plot3

ggplot(
  data = rgap_od_seed_summary,
  mapping = aes(x = algorithm, y = p90_rgap, fill = algorithm)
) +
  geom_boxplot(outlier.shape = NA, width = 0.5, linewidth = 0.6, alpha = 0.6) +
  geom_jitter(width = 0.1, height = 0, size = 2.5, shape = 21, color = "black") +
  scale_fill_manual(values = c(BM = "#2a78d6", duaIterate = "#eb6834")) +
  guides(fill = "none") +
  facet_wrap(~congestion_regime) +
  labs(
    title = "90th-percentile OD-specific R-gap",
    subtitle = "90th percentile of the OD-specific R-gap across the \n10 active OD pairs within each seed (final episode)",
    x = "Algorithm",
    y = "90th percentile R-gap (%)"
  )

ggsave(here("r","RQ1","figures","plot3_od_rgap_worstcase.png"))
```

## Interpretation plot 3:

- **Near free-flow**:
Both algorithms achieve excellent OD-specific equilibrium quality.
  - DuaIterate has slightly greater variability across seeds. 
Hence the low aggregated R-gap is accompanied by good equilibirum quality for nearly all OD pairs.

- **Moderate**:
Both algorithms achieve great OD-specific equilibrium, though with slightly greater values than in the near free flow case. Both algorithms, show for 2 seeds a 90th percentile r-gap between 5-10%, they would still be consider pretty close to equilibrium. 

- **High congestion**:
Both algorithms exhibit substantially larger 90th-percentile R-gap values than under lower demand. BM consistently presents higher values, with 3 seeds with a 90th percentile r-gap value above 10%.

This reflects that for higher congestions, despite achieving low aggregated r-gap values, some subset of OD pairs remain poorly equilibrated. 

<!-- The file refined_rgap_by_od is not needed to be analyzed. A plot using this file would not add anything new. -->



