---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ1:** Does Bush-Mosteller multi-agent RL converge to a Dynamic User Equilibrium
(DUE) in a microscopic SUMO traffic simulation?

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

**The plots to answer this question will be:**

1. Aggregate R-gap evolution (main convergence evidence).
2. Heatmap R-gap by departure interval (all episodes)
3. Worst-case (90th percentile) OD-specific R-gap by seed

For more details, check the research questions document.



```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(ggbeeswarm)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
congestion_regime_tiers <- c("Near free-flow", "Moderate congestion", "High congestion")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```


```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ1","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
data <- list.files(data_dir, pattern="\\.parquet$", full.names = TRUE) |> 
  (\(x) set_names(x, tools::file_path_sans_ext(basename(x))))() |> 
  map(read_parquet)
```

```{r}
#| label: representative_seeds_spaghetti_plots

# Randomly select 3 seeds (seeded so the same 3 - and their colors - are
# picked every time this re-renders)
set.seed(123)
representative_seeds <- sample(unique(data$bm_rgap$seed), size = 3)

labels_seeds <- c("S1", "S2", "S3")

color_lines <- c(
  "S1" = "#bcb8b1",  
  "S2" = "#4A4844", 
  "S3" = "#11110F"   
  )  
```

```{r}
#| label: get_warm_up_period(exploration)

# It's not perfectly precise — the true per-agent start is episode > 24 and all routes visited, so some agents start a few episodes later than 24. But it's the honest guaranteed floor ("no agent has started learning before this point"), it's already computed for every run via config.warm_up, and given the plot already treats each seed's trajectory independently without forcing alignment (per the existing code comment about seeds finishing at very different episode counts), a single shared floor line is consistent with that treatment rather than adding seed-specific precision the rest of the plot doesn't have either.

exploration_phase_floor <- as.numeric(unique(data$bm_rgap$warm_up))
```

```{r}
#| label: data_plot1

# Not aggregating mean +/- SD across seeds here: seeds finish training at very
# different episode counts (e.g. High regime: some stop by episode ~22, one
# runs to 129), so for a large share of each panel's episode range the "mean"
# would silently collapse to a single seed's raw value (SD undefined). Each
# seed's own trajectory is plotted instead - no cross-seed averaging, so no
# misleading shrinking-N problem.

rgap_df <- data$bm_rgap |>
  mutate(
    congestion_regime = make_congestion_regime(n_agents),
    ) |>
  filter(
    seed %in% representative_seeds
  ) |> 
  mutate(
    seed_fct = factor(seed, levels = representative_seeds, labels = labels_seeds)
  )

plot1_annotation <- rgap_df |> 
  group_by(seed, n_agents) |> 
  slice_max(episode, n = 1, with_ties = FALSE)
```

```{r}
#| label: plot1

fill <- "#E8E5DE"
alpha <- 0.8
color <- "#6F6C67"

ggplot(
  data = rgap_df,
  mapping = aes(x = episode, y = rgap, color = seed_fct, group = seed_fct)
) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = fill,
    alpha = alpha
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = color,
  linewidth = 0.5
  ) +
  annotate(
    "text",
    x = (exploration_phase_floor - 2) / 2,
    y = Inf,
    label = "warm-up \nphase",
    fontface = "bold",
    colour = color,
    vjust = 2.5,
    size = 4
  ) +
  geom_hline(
    yintercept = 5,
    colour = "grey40",
    linetype = "longdash",
    linewidth = 0.5
  ) +
  geom_line() +
  geom_point(data = plot1_annotation, size = 1.25) +
  geom_hline(yintercept = 0, alpha = 0.6, linewidth = 0.7, color = "#7D7D7D") +
  facet_grid(. ~ congestion_regime, scales = "free_x", space = "free_x") +
  scale_color_manual(name = "Seed", values = color_lines) +
  labs(
    title = "R-gap evolution across episodes",
    subtitle = "Training trajectories for three random seeds (BM)",
    x = "Episode",
    y = "R-gap (%)"
  ) +
  facetted_pos_scales(
    x = list(
      scale_x_continuous(breaks = scales::breaks_width(10)),
      scale_x_continuous(breaks = scales::breaks_width(10)),
      scale_x_continuous(breaks = scales::breaks_width(10))
    )
  ) +
  scale_y_continuous(breaks = scales::breaks_width(5)) +
  theme(
    panel.grid.major.y = element_line(
      color = "#E7E5DB",
      linewidth = 0.30
    ),
    panel.grid.minor.y = element_blank()
  )

ggsave(here("r","RQ1","figures","plot1_rgap_evolution.png"), width = 12, height = 6)
```

```{r}
#| label: data_plot2

# Not aggregating (mean) across seeds here, same reasoning as plot1: seeds
# finish training at very different episode counts, so a per-episode mean
# across seeds would silently collapse to a shrinking number of contributing
# seeds. Each of the 3 representative seeds is plotted as its own heatmap
# instead (rows), with congestion regime as columns - full seed set is still
# used for any statistical summaries elsewhere, just not for this plot.
#
# duaIterate is excluded (as before): its refined R-gap was only ever
# computed for the final iteration, not per-episode, so it has nothing to
# show along an episode x-axis.

refined_rgap_df <- data$bm_refined_rgap |>
  mutate(
    congestion_regime = make_congestion_regime(n_agents)
  ) |>
  filter(
    seed %in% representative_seeds
  ) |>
  mutate(
    seed_fct = factor(seed, levels = representative_seeds, labels = labels_seeds)
  )

cocient <- min(data$bm_refined_rgap$time_interval)

refined_rgap_df_agg <- refined_rgap_df |> 
  mutate(
    interval_group = time_interval %/% cocient
  ) |> 
  mutate(
    interval_group = factor(
      interval_group,
      levels = 1:max(interval_group),
      labels = paste0(
        seq(10, by = 10, length.out = max(interval_group)),
        "—",
        seq(19, by = 10, length.out = max(interval_group))
      )
      )
  ) |> 
  group_by(interval_group, seed_fct, congestion_regime, episode) |> 
  summarise(median_refined_rgap=median(refined_rgap)) |> 
  ungroup()

refined_rgap_df_agg
```

```{r}
#| label: plot2

# Choose x-axis ticks
max_episode <- max(refined_rgap_df_agg$episode)
step <- floor(max_episode / 40) * 10
breaks <- seq(0, max_episode, by = step)

# Remove breaks that are within step of the learning start
breaks <- breaks[abs(breaks - exploration_phase_floor) >= step]
# Add the learning start break
breaks <- sort(c(breaks, exploration_phase_floor))
labels <- as.character(breaks)
labels[breaks==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")

upper_limit_color_scale <- quantile(refined_rgap_df_agg$median_refined_rgap, probs = c(0.95))

ggplot(
  data = refined_rgap_df_agg,
  mapping = aes(x = episode, y = interval_group, fill = median_refined_rgap)
) +
  scale_x_continuous(breaks = breaks, labels = labels) +
  geom_tile() +
  scale_y_discrete(limits = rev) +
  scale_fill_viridis_c(
    limits = c(0, upper_limit_color_scale),
    oob = scales::squish,
    option = "C"
  ) +
  facet_grid(seed_fct ~ congestion_regime, scales = "free_x", space = "free_x") +
  labs(
    title = "Refined R-gap by departure interval across episodes",
    subtitle = "Departure intervals were aggregated into 10-interval groups \n(median refined R-gap per group).",
    x = "Episode",
    y = "Departure interval (group)",
    fill = "Refined\nR-gap (%)"
  ) + 
  geom_vline(xintercept = exploration_phase_floor, colour = "black", linewidth = 1.2)   +
  geom_vline(xintercept = exploration_phase_floor, colour = "white", linewidth = 0.5,  linetype = "dotdash") +
  theme(
    panel.grid.major.y = element_line(
      color = "#E7E5DB",
      linewidth = 0.30
    ),
    panel.grid.minor.y = element_blank()
  )

ggsave(here("r","RQ1","figures","plot2_refined_rgap_heatmap.png"), width = 12, height = 10)
```

```{r}
#| label: data_plot3
#| eval: false

# Plot3: worst-case (90th percentile) OD-specific r-gap, one value per
# algorithm x seed, computed across that seed's own final episode. 

rgap_od_df <- bind_rows(
  mutate(data$bm_rgap_by_od, algorithm = "BM"),
  mutate(data$dua_rgap_by_od, algorithm = "duaIterate")
) |>
  mutate(congestion_regime = make_congestion_regime(n_agents))

rgap_od_seed_summary <- rgap_od_df |>
  group_by(algorithm, congestion_regime, n_agents, seed) |>
  filter(episode == max(episode)) |>
  summarise(
    worst_case_rgap = max(rgap, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| label: plot3
#| eval: false

ggplot(
  data = rgap_od_seed_summary,
  mapping = aes(x = algorithm, y = worst_case_rgap, fill = algorithm)
) +
  geom_boxplot(outlier.shape = NA, width = 0.5, linewidth = 0.6, alpha = 0.6) +
  geom_jitter(width = 0.1, height = 0, size = 2.5, shape = 21, color = "black") +
  scale_fill_manual(values = c(BM = "#2a78d6", duaIterate = "#eb6834")) +
  guides(fill = "none") +
  facet_wrap(~congestion_regime) +
  labs(
    title = "Worst-case OD-specific R-gap",
    subtitle = "Worst-case of the OD-specific R-gap across the \n10 active OD pairs within each seed (final episode)",
    x = "Algorithm",
    y = "Worst-case R-gap (%)"
  )

# ggsave(here("r","RQ1","figures","plot3_od_rgap_worstcase.png"))
```

## Interpretation plot 3:

- **Near free-flow**:
Both algorithms achieve excellent OD-specific equilibrium quality.
  - DuaIterate has slightly greater variability across seeds. 
Hence the low aggregated R-gap is accompanied by good equilibirum quality for nearly all OD pairs.

- **Moderate**:
Both algorithms achieve great OD-specific equilibrium, though with slightly greater values than in the near free flow case. Both algorithms, show for 2 seeds a 90th percentile r-gap between 5-10%, they would still be consider pretty close to equilibrium. 

- **High congestion**:
Both algorithms exhibit substantially larger 90th-percentile R-gap values than under lower demand. BM consistently presents higher values, with 3 seeds with a 90th percentile r-gap value above 10%.

This reflects that for higher congestions, despite achieving low aggregated r-gap values, some subset of OD pairs remain poorly equilibrated. 

```{r}
#| label: data_plot4
#| 
rgap_od_df_plot4 <- rgap_od_df |>
  select(-c(run_id, warm_up, n_agents)) |> 
  filter(
    seed %in% representative_seeds[1]
  ) |> 
  group_by(algorithm, congestion_regime, seed, origin, destination) |> 
  slice_max(episode)

demand_odt_df_plot4 <- data$demand_odt |> 
  mutate(
    congestion_regime = make_congestion_regime(n_agents),
  ) |> 
  select(-c(run_id, warm_up, n_agents)) |> 
  group_by(
    origin, destination, seed, congestion_regime
  ) |> 
  summarise(
    demand = sum(count)
  ) |> 
  ungroup()

# Add the demand for each OD pair
rgap_od_df_plot4 <- rgap_od_df_plot4 |> 
  left_join(
    demand_odt_df_plot4,
    by = c("origin", "destination", "seed", "congestion_regime")
  )
```


```{r}
#| label: plot4

ggplot(
  data = rgap_od_df_plot4,
  mapping = aes(x = algorithm, y = rgap, fill = algorithm)
) +
  geom_violin(
    trim = FALSE,
    alpha = 0.7,
    linewidth = 0.3
  ) +
  geom_boxplot(
    width = 0.12,
    fill = "white",
    outlier.shape = NA,
    linewidth = 0.3
  ) +
  ggbeeswarm::geom_quasirandom(
  width = 0.15,
  alpha = 0.3,
  size = 0.6,
  ) +
  geom_hline(yintercept = 0, alpha = 0.75, linewidth = 0.7, color = "#7D7D7D") +
  facet_grid(~ congestion_regime) +
  labs(
    x = NULL,
    y = "OD-specific R-gap (%)",
    title = "Final OD-specific R-gap by Algorithm",
    subtitle = "Representative random seed; distributions computed from the final episodes."
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(face = "bold"),
    strip.text = element_text(
      size = 11,
      face = "bold"
    )
  ) +
  scale_fill_manual(
    values = c(
      "BM" = "#9999a1",
      "duaIterate" = "#e6e6e9"
    )
  ) +
  scale_y_continuous(breaks = scales::breaks_width(10))

ggsave(here("r","RQ1","figures","plot4_distribution_final_od_r_gap.png"))
```

<!-- The file refined_rgap_by_od is not needed to be analyzed. A plot using this file would not add anything new. -->



