---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ1:** Does Bush-Mosteller multi-agent RL converge to a Dynamic User Equilibrium
(DUE) in a microscopic SUMO traffic simulation?

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

**The plots to answer this question will be:**

1. Aggregate R-gap evolution (main convergence evidence).
2. Heatmap R-gap by departure interval (all episodes)
3. Worst-case (90th percentile) OD-specific R-gap by seed

For more details, check the research questions document.



```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
congestion_regime_tiers <- c("Near free-flow", "Moderate", "High")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```


```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ1","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
data <- list.files(data_dir, pattern="\\.parquet$", full.names = TRUE) |> 
  (\(x) set_names(x, tools::file_path_sans_ext(basename(x))))() |> 
  map(read_parquet)
```

```{r}
#| label: representative_seeds_spaghetti_plots

# Randomly select 3 seeds (seeded so the same 3 - and their colors - are
# picked every time this re-renders)
set.seed(123)
representative_seeds <- sample(unique(data$bm_rgap$seed), size = 3)

labels_seeds <- c("S1", "S2", "S3")

color_lines <- c(
  "S1" = "#0072B2",  # blue
  "S2" = "#E69F00",  # orange
  "S3" = "#009E73"   # bluish green
  )  
```


```{r}
#| label: data_plot1

# Not aggregating mean +/- SD across seeds here: seeds finish training at very
# different episode counts (e.g. High regime: some stop by episode ~22, one
# runs to 129), so for a large share of each panel's episode range the "mean"
# would silently collapse to a single seed's raw value (SD undefined). Each
# seed's own trajectory is plotted instead - no cross-seed averaging, so no
# misleading shrinking-N problem.

rgap_df <- data$bm_rgap |>
  mutate(
    congestion_regime = make_congestion_regime(n_agents),
    ) |>
  filter(
    seed %in% representative_seeds
  ) |> 
  mutate(
    seed_fct = factor(seed, levels = representative_seeds, labels = labels_seeds)
  )

plot1_annotation <- rgap_df |> 
  group_by(seed, n_agents) |> 
  slice_max(episode, n = 1, with_ties = FALSE)
```

```{r}
#| label: plot1

ggplot(
  data = rgap_df,
  mapping = aes(x = episode, y = rgap, color = seed_fct, group = seed_fct)
) +
  geom_line() +
  geom_point(data = plot1_annotation, size = 1.25) +
  geom_hline(yintercept = 0, alpha = 0.6, linewidth = 0.7, color = "#7D7D7D") +
  facet_grid(. ~ congestion_regime, scales = "free_x", space = "free_x") +
  scale_color_manual(name = "Seed", values = color_lines) +
  labs(
    title = "R-gap evolution across episodes",
    subtitle = "Training trajectories for three random seeds (BM)",
    x = "Episode",
    y = "R-gap (%)"
  ) +
  facetted_pos_scales(
    x = list(
      scale_x_continuous(breaks = scales::breaks_width(10)),
      scale_x_continuous(breaks = scales::breaks_width(10)),
      scale_x_continuous(breaks = scales::breaks_width(10))
    )
  ) +
  scale_y_continuous(breaks = scales::breaks_width(5))

ggsave(here("r","RQ1","figures","plot1_rgap_evolution.png"), width = 12, height = 6)
```

```{r}
#| label: data_plot2

# Not aggregating (mean) across seeds here, same reasoning as plot1: seeds
# finish training at very different episode counts, so a per-episode mean
# across seeds would silently collapse to a shrinking number of contributing
# seeds. Each of the 3 representative seeds is plotted as its own heatmap
# instead (rows), with congestion regime as columns - full seed set is still
# used for any statistical summaries elsewhere, just not for this plot.
#
# duaIterate is excluded (as before): its refined R-gap was only ever
# computed for the final iteration, not per-episode, so it has nothing to
# show along an episode x-axis.

refined_rgap_df <- data$bm_refined_rgap |>
  mutate(
    congestion_regime = make_congestion_regime(n_agents)
  ) |>
  filter(
    seed %in% representative_seeds
  ) |>
  mutate(
    seed_fct = factor(seed, levels = representative_seeds, labels = labels_seeds)
  )
```

```{r}
#| label: plot2

ggplot(
  data = refined_rgap_df,
  mapping = aes(x = episode, y = time_interval, fill = refined_rgap)
) +
  geom_tile() +
  scale_fill_viridis_c(
    limits = c(0, 50),
    oob = scales::squish
  ) +
  facet_grid(seed_fct ~ congestion_regime, scales = "free_x", space = "free_x") +
  labs(
    title = "Refined R-gap by departure interval across episodes",
    subtitle = "Per-seed trajectories for three random seeds (BM)",
    x = "Episode",
    y = "Departure interval",
    fill = "Refined\nR-gap (%)"
  )

ggsave(here("r","RQ1","figures","plot2_refined_rgap_heatmap.png"), width = 12, height = 10)
```

```{r}
#| label: data_plot3

# Plot3: worst-case (90th percentile) OD-specific r-gap, one value per
# algorithm x seed, computed across that seed's own final episode. 

rgap_od_df <- bind_rows(
  mutate(data$bm_rgap_by_od, algorithm = "BM"),
  mutate(data$dua_rgap_by_od, algorithm = "duaIterate")
) |>
  mutate(congestion_regime = make_congestion_regime(n_agents))

rgap_od_seed_summary <- rgap_od_df |>
  group_by(algorithm, congestion_regime, n_agents, seed) |>
  filter(episode == max(episode)) |>
  summarise(
    p90_rgap = quantile(rgap, 0.9),
    .groups = "drop"
  )
```

```{r}
#| label: plot3

ggplot(
  data = rgap_od_seed_summary,
  mapping = aes(x = algorithm, y = p90_rgap, fill = algorithm)
) +
  geom_boxplot(outlier.shape = NA, width = 0.5, linewidth = 0.6, alpha = 0.6) +
  geom_jitter(width = 0.08, height = 0, size = 2.5, shape = 21, color = "black") +
  scale_fill_manual(values = c(BM = "#2a78d6", duaIterate = "#eb6834")) +
  guides(fill = "none") +
  facet_wrap(~congestion_regime) +
  labs(
    title = "Worst-case OD-specific R-gap",
    subtitle = "90th percentile across each seed's 10 active OD pairs, final episode",
    x = "Algorithm",
    y = "90th percentile R-gap (%)"
  )

ggsave(here("r","RQ1","figures","plot3_od_rgap_worstcase.png"))
```



<!-- The file refined_rgap_by_od is not needed to be analyzed. A plot using this file would not add anything new. -->



