---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ1:** Does Bush-Mosteller multi-agent RL converge to a Dynamic User Equilibrium
(DUE) in a microscopic SUMO traffic simulation?

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

**The plots to answer this question will be:**

1. Aggregate R-gap evolution (main convergence evidence). 
  In particular:
  Figure 1. Learning curves
    Mean R-gap vs episode.
    Mean ± SD across seeds.
    Faceted by congestion regime.
    One line per algorithm.


2. Refined R-gap by departure interval in final episode (temporal heterogeneity).
3. OD-specific R-gap boxplot in final episode (spatial heterogeneity).
4. Maximum OD-specific R-gap evolution (robustness of convergence).
session_notes_2026-07-17.md
```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
```

```{r}
#| label: load_custom_functions

# library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())
```


```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ1","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
data <- list.files(data_dir, pattern="\\.parquet$", full.names = TRUE) |> 
  (\(x) set_names(x, tools::file_path_sans_ext(basename(x))))() |> 
  map(read_parquet)
```

```{r}
#| label: data_plot1

rgap_df <- bind_rows(
  mutate(data$bm_rgap, algorithm = "BM"),
  mutate(data$dua_rgap, algorithm = "duaIterate")
) |> 
  mutate(
    congestion_regime = factor(
      n_agents,
      levels = c(1500,2000,2500),
      labels = c("Low (1500)", "Moderate (2000)", "High (2500)")
    )
  )

rgap_summary <- rgap_df |> 
  group_by(algorithm, congestion_regime, episode) |> 
  summarise(
    mean_rgap = mean(rgap),
    sd_rgap = sd(rgap),
    .groups = "drop"
  )

rgap_summary <- rgap_summary |> 
  filter(algorithm == "BM")
```

```{r}
#| label: plot1

ggplot(
  data = rgap_summary,
  mapping = aes(x = episode, y = mean_rgap, color = algorithm, fill = algorithm)
) +
  geom_ribbon(
    aes(ymin = mean_rgap - sd_rgap, ymax = mean_rgap + sd_rgap),
    alpha = 0.2, color = NA
  ) +
  geom_line() +
  facet_wrap(~congestion_regime) +
  labs(
    title = "R-gap evolution across episodes",
    subtitle = "Mean ± SD across seeds",
    x = "Episode",
    y = "R-gap (%)",
    color = "Algorithm",
    fill = "Algorithm"
  )

ggsave(here("r","RQ1","figures","plot1_rgap_evolution.png"))
```

```{r}
#| label: data_plot2

# Plot2: Refined r-gap (by interval) of the last episode for the learning algorithms

refined_rgap_df <- data |> 
  map("refined_rgap") |> 
  imap(\(x,name) mutate(x, algorithm = name)) |> 
  list_rbind()

refined_rgap_df_last_ep <- refined_rgap_df |> 
  group_by(algorithm) |> 
  filter(episode == max(episode)) |> 
  ungroup()
```

```{r}
#| label: plot2

### OPTION A
ggplot(
  data = refined_rgap_df_last_ep,
  mapping = aes(
    x = time_interval,
    y = refined_rgap,
    color = algorithm
  )
) +
  geom_smooth(se = FALSE, span = 0.15) +
  labs(
    title = "Refined R-gap by Departure Interval (smoothed)",
    subtitle = "Comparison between algorithms in the final episode",
    x = "Departure Time Interval",
    y = "Refined R-gap (%) (smoothed)"
  )

### OPTION B
ggplot(
  data = refined_rgap_df_last_ep,
  mapping = aes(
    x = time_interval,
    y = refined_rgap,
    color = algorithm
  )
) +
  geom_line(alpha = 0.7) +
  labs(
    title = "Refined R-gap by Departure Interval",
    subtitle = "Comparison between algorithms in the final episode",
    x = "Departure Time Interval",
    y = "Refined R-gap (%)"
  )
ggsave(here("r","RQ1","figures","plot2a_refined_rgap.png"))
```

```{r}
#| label: data_plot3

# Analyze OD-specific R-gap (distribution)

rgap_od_df <- data |> 
  map("rgap_od") |> 
  imap(\(x,name) mutate(x, algorithm = name)) |> 
  list_rbind()

rgap_od_last_ep_df <- rgap_od_df |> 
  group_by(algorithm) |> 
  filter(episode == max(episode))
```

```{r}
#| label: plot3

# Option C
ggplot(
  data = rgap_od_last_ep_df,
  mapping = aes(
    x = algorithm,
    y = rgap,
    fill = algorithm
  ) 
) +
  geom_boxplot() +
  labs(
  title = "Distribution of OD-Specific R-gaps",
  subtitle = "Comparison between algorithms in the final episode",
  x = NULL,
  y = "OD-Specific R-gap (%)"
)
ggsave(here("r","RQ1","figures","plot3c_rgap_od.png"))
```

```{r}
#| label: data_plot4

# Visualize the maximum od-specific r-gap evolution

rgap_od_df_max <- 
  rgap_od_df |> 
  group_by(algorithm, episode) |> 
  summarise(max_rgap = max(abs(rgap))) |> filter(algorithm != "duaIterate")
```

```{r}
#| label: plot4

ggplot(
  data = rgap_od_df_max,
  mapping = aes(x = episode, y = max_rgap, color = algorithm)
) + 
  geom_line() +
  labs(
    title = "Evolution of maximum OD-Specific R-gap",
    subtitle = "Comparison between algorithms in the final episode",
    x = "Episode",
    y = "Max. od-specific rgap (%)" 
  )
ggsave(here("r","RQ1","figures","plot4_rgap_od_max.png"))
```

<!-- The file refined_rgap_by_od is not needed to be analyzed. A plot using this file would not add anything new. -->



