---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ11:** Analyze how incorporating travel-time variability
  (variance) into the perceived route cost affects route-choice
  behavior, learning dynamics, and convergence properties of BM
  model

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

**The plots to answer this question will be:**

1. Risk scatter (mechanism plot)
2. R-gap evolution across $\theta$
3. Final route flow share vs $\theta$
4. Flow evolution over episodes

For more details, check the research questions document.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(glue)
library(ggtext)
library(patchwork)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())
```


```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ11","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
data <- list.files(data_dir, pattern="\\.parquet$", full.names = TRUE) |> 
  (\(x) set_names(x, tools::file_path_sans_ext(basename(x))))() |> 
  map(read_parquet)
```

```{r}
#| label: random_seed
# Randomly select 1 seed 
set.seed(123)
random_seed <- sample(unique(data$bm_rgap$seed), size = 1)
```

```{r}
#| label: get_warm_up_period(exploration)

# It's not perfectly precise — the true per-agent start is episode > 24 and all routes visited, so some agents start a few episodes later than 24. But it's the honest guaranteed floor ("no agent has started learning before this point"), it's already computed for every run via config.warm_up, and given the plot already treats each seed's trajectory independently without forcing alignment (per the existing code comment about seeds finishing at very different episode counts), a single shared floor line is consistent with that treatment rather than adding seed-specific precision the rest of the plot doesn't have either.

exploration_phase_floor <- as.numeric(unique(data$bm_rgap$warm_up))
```

```{r}
#| label: representative_OD

# Choose an OD whose shortest route is not the most stable (hence, there exist alternative routes that are longer but more stable).
# Is gonna be chosen taking into account equally weighted means (without memory mechanism) and with theta=0 (under normal drivers scenario). 
# We do it this way because we merely wanna inspect the network/demand structure to choose our OD and isolate it from the effect of other variables

# We wamt that the shortest route is at least 5 seconds faster than the most stable route
# (this way we could see a difference in the r-gap when implemented risk sensitivity model)
min_mean_tt_diff <- 5

rewards_df <- data$rewards |> 
  mutate(
    reliability_sensitivity = as.numeric(reliability_sensitivity)
  ) |> 
  filter(
    seed == random_seed,
    reliability_sensitivity == 0
  ) |> 
  select(-c(run_id, seed, warm_up, reliability_sensitivity))

actions_df <- data$actions |> 
  mutate(
    reliability_sensitivity = as.numeric(reliability_sensitivity)
  ) |> 
  filter(
    reliability_sensitivity == 0,
    seed == random_seed
  ) |> 
  select(-c(run_id, seed, warm_up, reliability_sensitivity))

agents_od_df <- data$agents_od |> 
  mutate(
    reliability_sensitivity = as.numeric(reliability_sensitivity)
  ) |> 
  filter(
    seed == random_seed,
    reliability_sensitivity == 0
  ) |> 
  select(-c(post_warm_up, run_id, seed, warm_up, reliability_sensitivity, departure_time)) |> 
  rename(agent_id = id)

representative_od_df <- rewards_df |> 
  left_join(actions_df, by = c("episode", "agent_id")) |> 
  left_join(agents_od_df, by = "agent_id")

representative_od_df_summary <- representative_od_df |> 
  group_by(origin, destination, action) |> 
  summarise(
    mean_tt = mean(reward),
    sd_tt = sd(reward),
    count = n()
  ) |> 
  ungroup() |> 
  filter(
    count > 15
  )

representative_od_decision_df <- representative_od_df_summary |> 
  group_by(
    origin, destination
  ) |> 
  mutate(
    is_shortest = mean_tt == min(mean_tt),
    is_stablest = sd_tt == min(sd_tt),
    candidate_od = is_shortest != is_stablest
  ) |> 
  filter(
    candidate_od
  )

representative_od <- representative_od_decision_df |> 
  summarise(
    diff = abs(mean_tt[1] - mean_tt[2]) > min_mean_tt_diff
  ) |> 
  filter(diff) |> 
  arrange(desc(diff)) |> 
  slice_head(n = 1) |> 
  select(origin, destination)
  
representative_od_routes <- representative_od_df_summary |> 
  filter(
    origin == representative_od$origin,
    destination == representative_od$destination
  ) |> 
  arrange(mean_tt)
```

```{r}
#| label: data_plot1_points
agents_representative_od_df <- agents_od_df |> 
  filter(
    origin == representative_od$origin,
    destination == representative_od$destination
  ) 
agent <- agents_representative_od_df[ceiling(nrow(agents_representative_od_df) / 2), ]

# The points in the plot 1, will represent the beliefs of an agent about 
# the mean_tt and sd of its routes. 
# We just need one agent, because that way we can easily illustrate 
# the mechanisms and how an agent perception about routes would change by 
# considering different thetas we just need to use one agent.
# It is an agent belonging to the representative OD that we are gonna talk
# about in the following plots
# We want the agent to be in the middle of the simulation horizon

agent_last_episode_info <- data$bm_results |> 
  mutate(
    reliability_sensitivity = as.numeric(reliability_sensitivity)
  ) |> 
  filter(
    reliability_sensitivity == 0,
    seed == random_seed
  ) |> 
  select(-c(run_id, seed, warm_up, reliability_sensitivity)) |> 
  filter(
    agent_id == agent$agent_id,
    episode == max(episode)
  ) |> 
  select(
    episode, route_id, PT, sigma_r
  )
```

```{r}
#| label: theta_values

theta_values <- as.numeric(data$demand_odt$reliability_sensitivity |> unique())
```

```{r}
#| label: data_plot1_lines

# Intuition about the plot:
# 1) I wanna see which route is the best for each theta value.
# 2) Draw the contour lines for each theta value. In particular, draw the
# contour corresponding to the minimum (best) route cost. This shows which
# other combinations of values result in the same cost as the best route.

# map_dfr: runs the function once for each value in theta_values, and stacks the three results into one dataframe

iso_cost_lines_tmp <- map_dfr(theta_values, function(theta){
  # Choose route with the minimum generalized cost
  best <- agent_last_episode_info |> 
    mutate(
      cost = PT + theta * sigma_r
    ) |> 
    slice_min(cost, n = 1)
  
  # Line when we do not take into account the variability.
  # Basically the contour line is vertical because variability does not matter for the generalized cost
  if (theta == 0){
    tibble(theta, type = "vertical", xintercept = best$PT, route_id = best$route_id, cost = best$cost)
  } else{
    # (σ_r = -PT_r/θ + cost/θ, i.e. slope -1/θ, intercept cost/θ)
    tibble(theta, type = "line", slope = -1 / theta, intercept = best$cost / theta, route_id = best$route_id, cost = best$cost)
  }
})

iso_cost_lines <- iso_cost_lines_tmp |> mutate(
  theta = factor(theta)
)
```

```{r}
#| label: Tweaks_to_improve_visualization

# 1. Improve route_id labels
path_levels <- agent_last_episode_info$route_id |> unique()
path_levels_labels <- map_chr(path_levels, \(x) paste0("Path ", x+1))
agent_last_episode_info_plot <- agent_last_episode_info |> 
  mutate(
    path_id = factor(route_id, levels = path_levels, labels = path_levels_labels)
  ) |> 
  select(-route_id)

# 2. Choose only iso cost lines that have different minimum cost routes
iso_cost_lines_plot_tmp <- iso_cost_lines |> 
  mutate(
    theta_num = as.numeric(theta)
  ) |> 
  group_by(route_id) |> 
  mutate(
    plot = theta_num == max(theta_num)
  ) |> 
  filter(plot) |> 
  select(-c(plot, theta_num))

# 3. Highlight min_cost routes
min_cost_routes <- iso_cost_lines_plot$route_id 
min_cost_routes_df <- agent_last_episode_info |> 
  filter(
    route_id %in% min_cost_routes
  )

# 4. Label the iso cost lines
iso_cost_lines_plot <- iso_cost_lines_plot_tmp |> 
  left_join(min_cost_routes_df, by = "route_id") |> 
  mutate(
    theta_num = as.numeric(theta)
  ) |> 
  mutate(
    x = ifelse(type == "vertical", xintercept, (cost - theta_num * sigma_r)),
    y = ifelse(type == "vertical", max(agent_last_episode_info_plot$sigma_r), 0),
    label = glue("\u03b8 = {theta}, C = {cost} s")
  )
```




```{r}
#| label: plot_1
ggplot(
  data = agent_last_episode_info_plot, 
  mapping = aes(x = PT, y = sigma_r)
) +
  geom_point() +
  geom_point(
    data = min_cost_routes_df,
    mapping = aes(x = PT, y = sigma_r),
    size = 4, 
    shape = "circle open", 
  ) +
  geom_vline(
    data = filter(iso_cost_lines_plot, type == "vertical"),
    mapping = aes(
      xintercept = xintercept, 
      ),
      linetype = "dashed"
  ) +
  geom_abline(
    data = filter(iso_cost_lines_plot, type == "line"),
    mapping = aes(
      slope = slope, 
      intercept = intercept, 
      ), 
    linetype = "dashed"
  ) +
  ggrepel::geom_text_repel(
    mapping = aes(label = path_id),
    size = 3.5,
    box.padding = 0.5,
    point.padding = 0.3,
    force = 1,
    max.overlaps = Inf,
    seed = 123,
  ) +
  ggrepel::geom_text_repel(
    data = iso_cost_lines_plot,
    mapping = aes(
      x = x,
      y = y,
      label = label
    ),
    size = 3.5,
    box.padding = 1,
    point.padding = 0.3,
    force = 1,
    max.overlaps = Inf,
    seed = 123,
    nudge_x = 2,
    fontface = "bold",
    segment.alpha = 0.5,
    segment.size = 0.3,
  ) +
  labs(
    title = "Risk sensitivity changes the minimum-cost route",
    subtitle = "One representative agent; highlighted points show \nthe minimum-cost route. Dashed lines indicate equal perceived cost.",
  x = expression("Perceived travel time " * PT[r] * " (s)"),
  y = expression("Perceived travel time variability " * sigma[r] * " (s)"),
  )

ggsave(here("r","RQ11","figures","plot1_risk_sensitivity_mechanism.png"), width = 12, height = 6)
```

```{r}
#| label: representative_seeds
representative_seeds <- sample(unique(data$bm_rgap$seed), size = 3)
labels_seeds <- c("S1", "S2", "S3")
```

```{r}
#| label: data_plot2

representative_reliability_sensitivity_levels <- c(0, 10)
representative_reliability_sensitivity_levels_labels <- map_chr(
  representative_reliability_sensitivity_levels,
  \(x) paste0("\u03b8 = ", x)
)

# Not aggregating mean +/- SD across seeds here: seeds finish training at very
# different episode counts (e.g. High regime: some stop by episode ~22, one
# runs to 129), so for a large share of each panel's episode range the "mean"
# would silently collapse to a single seed's raw value (SD undefined). Each
# seed's own trajectory is plotted instead - no cross-seed averaging, so no
# misleading shrinking-N problem.

rgap_df <- data$bm_rgap |> 
  mutate(
    reliability_sensitivity = as.numeric(reliability_sensitivity)
  ) |> 
  filter(
    seed %in% representative_seeds,
    reliability_sensitivity %in% representative_reliability_sensitivity_levels
  ) |> 
  mutate(
    seed_fct = factor(seed, levels = representative_seeds, labels = labels_seeds)
  ) |> 
  select(-c(run_id, warm_up, seed)) |> 
  mutate(
    reliability_sentivity_fct = 
      factor(reliability_sensitivity,
             levels = representative_reliability_sensitivity_levels,
             labels = representative_reliability_sensitivity_levels_labels
      )
  ) |> 
  mutate(
    reliability_sentivity_fct = reliability_sentivity_fct
  )
```

```{r}
#| label: breaks_plot2

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
step <- floor(max_episode / 30) * 10
breaks <- seq(0, max_episode, by = step)

# Remove breaks that are within step of the learning start
breaks <- breaks[abs(breaks - exploration_phase_floor) >= (step/2.25)]
# Add the learning start break
breaks <- sort(c(breaks, exploration_phase_floor))
labels <- as.character(breaks)
labels[breaks==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```

```{r}
#| label: plot2

upper_limit_y_axis <- quantile(rgap_df$rgap, probs = c(0.99))

ggplot(
  data = rgap_df,
  mapping = aes(x = episode, y = rgap, group = seed_fct)
) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = "red",
    alpha = 0.08
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.5
  ) +
  geom_line(alpha = 0.6, linewidth = 0.4) +
  geom_hline(yintercept = 0, alpha = 0.6, linewidth = 0.7, color = "#7D7D7D") +
  geom_hline(
    yintercept = 5,
    colour = "grey25",
    linetype = "longdash",
    linewidth = 0.6,
  ) +
  facet_grid(reliability_sentivity_fct ~ .) +
  labs(
    title = "Risk sensitivity affects R-gap convergence",
    subtitle = glue("Representative learning trajectories for θ = {representative_reliability_sensitivity_levels[1]} and θ = {representative_reliability_sensitivity_levels[2]} across different seeds"),
    x = "Episode",
    y = "R-gap (%)"
  ) +
  coord_cartesian(ylim=c(-5,upper_limit_y_axis)) +
  theme(
    axis.title.x = element_text(margin = margin(t = 1)),
  ) +
  scale_x_continuous(breaks = breaks, labels = labels) +
  scale_y_continuous(
  breaks = function(x) {
    c(scales::breaks_pretty()(x), 5)
  }
)

ggsave(here("r","RQ11","figures","plot_2_convergence_plot.png"), width = 12, height = 6)
```



