---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ11:** Analyze how incorporating travel-time variability
  (variance) into the perceived route cost affects route-choice
  behavior, learning dynamics, and convergence properties of BM
  model

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

**The plots to answer this question will be:**

1. Risk scatter (mechanism plot)
2. R-gap evolution across $\theta$
3. Final route flow share vs $\theta$
4. Flow evolution over episodes

For more details, check the research questions document.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(glue)
library(ggtext)
library(patchwork)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())
```


```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ11","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
data <- list.files(data_dir, pattern="\\.parquet$", full.names = TRUE) |> 
  (\(x) set_names(x, tools::file_path_sans_ext(basename(x))))() |> 
  map(read_parquet)
```

```{r}
#| label: random_seed
# Randomly select 1 seed 
set.seed(123)
random_seed <- sample(unique(data$bm_rgap$seed), size = 1)
```

```{r}
#| label: get_warm_up_period(exploration)

# It's not perfectly precise — the true per-agent start is episode > 24 and all routes visited, so some agents start a few episodes later than 24. But it's the honest guaranteed floor ("no agent has started learning before this point"), it's already computed for every run via config.warm_up, and given the plot already treats each seed's trajectory independently without forcing alignment (per the existing code comment about seeds finishing at very different episode counts), a single shared floor line is consistent with that treatment rather than adding seed-specific precision the rest of the plot doesn't have either.

exploration_phase_floor <- as.numeric(unique(data$bm_rgap$warm_up))
```

```{r}
#| label: representative_OD

# Choose an OD whose shortest route is not the most stable (hence, there exist alternative routes that are longer but more stable).
# Is gonna be chosen taking into account equally weighted means (without memory mechanism) and with theta=0 (under normal drivers scenario). 
# We do it this way because we merely wanna inspect the network/demand structure to choose our OD and isolate it from the effect of other variables

# We wamt that the shortest route is at least 5 seconds faster than the most stable route
# (this way we could see a difference in the r-gap when implemented risk sensitivity model)
min_mean_tt_diff <- 5

rewards_df <- data$rewards |> 
  mutate(
    reliability_sensitivity = as.numeric(reliability_sensitivity)
  ) |> 
  filter(
    seed == random_seed,
    reliability_sensitivity == 0
  ) |> 
  select(-c(run_id, seed, warm_up, reliability_sensitivity))

actions_df <- data$actions |> 
  mutate(
    reliability_sensitivity = as.numeric(reliability_sensitivity)
  ) |> 
  filter(
    reliability_sensitivity == 0,
    seed == random_seed
  ) |> 
  select(-c(run_id, seed, warm_up, reliability_sensitivity))

agents_od_df <- data$agents_od |> 
  mutate(
    reliability_sensitivity = as.numeric(reliability_sensitivity)
  ) |> 
  filter(
    seed == random_seed,
    reliability_sensitivity == 0
  ) |> 
  select(-c(post_warm_up, run_id, seed, warm_up, reliability_sensitivity, departure_time)) |> 
  rename(agent_id = id)

representative_od_df <- rewards_df |> 
  left_join(actions_df, by = c("episode", "agent_id")) |> 
  left_join(agents_od_df, by = "agent_id")

representative_od_df_summary <- representative_od_df |> 
  group_by(origin, destination, action) |> 
  summarise(
    mean_tt = mean(reward),
    sd_tt = sd(reward),
    count = n()
  ) |> 
  ungroup() |> 
  filter(
    count > 15
  )

representative_od_decision_df <- representative_od_df_summary |> 
  group_by(
    origin, destination
  ) |> 
  mutate(
    is_shortest = mean_tt == min(mean_tt),
    is_stablest = sd_tt == min(sd_tt),
    candidate_od = is_shortest != is_stablest
  ) |> 
  filter(
    candidate_od
  )

representative_od <- representative_od_decision_df |> 
  summarise(
    diff = abs(mean_tt[1] - mean_tt[2]) > min_mean_tt_diff
  ) |> 
  filter(diff) |> 
  arrange(desc(diff)) |> 
  slice_head(n = 1) |> 
  select(origin, destination)
  
representative_od_routes <- representative_od_df_summary |> 
  filter(
    origin == representative_od$origin,
    destination == representative_od$destination
  ) |> 
  arrange(mean_tt)
```

```{r}
#| label: data_plot1_points
agents_representative_od_df <- agents_od_df |> 
  filter(
    origin == representative_od$origin,
    destination == representative_od$destination
  ) 
agent <- agents_representative_od_df[ceiling(nrow(agents_representative_od_df) / 2), ]

# The points in the plot 1, will represent the beliefs of an agent about 
# the mean_tt and sd of its routes. 
# We just need one agent, because that way we can easily illustrate 
# the mechanisms and how an agent perception about routes would change by 
# considering different thetas we just need to use one agent.
# It is an agent belonging to the representative OD that we are gonna talk
# about in the following plots
# We want the agent to be in the middle of the simulation horizon

agent_last_episode_info <- data$bm_results |> 
  mutate(
    reliability_sensitivity = as.numeric(reliability_sensitivity)
  ) |> 
  filter(
    reliability_sensitivity == 0,
    seed == random_seed
  ) |> 
  select(-c(run_id, seed, warm_up, reliability_sensitivity)) |> 
  filter(
    agent_id == agent$agent_id,
    episode == max(episode)
  ) |> 
  select(
    episode, route_id, PT, sigma_r
  )
```

```{r}
#| label: theta_values

theta_values <- as.numeric(data$demand_odt$reliability_sensitivity |> unique())
```

```{r}
#| label: data_plot1_lines

# Intuition about the plot:
# 1) I wanna see which route is the best for each theta value.
# 2) Draw the contour lines for each theta value. In particular, draw the
# contour corresponding to the minimum (best) route cost. This shows which
# other combinations of values result in the same cost as the best route.

# map_dfr: runs the function once for each value in theta_values, and stacks the three results into one dataframe

iso_cost_lines_tmp <- map_dfr(theta_values, function(theta){
  # Choose route with the minimum generalized cost
  best <- agent_last_episode_info |> 
    mutate(
      cost = PT + theta * sigma_r
    ) |> 
    slice_min(cost, n = 1)
  
  # Line when we do not take into account the variability.
  # Basically the contour line is vertical because variability does not matter for the generalized cost
  if (theta == 0){
    tibble(theta, type = "vertical", xintercept = best$PT)
  } else{
    # (σ_r = -PT_r/θ + cost/θ, i.e. slope -1/θ, intercept cost/θ)
    tibble(theta, type = "line", slope = -1 / theta, intercept = best$cost / theta)
  }
})

iso_cost_lines <- iso_cost_lines_tmp |> mutate(
  theta = factor(theta)
)
```



```{r}
#| label: plot_1
ggplot(
  data = agent_last_episode_info, 
  mapping = aes(x = PT, y = sigma_r)
) +
  geom_point(size = 3) +
  geom_vline(
    data = filter(iso_cost_lines, type == "vertical"),
    mapping = aes(
      xintercept = xintercept, 
      color = theta),
      linetype = "dashed"
  ) +
  geom_abline(
    data = filter(iso_cost_lines, type == "line"),
    mapping = aes(
      slope = slope, 
      intercept = intercept, 
      color = theta), 
    linetype = "dashed"
  ) +
  ggrepel::geom_text_repel(aes(label = route_id)) +
  labs(
    x = "Perceived travel time", 
    y = "Perceived std. dev.", 
    color = expression(theta)
  )
```


