---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ9:** Analyze how the memory level of agents affects route-choice concentration and diversity in the BM model.

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

**The plots to answer this question will be:**

1. Boxplot of route-choice diversity across memory levels (one seed)
2. Sensitivity analysis across seeds

For more details, check the research questions document.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(glue)
library(ggtext)
library(patchwork)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
congestion_regime_tiers <- c("Near free-flow", "Moderate congestion", "High congestion")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```

```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ9","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
data <- list.files(data_dir, pattern="\\.parquet$", full.names = TRUE) |> 
  (\(x) set_names(x, tools::file_path_sans_ext(basename(x))))() |> 
  map(read_parquet)
```



```{r}
#| label: od_routes

# Data structure that stores for each seed, the ODs and their routes
seed_od_routes <- data$od_routes |> 
  select(-run_id) |> 
  filter(n_agents=='2000') |> 
  select(-n_agents) |> 
  group_by(seed, origin, dest, route_id) |> 
  slice_head(n = 1) |> 
  ungroup() |> 
  select(seed, origin, dest, route_id)

# Check to make sure all ODs across all different seeds have at least 2 routes
n_OD_with_1_route <- seed_od_routes |> 
  group_by(seed, origin, dest) |> 
  summarise(
    n = n()
  ) |>
  ungroup() |> 
  filter(n == 1) |> 
  nrow()

if (n_OD_with_1_route>0){
  stop("Some OD pair of some seed has only one route in their route set.")
}
```

```{r}
#| label: flow_od_routes

flow_routes <- data$flow_paths |> 
  filter(n_agents == '2000') |> 
  select(-run_id, -n_agents) |> 
  group_by(seed, memory_level) |> 
  filter(
    episode == max(episode)
  ) |> 
  ungroup() |> 
  group_by(
    origin, destination, path, seed, memory_level
  ) |> 
  summarise(
    flow = sum(count)
  ) |> 
  rename(
    route_id = path
  ) |> 
  ungroup() |> 
  mutate(
    memory_level = as.numeric(memory_level)
  ) |> 
  select(seed, memory_level, origin, destination, route_id, flow)
```

```{r}
#| label: demand_per_od

demand_per_od <- flow_routes |> 
  filter(
    memory_level == min(memory_level)
  ) |> 
  select(-memory_level) |> 
  group_by(origin, destination, seed) |> 
  summarise(
    demand = sum(flow)
  ) |> 
  select(seed, origin, destination, demand) |> 
  ungroup()
```

```{r}
#| label: n_alternatives_per_OD

n_alternatives <- seed_od_routes |> 
  group_by(seed, origin, dest) |> 
  summarise(
    n_alternatives = n()
  ) |> 
  rename(
    destination = dest
  ) |> 
  ungroup()
```


```{r}
#| label: path_share

path_share <- flow_routes |> 
  left_join(demand_per_od, by = c("seed", "origin", "destination")) |> 
  mutate(
    share = flow / demand
  )
```

```{r}
#| label: n_eff_routes

df_n_eff_routes <- path_share |> 
  mutate(
    log_share = log(share)
  ) |> 
  mutate(
    term_i = share * log_share
  ) |> 
  group_by(
    seed, memory_level, origin, destination
  ) |> 
  summarise(
    entropy = -sum(term_i),
    check = sum(share)
  ) |> 
  mutate(
    n_eff_routes = exp(entropy)
  ) |> 
  left_join(
    n_alternatives, by = c("seed", "origin", "destination")
  ) |> 
  mutate(
    normalized_n_eff_routes = (n_eff_routes - 1) / (n_alternatives - 1)
  ) |> 
  ungroup() 
```

```{r}
#| label: plot1

# We only take into account one seed
random_seed <- 8
df_plot_1 <- df_n_eff_routes |> filter(seed == random_seed)

ggplot(
  data = df_plot_1,
  mapping = aes(
    x = factor(memory_level),
    y = normalized_n_eff_routes
  )
) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(
  width = 0.125,
  alpha = 0.5,
  size = 0.8
  ) +
  labs(
    x = "Memory level",
    y = "Normalized nº effective routes",
    title = "Route-choice diversity",
    subtitle = glue("Normalized effective number of routes across OD pairs \nSeed = {random_seed} | High congestion (2000 veh)")
  )

ggsave(here("r","RQ9","figures","plot1_boxplot_n_eff_route.png"))
```





