---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ2:** Analyze whether reducing the memory level of agents
  prevents the BM model from converging towards a DUE state

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

**The plots to answer this question will be:**

1. Aggregated R-gap for different memory levels across episodes (representative learning dynamics)
2. Sensitivity analysis
3. Variability across seeds

For more details, check the research questions document.



```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(glue)
library(ggtext)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
congestion_regime_tiers <- c("Near free-flow", "Moderate congestion", "High congestion")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```


```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ2","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
data <- list.files(data_dir, pattern="\\.parquet$", full.names = TRUE) |> 
  (\(x) set_names(x, tools::file_path_sans_ext(basename(x))))() |> 
  map(read_parquet)
```

```{r}
#| label: representative_seeds_spaghetti_plots

# Randomly select 3 seeds (seeded so the same 3 - and their colors - are
# picked every time this re-renders)
set.seed(123)
representative_seeds <- sample(unique(data$bm_rgap$seed), size = 3)

labels_seeds <- c("S1", "S2", "S3")

color_lines <- c(
  "1.0" = "#0072B2",  # blue
  "0.6" = "#E69F00",  # orange
  "0.2" = "#009E73"  # bluish green
) 
```

```{r}
#| label: get_warm_up_period(exploration)

# It's not perfectly precise — the true per-agent start is episode > 24 and all routes visited, so some agents start a few episodes later than 24. But it's the honest guaranteed floor ("no agent has started learning before this point"), it's already computed for every run via config.warm_up, and given the plot already treats each seed's trajectory independently without forcing alignment (per the existing code comment about seeds finishing at very different episode counts), a single shared floor line is consistent with that treatment rather than adding seed-specific precision the rest of the plot doesn't have either.

exploration_phase_floor <- as.numeric(unique(data$bm_rgap$warm_up))
```

## Network equilibrium

```{r}
#| label: data_plot1

representative_memory_levels <- c("1.0","0.6","0.2")
representative_memory_levels_labels <- map_chr(representative_memory_levels, \(x) paste0("Memory = ", x))

# Not aggregating mean +/- SD across seeds here: seeds finish training at very
# different episode counts (e.g. High regime: some stop by episode ~22, one
# runs to 129), so for a large share of each panel's episode range the "mean"
# would silently collapse to a single seed's raw value (SD undefined). Each
# seed's own trajectory is plotted instead - no cross-seed averaging, so no
# misleading shrinking-N problem.

rgap_df <- data$bm_rgap |>
  mutate(
    congestion_regime = make_congestion_regime(n_agents),
    ) |>
  filter(
    seed %in% representative_seeds
  ) |> 
  mutate(
    seed_fct = factor(seed, levels = representative_seeds, labels = labels_seeds)
  ) |> 
  filter(
    memory_level %in% representative_memory_levels
  ) |> 
  mutate(
    memory_level_fct = factor(memory_level, levels = representative_memory_levels, labels = representative_memory_levels_labels) 
  ) |> 
  mutate(memory_level_fct = fct_rev(memory_level_fct))
```

```{r}
#| label: breaks_plot1

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
step <- floor(max_episode / 30) * 10
breaks <- seq(0, max_episode, by = step)

# Remove breaks that are within step of the learning start
breaks <- breaks[abs(breaks - exploration_phase_floor) >= (step/2.25)]
# Add the learning start break
breaks <- sort(c(breaks, exploration_phase_floor))
labels <- as.character(breaks)
labels[breaks==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```


```{r}
#| label: plot1

upper_limit_y_axis <- quantile(rgap_df$rgap, probs = c(0.95))

ggplot(
  data = rgap_df,
  mapping = aes(x = episode, y = rgap, group = seed_fct)
) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = "red",
    alpha = 0.08
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.5
  ) +
  geom_line(alpha = 0.6, linewidth = 0.4) +
  geom_hline(yintercept = 0, alpha = 0.6, linewidth = 0.7, color = "#7D7D7D") +
  geom_hline(
    yintercept = 5,
    colour = "grey40",
    linetype = "longdash",
    linewidth = 0.5
  ) +
  facet_grid(memory_level_fct ~ congestion_regime) +
  labs(
    title = "R-gap evolution across episodes",
    subtitle = "Representative learning trajectories for three memory levels and three random seeds (BM)",
    x = "Episode",
    y = "R-gap (%)"
  ) +
coord_cartesian(ylim=c(-5,upper_limit_y_axis)) +
   scale_x_continuous(breaks = breaks, labels = labels) +
theme(
  axis.title.x = element_text(margin = margin(t = 1)),
  axis.text.x = element_text(angle = 30, hjust = 1),
  strip.text.y = element_text(angle = 0, size = 10)
)

ggsave(here("r","RQ2","figures","plot1_rgap_evolution_mem_seed.png"), width = 12, height = 6)
```

## Interpretation plot 1

- **Oscillatory behavior**:
There is no evidence that decreasing memory produces oscillatory behavior.

- **DUE convergence**:
All memory levels eventually approach similar low R-gap values (DUE state)

- **Behavioral convergence**:
Higher memory levels lead to earlier stabilization of the agents policies.

- **Robustness across seeds**:
Similar results are obtained across seeds.

Hence memory has relatively influence on the evolution of the aggregated R-gap, but it has a clear influence on behavioral convergence.

Intuitively, higher memory enables agents to accumulate knowledge over time, leading to earlier stabilization of route-choice policies. Lower memory places greater emphasis on recent experiences, causing policies to continue adapting for more episodes.

```{r}
#| label: data_plot2
#| eval: false

sensitivity_df <- data$bm_rgap |> 
  mutate(
    memory_level = as.numeric(memory_level),
    congestion_regime = make_congestion_regime(n_agents),
  ) |> 
  select(-c(n_agents)) |> 
  group_by(seed, congestion_regime, memory_level) |> 
  slice_max(episode, n = 1) |> 
  ungroup() 

sensitivity_df_summary <- sensitivity_df |> 
  group_by(memory_level, congestion_regime) |> 
  summarise(
    median_rgap = median(rgap)
  ) |> 
  ungroup()

```

```{r}
#| label: plot2
#| eval: false

ggplot(
  data = sensitivity_df_summary,
  mapping = aes(
    x = memory_level,
    y = median_rgap
  )
) +
  geom_point() +
  geom_line(linewidth=1.4) +
  geom_line(
    data = sensitivity_df,
    mapping = aes(
      x = memory_level,
      y = rgap,
      group = seed
    ),
    linewidth = 0.45,
    alpha = 0.6,
    colour = "grey60"
  ) +
  facet_wrap(~congestion_regime) +
  labs(
    title = "Sensitivity of R-gap to memory level",
    subtitle = "Final-episode R-gap for five random seeds (thin lines) \nand their median (thick line)",
    x = "Memory level",
    y = "R-gap (%)"
  ) + 
  scale_x_continuous(
    breaks = unique(sensitivity_df$memory_level)
  )

ggsave(here("r","RQ2","figures","plot2_sensitivity_analysis_memory.png"), width = 12, height = 6)
```

The second plot relative to the sensitivity of the r-gap to the memory level, 
is not included. When there is no sensitivity, it's enough with the first plot. The second plot is only necessary in the case that there is sensitivity.


## Behavioral convergence
```{r}
#| label: data_plot_3_route_flow

# Things to take into account:
# 1. OD matrix is different across congestion_regimes (keeping everything else constant)
# 2. OD matrix is different across seeds (keeping everything else constant)
# 3. OD matrix is the same across memory levels (keeping everything else constant)

# When analyzing the impact of memory level on the route flow evolution, we just want to analyze one OD from a FIXED OD MATRIX!!!
###########################
# Hence to choose our representative OD we have to pick:
# #########################
# 1. representative_seed
representative_seed <- representative_seeds[1]
# 2. representative_congestion
representative_congestion <- 2000
representative_congestion_label <- "High congestion (2000)"
# 3. random_memory_level
set.seed(123)
random_memory_level <- sample(unique(data$demand_odt$memory_level), size = 1)

# To choose a representative OD, basically we are going to select an OD with a high number of available routes and a high demand. So we need two things:

# 1. OD matrix (contains the demand on each OD for this scenario)
OD_matrix <- data$demand_odt |> 
  filter(
    memory_level == random_memory_level,
    seed == representative_seed, 
    n_agents == representative_congestion
  ) |> 
  select(-c(run_id, warm_up, memory_level)) |> 
  group_by(origin, destination, seed, n_agents) |> 
  summarise(demand = sum(count)) |> 
  ungroup() |> 
  rename(
    dest = destination
  )

# 2. OD_routes (available routes for each OD)
OD_routes <- data$od_routes |> 
  filter(
    memory_level == random_memory_level,
    seed == representative_seed, 
    n_agents == representative_congestion
  ) |> 
  select(-c(run_id, seed, memory_level, n_agents, warm_up)) |> 
  group_by(origin, dest) |> 
  summarise(
    n_routes = n_distinct(route_id)
  ) |> 
  ungroup()

# JOIN: od_matrix + od_routes
OD_info <- OD_matrix |> 
  left_join(OD_routes, by = c("origin", "dest"))

representative_OD <- OD_info |> 
  arrange(desc(n_routes),desc(demand)) |> 
  slice_head(n = 1) |> 
  select(-c(seed, n_agents))
```

```{r}
#| label: data_plot_3_route_flow_2

# Now that we have the scenario (defined by the seed and congestion_regime) 
# and we also have the representative OD whose paths we want to analyze, 
# we need to do the following:

# 1. Compute the flow on each route of the representative OD per episode
# Efficiently dealing with large dataframes

# 1. Partition parquet file
pq_path_flow_paths <- here("r/RQ2/data/flow_paths_partitioned")
data$flow_paths |> 
  group_by(path) |> 
  write_dataset(path = pq_path_flow_paths, format = "parquet")

# 2. List partitioned parquet files
tibble(
  files = list.files(pq_path_flow_paths, recursive = TRUE),
  size_MB = file.size(file.path(pq_path_flow_paths, files)) / 1024^2
)

# 3. Use dplyr with arrow
flow_paths_pq <- open_dataset(pq_path_flow_paths)
query <- flow_paths_pq |> 
  filter(
    # Filter by the particular scenario
    seed == representative_seed, 
    n_agents == representative_congestion,
    memory_level %in% representative_memory_levels,
    # Filter by the particular OD
    origin == representative_OD$origin,
    destination == representative_OD$dest,
  ) |> 
  group_by(episode, path, memory_level) |> 
  summarise(
    flow = sum(count)
  ) |> 
  ungroup()

flow_paths_df_query <- query |> collect()

# Check: The sum of flows across paths for the particular OD for one episode should be equal to the demand on that OD
check_demand_od_df <- flow_paths_df_query |> 
  group_by(episode, memory_level) |> 
  summarise(
    demand = sum(flow)
  ) |> ungroup()

demand_od_sum_paths <- unique(check_demand_od_df$demand)
demand_od <- representative_OD$demand

if (demand_od != demand_od_sum_paths) {
  stop("Something is wrong in the computation of path flows for the particular OD and scenario.")
}
```

```{r}
#| label: data_plot_3_a_part_3

# Now we still have to manipulate a bit the flow paths on the particular OD and scenario, so that is properly visualized in the plot

# 1. Handling some things to convert path variable to factor
path_levels <- as.character(sort(unique(flow_paths_df_query$path)))
path_levels_labels <- as.character(sort(unique(flow_paths_df_query$path + 1)))

flow_paths_df <- flow_paths_df_query |> 
  mutate(
    # Convert to factors
    congestion_regime = representative_congestion_label,
    memory_level_fct = forcats::fct_rev(factor(
      memory_level, 
      levels = representative_memory_levels, 
      labels = representative_memory_levels_labels)),
    path = factor(
      path, 
      levels = path_levels,
      labels = path_levels_labels
    )
  ) |> 
  select(-c(memory_level))
```

```{r}
#| label: breaks_plot3_a
#| eval: False

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
step_plot3 <- floor(max_episode / 50) * 10
breaks_plot3 <- seq(0, max_episode, by = step_plot3)

# Remove breaks that are within step of the learning start
breaks_plot3 <- breaks[abs(breaks - exploration_phase_floor) >= step_plot3 / 2]
# Add the learning start break
breaks_plot3 <- sort(c(breaks_plot3, exploration_phase_floor))
labels_plot3 <- as.character(breaks_plot3)
labels_plot3[breaks_plot3==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```

```{r}
#| label: plot3_a

# This plots refers to a particular OD, for a particular seed and congestion regime, and facet for some representative memory_levels. 

ggplot(
  data = flow_paths_df, 
  mapping = aes(x = episode, y = flow, group = path, color = path)
) +
  geom_line(linewidth = 0.7) +
  facet_grid(memory_level_fct ~ .) +
  labs(
    color = "Route",
    y = "Route flow",
    x = "Episode",
    title = "Behavioral convergence: Evolution of route flows",
    subtitle = glue(
      "{representative_congestion_label} | Seed {representative_seed} | OD pair: {representative_OD$origin} → {representative_OD$dest}"
    )
  ) +
  # Shaded region
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = "red",
    alpha = 0.06
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.4
  ) +
  annotate(
    "text",
    x = (exploration_phase_floor - 2) / 2,
    y = Inf,
    label = "pre-learning \nphase",
    fontface = "bold",
    colour = "firebrick",
    vjust = 2.5,
    size = 3.5
  ) +
  # Change color scale
  scale_color_brewer(palette = "Dark2") +
  # Tweak miscellaneous stuff
  theme(
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10),
    legend.key.size = unit(0.6, "lines"),
    plot.subtitle = element_text(size = 11)
) 

ggsave(here("r","RQ2","figures","plot3_route_flow_evolution.png"), width = 12, height = 6)
```

```{r}
#| label: data_plot3_b

# Plot relative to the mean travel time of the routes, to check
# redistribution occurs among near-equilibrium routes

# 1. Compute the mean travel time on each route per episode
# Efficiently dealing with large dataframes

# 1. Partition parquet file
pq_path_cost_paths <- here("r/RQ2/data/cost_paths_partitioned")
data$cost_paths |> 
  group_by(path) |> 
  write_dataset(path = pq_path_cost_paths, format = "parquet")

# 2. List partitioned parquet files
tibble(
  files = list.files(pq_path_cost_paths, recursive = TRUE),
  size_MB = file.size(file.path(pq_path_cost_paths, files)) / 1024^2
)

# 3. Use dplyr with arrow
cost_paths_pq <- open_dataset(pq_path_cost_paths)
query <- cost_paths_pq |> 
  filter(
    memory_level %in% representative_memory_levels,
    seed == representative_seed, 
  ) |> 
  group_by(episode, path, memory_level, n_agents) |> 
  summarise(
    flow = sum(count)
  ) |> 
  ungroup()

flow_paths_df_query <- query |> collect()

```


