---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ4:** Analyze whether the BM model is capable of
  readapting and recovering the initial equilibrium traffic state after
  a temporary degradation of link capacity

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

**The plots to answer this question will be:**

1. R-gap and Mean Travel Time evolution
2. Flow on disrupted edge
3. Path share evolution

For more details, check the research questions document.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(patchwork)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
congestion_regime_tiers <- c("Moderate", "High")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```


```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ4","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
data <- list.files(data_dir, pattern="\\.parquet$", full.names = TRUE) |> 
  (\(x) set_names(x, tools::file_path_sans_ext(basename(x))))() |> 
  map(read_parquet)
```

```{r}
#| label: data_plot_1_a

disruption_start <- as.numeric(unique(data$bm_rgap$degradation_start_episode))
disruption_end <- as.numeric(unique(data$bm_rgap$degradation_end_episode))

rgap_df <- data$bm_rgap |> 
  mutate(
    congestion_regime = make_congestion_regime(n_agents)
  ) |> 
  select(-c(degradation_start_episode, degradation_end_episode, n_agents, network_degraded, run_id))

rgap_df_summary <- rgap_df |> 
  group_by(episode, congestion_regime) |> 
  summarise(
    median_rgap = median(rgap),
    iqr_rgap = IQR(rgap)
  ) |> 
  ungroup()
```

```{r}
#| label: plot_1_a

p1_a <- ggplot(
) +
  annotate(
    "rect",
    xmin = disruption_start,
    xmax = disruption_end,
    ymin = -Inf,
    ymax = Inf,
    fill = "red",
    alpha = 0.08,
  ) +
  annotate(
    "text",
    x = (disruption_start + disruption_end) / 2,
    y = Inf,
    label = "Disruption",
    fontface = "bold",
    colour = "firebrick",
    vjust = 2.5,
    size = 4
  ) +
  geom_vline(
  xintercept = disruption_start,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.5
  ) +
  geom_vline(
  xintercept = disruption_end,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.5
  ) +
  geom_line(
    data = rgap_df_summary,
    mapping = aes(x = episode, y = median_rgap),
  ) + 
  geom_ribbon(
    data = rgap_df_summary,
    mapping = aes(x = episode, ymin = median_rgap - iqr_rgap, ymax = median_rgap + iqr_rgap),
    alpha = 0.1
  ) +
  facet_wrap(~congestion_regime) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
) +
  labs(
    x = "Episode",
    y = "R-gap (%)"
  )
  
```

```{r}
#| label: data_plot_1_b

# Efficiently dealing with large dataframes

# 1. Partition parquet file
pq_path_trip_info <- here("r/RQ4/data/trips_info_partitioned")
data$bm_trips_info |> 
  group_by(seed) |> 
  write_dataset(path = pq_path_trip_info, format = "parquet")

# 2. List partitioned parquet files
tibble(
  files = list.files(pq_path_trip_info, recursive = TRUE),
  size_MB = file.size(file.path(pq_path_trip_info, files)) / 1024^2
)

# 3. Use dplyr with arrow
trips_info_pq <- open_dataset(pq_path_trip_info)
query <- trips_info_pq |> 
  mutate(
    duration = as.numeric(duration)
  ) |> 
  group_by(episode, seed, n_agents) |> 
  summarise(
    mean_travel_time = mean(duration)
  ) |> 
  ungroup()

mean_tt_df <- query |> collect() |> 
  mutate(
    congestion_regime = make_congestion_regime(n_agents)
  ) |> 
  select(-n_agents)

mean_tt_df_summary <- mean_tt_df |> 
  group_by(episode, congestion_regime) |> 
  summarise(
    median_mean_tt = median(mean_travel_time),
    iqr_mean_tt = IQR(mean_travel_time)
  ) |> 
  ungroup()
```

```{r}
#| label: plot_1_b

p1_b <- ggplot(
) +
  annotate(
    "rect",
    xmin = disruption_start,
    xmax = disruption_end,
    ymin = -Inf,
    ymax = Inf,
    fill = "red",
    alpha = 0.08,
  ) +
  annotate(
    "text",
    x = (disruption_start + disruption_end) / 2,
    y = Inf,
    label = "Disruption",
    fontface = "bold",
    colour = "firebrick",
    vjust = 2.5,
    size = 4
  ) +
  geom_vline(
  xintercept = disruption_start,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.5
  ) +
  geom_vline(
  xintercept = disruption_end,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.5
  ) +
  geom_line(
    data = mean_tt_df_summary,
    mapping = aes(x = episode, y = median_mean_tt),
  ) + 
  geom_ribbon(
    data = mean_tt_df_summary,
    mapping = aes(x = episode, ymin = median_mean_tt - iqr_mean_tt, ymax = median_mean_tt + iqr_mean_tt),
    alpha = 0.1
  ) +
  facet_wrap(~congestion_regime) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
) +
  labs(
    x = "Episode",
    y = "Mean Travel Time (s)"
  )
```

```{r}
#| label: plot_1

p1_a / p1_b +
  plot_annotation(
    title = "Adaptation to temporary network disruption",
    subtitle = "Evolution before, during, and after the disruption"
  )

ggsave(here("r","RQ4","figures","plot1_rgap_meantt_evolution.png"), width = 12, height = 6)
```

```{r}
#| label: data_plot_2

degraded_edges <- c("E10_15_NB", "E10_15_SB")

# Efficiently dealing with large dataframes

# 1. Partition parquet file
pq_path_edgedata <- here("r/RQ4/data/edgedata_partitioned")
data$bm_edgedata |> 
  group_by(seed) |> 
  write_dataset(path = pq_path_edgedata, format = "parquet")

# 2. List partitioned parquet files
tibble(
  files = list.files(pq_path_edgedata, recursive = TRUE),
  size_MB = file.size(file.path(pq_path_edgedata, files)) / 1024^2
)

# 3. Use dplyr with arrow
edgedata_pq <- open_dataset(pq_path_edgedata)
query <- edgedata_pq |> 
  # Filter only by degraded edges
  filter(edge %in% degraded_edges) 
# |> 
  # mutate(
  #   entered = as.numeric(entered)
  # ) |> 
  # group_by(episode, seed, n_agents) |> 
  # summarise(
  #   flow_edge = sum(entered)
  # ) |> 
  # ungroup()

flow_degraded_edges_df <- query |> collect()

```








