---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ4:** Analyze whether the BM model is capable of
  readapting and recovering the initial equilibrium traffic state after
  a temporary degradation of link capacity
  
Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

**The plots to answer this question will be:**

1. R-gap and Mean Travel Time evolution
2. Flow on disrupted edge
3. Path share evolution

For more details, check the research questions document.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(patchwork)
library(xml2)
library(glue)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
congestion_regime_tiers <- c("High")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```

```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ4","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
parquet_files <- c(
  flow_paths = "flow_paths.parquet",
  bm_rap = "bm_rgap.parquet",
  demand_odt = "demand_odt.parquet",
  od_routes = "od_routes.parquet"
)

data <- parquet_files |>
  purrr::set_names() |>
  purrr::map(\(f) arrow::read_parquet(file.path(data_dir, f)))
```

```{r}
#| label: get_warm_up_period(exploration)

# It's not perfectly precise — the true per-agent start is episode > 24 and all routes visited, so some agents start a few episodes later than 24. But it's the honest guaranteed floor ("no agent has started learning before this point"), it's already computed for every run via config.warm_up, and given the plot already treats each seed's trajectory independently without forcing alignment (per the existing code comment about seeds finishing at very different episode counts), a single shared floor line is consistent with that treatment rather than adding seed-specific precision the rest of the plot doesn't have either.

exploration_phase_floor <- as.numeric(unique(data$bm_rgap$warm_up))
```

```{r}
#| label: data_plot3_prerequisites

# Infer the degraded edges (with their NB/SB/EB/WB direction suffixes)
# straight from the network files instead of hardcoding them: an edge
# counts as degraded if its lane speed differs between the baseline
# network and the one config.network_degraded was swapped in with.
network_degraded_path <- unique(data$bm_rgap$network_degraded)
stopifnot(length(network_degraded_path) == 1)
network_normal_path <- file.path(dirname(network_degraded_path), "Normal_Sioux_Falls.net.xml")

get_lane_speeds <- function(net_xml_path) {
  # Read the XML file and find all <lane> elements
  lanes <- xml_find_all(read_xml(net_xml_path), "//lane")
  
  # Extract the edge ID by removing the lane suffix (_0, _1, etc.)
  edge_id <- sub("_[0-9]+$", "", xml_attr(lanes, "id"))
  
  # Extract lane speeds, convert them to numeric,
  # and use edge IDs as the names of the vector
  set_names(as.numeric(xml_attr(lanes, "speed")), edge_id)
}

normal_speeds <- get_lane_speeds(network_normal_path)
degraded_speeds <- get_lane_speeds(network_degraded_path)
degraded_edges <- names(degraded_speeds)[degraded_speeds != normal_speeds[names(degraded_speeds)]]
```

```{r}
#| label: representative_memory_levels

representative_memory_levels <- c("1", "0.1")
representative_memory_levels_labels <- map_chr(representative_memory_levels, \(x) paste0("Memory = ", x))
```

```{r}
#| label: data_plot3

# 1. Choose a representative OD pair:
# (i) it has relatively high travel demand
# (ii) it possesses several feasible alternative routes
# (iii) at least one of its frequently used routes traverses the disrupted edge while alternative routes bypass it.

# # Things to take into account:
# 1. OD matrix is different across congestion_regimes (keeping everything else constant)
# 2. OD matrix is different across seeds (keeping everything else constant)

###########################
# Hence to choose our representative OD we have to pick:
# #########################
# 1. representative_seed
set.seed(123)
representative_seed <- sample(data$bm_rgap.parquet$seed, size = 1)
# 2. representative_congestion
representative_congestion <- 2000
representative_congestion_label <- "High congestion (2000)"
# 3. Random memory level (OD matrix doesnt vary with the memory level)
random_memory_level <- sample(unique(data$demand_odt$memory_level), size = 1)

# We are gonna follow the criteria described above to choose a representative OD.
# Hence we need three things:
# 1. OD matrix (contains the demand on each OD for this seed and congestion)
OD_matrix <- data$demand_odt |> 
  filter(
    seed == representative_seed, 
    n_agents == representative_congestion,
    memory_level == random_memory_level,
  ) |> 
  select(-c(run_id, warm_up, memory_level)) |> 
  group_by(origin, destination, seed, n_agents) |> 
  summarise(demand = sum(count)) |> 
  ungroup() |> 
  rename(
    dest = destination
  )

# Check to ensure everything is ok
if (representative_congestion==2000){
  sum_demand <- sum(OD_matrix$demand)
  if (sum_demand != 1847){
    stop("Review the OD_matrix data wrangling")
  }
}

# 2. OD_routes (available routes for each OD)
OD_routes <- data$od_routes |> 
  filter(
    memory_level == random_memory_level,
    seed == representative_seed, 
    n_agents == representative_congestion
  ) |> 
  select(-c(run_id, seed, n_agents, warm_up, memory_level)) |> 
  group_by(origin, dest)

OD_routes_summary <- OD_routes |> 
  summarise(
    n_routes = n_distinct(route_id)
  ) |> 
  ungroup() 

# 3. OD_routes_disrupted (which routes are disrupted and which unaffected for all the ODs)
OD_routes_disrupted <- data$od_routes |> 
  filter(
    memory_level == random_memory_level,
    seed == representative_seed, 
    n_agents == representative_congestion
  ) |> 
  select(-c(run_id, seed, n_agents, warm_up, network_degraded, degradation_start_episode, degradation_end_episode, memory_level)) |> 
  group_by(origin, dest, route_id) |> 
  summarise(
    disrupted = any(edge %in% degraded_edges)
  ) |> 
  ungroup() |> 
  group_by(origin, dest) |> 
  summarise(
    n_disrupted_routes = sum(disrupted)
  ) |> 
  ungroup()

representative_OD_df <- 
  OD_matrix |> 
  left_join(OD_routes_summary, by = c("origin", "dest")) |>
  left_join(OD_routes_disrupted, by = c("origin", "dest")) |> 
  select(-c("seed", "n_agents")) |> 
  filter(n_routes > n_disrupted_routes) |> 
  filter(n_disrupted_routes > 0) |> 
  arrange(desc(demand), desc(n_routes), desc(n_disrupted_routes))

representative_OD <- representative_OD_df |> 
  slice_head(n = 1)
```

```{r}
#| label: data_plot3_part_2
# 2. Once the representative OD has been chosen, the next step is to check the paths 
# for the OD and create a column that mark which paths are disrupted and which unaffected

# OD_routes <- read_parquet("/home/miguel/6.Projects/Thesis/data/environment/od_routes.parquet")

OD_routes_classified <- OD_routes |> 
  filter(origin == representative_OD$origin, dest == representative_OD$dest) |> 
  group_by(origin, dest, route_id) |> 
  summarise(
    disrupted = any(edge %in% degraded_edges)
  ) |> 
  ungroup()
```

```{r}
#| label: data_plot3_part_3

# 3. Compute the flow on each route of the representative OD per episode

# Efficiently dealing with large dataframes

# 1. Partition parquet file
pq_path_flow_paths <- here("r/RQ4/data/flow_paths_partitioned")

# Make sure folder is empty before writing, so we have a clean slate,
# and dont gets mixed with previous stuff
if (dir.exists(pq_path_flow_paths)) {
  unlink(pq_path_flow_paths, recursive = TRUE, force = TRUE)
}

dir.create(pq_path_flow_paths, recursive = TRUE)

data$flow_paths |> 
  group_by(path) |> 
  write_dataset(path = pq_path_flow_paths, format = "parquet")

# 2. List partitioned parquet files
tibble(
  files = list.files(pq_path_flow_paths, recursive = TRUE),
  size_MB = file.size(file.path(pq_path_flow_paths, files)) / 1024^2
)

# 3. Use dplyr with arrow
flow_paths_pq <- open_dataset(pq_path_flow_paths)
query <- flow_paths_pq |> 
  filter(
    # Filter by the particular scenario
    memory_level %in% representative_memory_levels,
    seed == representative_seed, 
    n_agents == representative_congestion,
    # Filter by the particular OD
    origin == representative_OD$origin,
    destination == representative_OD$dest,
  ) |> 
  group_by(episode, path, memory_level) |> 
  summarise(
    flow = sum(count)
  ) |> 
  ungroup()

flow_paths_df_query <- query |> collect()

# Check: The sum of flows across paths for the particular OD for one episode should be equal to the demand on that OD
check_demand_od_df <- flow_paths_df_query |> 
  group_by(episode, memory_level) |> 
  summarise(
    demand = sum(flow)
  ) |> ungroup()

demand_od_sum_paths <- unique(check_demand_od_df$demand)
demand_od <- representative_OD$demand

if (demand_od != demand_od_sum_paths) {
  stop("Something is wrong in the computation of path flows for the particular OD and scenario.")
}
```

```{r}
#| label: data_plot_3_part_4

# 4. Transforming df 

# a) Necessary stuff to convert path variable to factor
path_levels <- as.character(sort(unique(flow_paths_df_query$path)))
path_levels_labels <- as.character(sort(unique(flow_paths_df_query$path + 1)))

# b) Prepare df to add information regarding route is disrupted or unaffected
OD_routes_classified_transformed <- OD_routes_classified |> 
  select(-c(origin, dest)) |> 
  rename(path = route_id)

flow_paths_df <- flow_paths_df_query |> 
  left_join(OD_routes_classified_transformed, by = "path") |> 
  mutate(
    # 1. Convert to factors
    congestion_regime = representative_congestion_label,
    path = factor(
      path, 
      levels = path_levels,
      labels = path_levels_labels
    )
  ) |> 
    # 2. Compute flow in relative terms
  mutate(
    flow = flow / demand_od
  ) |> 
  mutate(
    memory_level_fct = factor(memory_level, levels = representative_memory_levels, labels = representative_memory_levels_labels) 
  ) |> 
  mutate(memory_level_fct = fct_rev(memory_level_fct))
  

# Check flow share is correct
if(!all(flow_paths_df |> group_by(episode, memory_level_fct) |> summarise(total=sum(flow)) |> pull(total) == "1")){
  stop("There is some error with the relative flow computation")
}
```

```{r}
#| label: breaks_plot3

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
breaks <- seq(0,max_episode, by = 75)

# Remove breaks that are within step of the learning start
breaks <- breaks[abs(breaks - exploration_phase_floor) >= (50)]
# Add the learning start break
breaks <- sort(c(breaks, exploration_phase_floor))
labels <- as.character(breaks)
labels[breaks==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```


```{r}
#| label: disruption_episodes
disruption_start <- as.numeric(unique(data$bm_rgap$degradation_start_episode))
disruption_end <- as.numeric(unique(data$bm_rgap$degradation_end_episode))
```

```{r}
#| label: Determine_relevant_paths_plot3

# On this plot, we only wanna focus on the 
# paths whose path share changes considerably 
# because of the disruption. 
# For the paths that do not vary, we are not interested in them
# So we will set alpha for those paths that are irrelevant
# To detect in code those paths whose flow vary because of disruption we will use the following rule:
# RULE:
# Median flow last 10 episodes before disruption
# minus
# Median flow last 10 episodes during disruption 
# Filter by those that change > 5%.

median_flow_paths_before <- flow_paths_df |> 
  filter(
    episode>=disruption_start - 10,
    episode<disruption_start
  ) |> 
  group_by(path, memory_level) |> 
  summarise(median_flow_before = median(flow))

median_flow_paths_during <- flow_paths_df |> 
  filter(
    episode>= disruption_end - 10,
    episode<disruption_end
  ) |> 
  group_by(path, memory_level) |> 
  summarise(median_flow_during = median(flow))

all_paths_memory_levels <- tidyr::crossing(
  path = unique(flow_paths_df$path),
  memory_level = representative_memory_levels
)

relevant_paths_plot3 <- all_paths_memory_levels |> 
  left_join(median_flow_paths_before, by = c("path", "memory_level")) |> 
  left_join(median_flow_paths_during, by = c("path", "memory_level")) |> 
  mutate(
    median_flow_before = coalesce(median_flow_before, 0),
    median_flow_during = coalesce(median_flow_during, 0),
    diff = abs(median_flow_during - median_flow_before)
  ) |> 
  mutate(
    relevant = diff > 0.1
  ) |> 
  select(-c(median_flow_before, median_flow_during, diff))

flow_paths_df_plot <- flow_paths_df |> 
  left_join(relevant_paths_plot3, by = c("path", "memory_level")) |> 
  select(-memory_level)
```

```{r}
#| label: plot3

path_labels_relevant <- flow_paths_df_plot |> 
  group_by(path, memory_level_fct) |> 
  filter(episode == max(episode, na.rm = TRUE)) |> 
  mutate(
    path_type = if_else(disrupted, "Disrupted", "Unaffected")
  ) |> 
  filter(relevant)

# This plots refers to a particular OD, for a particular seed and congestion regime
upper_y_axisx <- ceiling(max(flow_paths_df$flow) * 100 %/% 10) / 10


ggplot(
  data = flow_paths_df_plot,
  mapping = aes(x = episode, y = flow)
) +
  # Lines
  geom_line(
    mapping = aes(color = path, linewidth = relevant)
  ) +
  scale_linewidth_manual(
    values = c("FALSE" = 0.2, "TRUE" = 0.9),
    guide = 'none'
  ) +
  # Path type labels
  ggrepel::geom_text_repel(
    data = path_labels_relevant,
    mapping = aes(
      x = episode,
      y = flow,
      label = path_type,
      color = path
    ),
    size = 4,
    show.legend = FALSE,
    nudge_x = 10,
    box.padding = 0.2,
    point.padding = 1,
    max.overlaps = Inf,
    hjust = 1
  ) +
  # Pre-learning shaded region
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = "red",
    alpha = 0.06
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.4
  ) +
  scale_x_continuous(breaks = breaks, labels = labels) +
  # Disruption phase shaded region
  annotate(
    "rect",
    xmin = disruption_start,
    xmax = disruption_end,
    ymin = -Inf,
    ymax = Inf,
    fill = "orange",
    alpha = 0.20,
  ) +
  annotate(
    "text",
    x = (disruption_start + disruption_end) / 2,
    y = Inf,
    label = "Disruption",
    fontface = "bold",
    colour = "firebrick",
    vjust = 2.5,
    size = 4
  ) +
  geom_vline(
  xintercept = disruption_start,
  linetype = 2,
  alpha = 0.5,
  color = "orange",
  linewidth = 0.5
  ) +
  geom_vline(
  xintercept = disruption_end,
  linetype = 2,
  alpha = 0.5,
  color = "orange",
  linewidth = 0.5
  ) +
  # Change y-axis to %
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, upper_y_axisx),
    breaks = seq(0, 1, 0.2)
  ) +
  # Change color scale
  scale_color_brewer(palette = "Dark2", ) +
  theme(
    axis.text.x = element_text(size = 10),
  ) +
  labs(
    color = "Route",
    x = "Episode",
    y = "Path share (%)",
    title = "Path-share evolution during a temporary network disruption",
    subtitle = glue(
      "{representative_congestion_label} | Seed {representative_seed} | OD pair: {representative_OD$origin} → {representative_OD$dest}"
  )
) +
  facet_grid(memory_level_fct ~ .)

ggsave(here("r","RQ4","figures","plot3_flow_share_evolution_paths.png"), width = 12, height = 6)

```




