---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "11/aug/2026"
format:
  html:
    self-contained: true
execute:
  warning: false
  message: false
editor_options:
  chunk_output_type: console
---

**RQ8:** Quantitative comparison of link flows produced by Bush-Mosteller vs
`duaIterate`, across memory levels.

Split out of RQ7 (which stays a single-combination qualitative case study at
the base-config default `memory_level=1`) so this sweep can vary
`memory_level` and replicate across seeds without touching RQ7's qualitative
artifacts — see `experiments/rq8/design.yaml`.

For more details, check the research questions document.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(patchwork)
library(xml2)
library(glue)
library(gt)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
congestion_regime_tiers <- c("High")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```

```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ8","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
parquet_files <- c(
  od_routes_bm = "od_routes_bm.parquet",
  od_routes_dua = "od_routes_dua.parquet",
  flows_bm = "flows_bm.parquet",
  flows_dua = "flows_dua.parquet"
)

data <- parquet_files |>
  purrr::set_names() |>
  purrr::map(\(f) arrow::read_parquet(file.path(data_dir, f)))
```

```{r}
#| label: representative_memory_levels

representative_memory_levels <- c("1.0", "0.5" ,"0.01")
```

```{r}
#| label: Link_Flow_RMSE_bm

# route_id is only unique within an OD pair (routes are numbered 0..k-1 per
# OD), so origin/dest must be carried through the join alongside route_id -
# joining on route_id alone pools flow across unrelated OD pairs that happen
# to share the same route number. 
# seed/memory_level identify the run the data came from
bm_edges_routes <- data$od_routes_bm.parquet |>
  distinct(seed, memory_level, origin, dest, route_id, edge)

bm_flow_routes <- data$flows_bm.parquet |>
  group_by(seed, memory_level) |>
  filter(episode == max(episode)) |>
  ungroup() |>
  group_by(seed, memory_level, origin, destination, path) |>
  summarise(flow = sum(count), .groups = "drop") |>
  rename(dest = destination, route_id = path)

bm_edges_flows <- bm_edges_routes |>
  left_join(
    bm_flow_routes,
    by = c("seed", "memory_level", "origin", "dest", "route_id")
  ) |>
  mutate(flow = coalesce(flow, 0)) |>
  group_by(seed, memory_level, edge) |>
  summarise(flow = sum(flow), .groups = "drop")
```

```{r}
#| label: Link_Flow_RMSE_dua

# duaIterate doesn't depend on BM's memory_level (only on seed), so every
# memory_level within a seed logged an identical copy of the DUA routes/
# flows. distinct()/group_by() below drop the memory_level column before
# aggregating, collapsing those duplicates into one DUA reference per seed.
dua_edges_routes <- data$od_routes_dua.parquet |>
  distinct(seed, origin, dest, route_id, edge)

dua_flow_routes <- data$flows_dua.parquet |>
  distinct(seed, origin, destination, path, count) |>
  group_by(seed, origin, destination, path) |>
  summarise(flow = sum(count), .groups = "drop") |>
  rename(dest = destination, route_id = path)

dua_edges_flows <- dua_edges_routes |>
  left_join(dua_flow_routes, by = c("seed", "origin", "dest", "route_id")) |>
  mutate(flow = coalesce(flow, 0)) |>
  group_by(seed, edge) |>
  summarise(flow = sum(flow), .groups = "drop")
```

```{r}
#| label: Link_Flow_RMSE_computation

# dua_edges_flows has no memory_level column, so each seed's single DUA row
# per edge is broadcast against every memory_level's BM row for that same
# (seed, edge) — the correct shared reference for the whole sweep.
edge_flows_table <- full_join(
  bm_edges_flows,
  dua_edges_flows,
  by = c("seed", "edge"),
  suffix = c("_BM", "_DUE")
) |>
  mutate(
    flow_BM = coalesce(flow_BM, 0),
    flow_DUE = coalesce(flow_DUE, 0)
  )

link_flow_rmse_df <- edge_flows_table |>
  mutate(squared_diff = (flow_BM - flow_DUE) ** 2) |>
  group_by(seed, memory_level) |>
  summarise(
    M = n_distinct(edge),
    sum_squared_diff = sum(squared_diff),
    sum_flow_DUE = sum(flow_DUE),
    .groups = "drop"
  ) |>
  mutate(
    numerator = sqrt(sum_squared_diff / M),
    denominator = sum_flow_DUE / M,
    link_flow_rmse = numerator / denominator
  ) |> 
  mutate(
    memory_level_fct = factor(memory_level, levels = representative_memory_levels) 
  ) |> 
  mutate(memory_level_fct = fct_rev(memory_level_fct)) |> 
  select(c(seed, link_flow_rmse, memory_level_fct))
```

```{r}
#| label: Link_Flow_RMSE_summary

link_flow_rmse_summary <- link_flow_rmse_df |>
  group_by(memory_level_fct) |>
  summarise(median_rmse = median(link_flow_rmse), .groups = "drop")
```

```{r}
#| label: plot_rmse_vs_memory

ggplot(
  data = link_flow_rmse_summary,
  mapping = aes(x = memory_level_fct, y = median_rmse)
) +
  geom_point() +
  geom_line(linewidth = 1.4) +
  geom_line(
    data = link_flow_rmse_df,
    mapping = aes(x = memory_level_fct, y = link_flow_rmse, group = seed),
    linewidth = 0.45,
    alpha = 0.6,
    colour = "grey40"
  ) +
  labs(
    title = "Sensitivity of link-flow RMSE to memory level",
    subtitle = "Per-seed link-flow NRMSE against duaIterate (thin lines) \nand their median (thick line)",
    x = "Memory level",
    y = "Link-flow RMSE (normalized)"
  )
  

ggsave(here("r","RQ8","figures","plot_rmse_vs_memory.png"), width = 12, height = 6)
 








  
```