---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ5:** Analyzing how introducing heterogeneous memory decay
  parameter values across agents affects the learning dynamics of the model?

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

**The plots to answer this question will be:**

## Learning dynamics in a disrupted network scenario
1. R-gap evolution
2. Flow on disrupted edge
3. Path share evolution

For more details, check the research questions document.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(patchwork)
library(xml2)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
congestion_regime_tiers <- c("High")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```

```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ5","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
parquet_files <- c(
  flow_paths = "bm_rgap.parquet",
  dua_rgap = "dua_rgap.parquet",
  bm_edgedata = "bm_edgedata.parquet",
  od_routes = "od_routes.parquet",
  demand_odt = "demand_odt.parquet",
  bm_results = "bm_results.parquet"
)

data <- parquet_files |>
  purrr::set_names() |>
  purrr::map(\(f) arrow::read_parquet(file.path(data_dir, f)))
```

```{r}
#| label: get_warm_up_period(exploration)

# It's not perfectly precise — the true per-agent start is episode > 24 and all routes visited, so some agents start a few episodes later than 24. But it's the honest guaranteed floor ("no agent has started learning before this point"), it's already computed for every run via config.warm_up, and given the plot already treats each seed's trajectory independently without forcing alignment (per the existing code comment about seeds finishing at very different episode counts), a single shared floor line is consistent with that treatment rather than adding seed-specific precision the rest of the plot doesn't have either.

exploration_phase_floor <- as.numeric(unique(data$bm_rgap$warm_up))
```

```{r}
#| label: types_memory

representative_types_memory <- c("True", "False")
representative_types_memory_labels <- c("Heterogeneous", "Homogeneous")
```

```{r}
#| label: check_heterogeneous_memory

# Make sure agents have heteregeneous memory
heterogeneous_memory_agents <- data$bm_results |> 
  filter(heterogeneous_memory == "True") |> 
  group_by(agent_id) |> 
  slice_head(n=1) |> 
  select(agent_id, memory_level, heterogeneous_memory)

empirical_mean <- mean(heterogeneous_memory_agents$memory_level)
theorical_mean <- unique(as.numeric(data$bm_rgap$memory_mean))

diff <- abs(empirical_mean - theorical_mean) < 0.05

if (!diff){
  stop("There is some error related to the memory of the heterogeneous memory agents")
}
```


```{r}
#| label: breaks_plot1

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
breaks <- seq(0,max_episode, by = 75)

# Remove breaks that are within step of the learning start
breaks <- breaks[abs(breaks - exploration_phase_floor) >= (50)]
# Add the learning start break
breaks <- sort(c(breaks, exploration_phase_floor))
labels <- as.character(breaks)
labels[breaks==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```

```{r}
#| label: data_plot_1_a

congestion_to_analyze <- "155"
disruption_start <- as.numeric(unique(data$bm_rgap$degradation_start_episode))
disruption_end <- as.numeric(unique(data$bm_rgap$degradation_end_episode))

rgap_df <- data$bm_rgap |> 
  filter(
    n_agents == congestion_to_analyze,
  ) |> 
  mutate(
    congestion_regime = make_congestion_regime(n_agents)
  ) |> 
  select(-c(degradation_start_episode, degradation_end_episode, n_agents, network_degraded, run_id, warm_up)) |> 
  mutate(
    type_memory_fct = factor(
      heterogeneous_memory,
      levels = representative_types_memory,
      labels = representative_types_memory_labels
    )
  ) |> 
  mutate(type_memory_fct = fct_rev(type_memory_fct)) |> 
  select(-heterogeneous_memory)
 
# Median and quantiles across seeds (here I always run the same episodes, 
# so i can compute summaries on seeds, instead of having to plot individual trajectories)
rgap_df_summary <- rgap_df |> 
  group_by(episode, congestion_regime, type_memory_fct) |> 
  summarise(
    median_rgap = median(rgap),
    q25_rgap    = quantile(rgap, 0.25),
    q75_rgap    = quantile(rgap, 0.75)
  ) |> 
  ungroup()
```

```{r}
#| label: plot_1_a

ggplot(
) +
  annotate(
    "rect",
    xmin = disruption_start,
    xmax = disruption_end,
    ymin = -Inf,
    ymax = Inf,
    fill = "orange",
    alpha = 0.25,
  ) +
  annotate(
    "text",
    x = (disruption_start + disruption_end) / 2,
    y = Inf,
    label = "Disruption",
    fontface = "bold",
    colour = "firebrick",
    vjust = 2.5,
    size = 4
  ) +
  geom_vline(
  xintercept = disruption_start,
  linetype = 2,
  alpha = 0.7,
  color = "orange",
  linewidth = 0.5
  ) +
  geom_vline(
  xintercept = disruption_end,
  linetype = 2,
  alpha = 0.7,
  color = "orange",
  linewidth = 0.5
  ) +
  geom_line(
    data = rgap_df_summary,
    mapping = aes(x = episode, y = median_rgap),
  ) + 
  geom_ribbon(
    data = rgap_df_summary,
    mapping = aes(x = episode, ymin = q25_rgap, ymax = q75_rgap),
    alpha = 0.1
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 1, size = 9),
) +
  labs(
    x = "Episode",
    y = "R-gap (%)",
    title = "Adaptation to temporary network disruption",
    subtitle = "Evolution before, during, and after the disruption. High congestion (2000)"
  ) +
  scale_x_continuous(breaks = breaks, labels = labels) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = "red",
    alpha = 0.08
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.5
  ) +
  geom_hline(yintercept = 0, alpha = 0.6, linewidth = 0.7, color = "#7D7D7D") +
  geom_hline(
    yintercept = 5,
    colour = "grey40",
    linetype = "longdash",
    linewidth = 0.5
  ) +
  # Automatically chosen breaks and add one extra tick
  scale_y_continuous(
    breaks = function(x) sort(unique(c(scales::breaks_pretty()(x), 5)))
) +
  facet_grid(type_memory_fct ~ .)
ggsave(here("r","RQ5","figures","plot1_rgap_evolution.png"), width = 12, height = 6)
  
```

```{r}
#| label: data_plot_2

# Infer the degraded edges (with their NB/SB/EB/WB direction suffixes)
# straight from the network files instead of hardcoding them: an edge
# counts as degraded if its lane speed differs between the baseline
# network and the one config.network_degraded was swapped in with.
network_degraded_path <- unique(data$bm_rgap$network_degraded)
stopifnot(length(network_degraded_path) == 1)
network_normal_path <- file.path(dirname(network_degraded_path), "Normal_Sioux_Falls.net.xml")

get_lane_speeds <- function(net_xml_path) {
  # Read the XML file and find all <lane> elements
  lanes <- xml_find_all(read_xml(net_xml_path), "//lane")
  
  # Extract the edge ID by removing the lane suffix (_0, _1, etc.)
  edge_id <- sub("_[0-9]+$", "", xml_attr(lanes, "id"))
  
  # Extract lane speeds, convert them to numeric,
  # and use edge IDs as the names of the vector
  set_names(as.numeric(xml_attr(lanes, "speed")), edge_id)
}

normal_speeds <- get_lane_speeds(network_normal_path)
degraded_speeds <- get_lane_speeds(network_degraded_path)
degraded_edges <- names(degraded_speeds)[degraded_speeds != normal_speeds[names(degraded_speeds)]]

# Efficiently dealing with large dataframes

# 1. Partition parquet file
pq_path_edgedata <- here("r/RQ5/data/edgedata_partitioned")

# Make sure folder is empty before writing, so we have a clean slate,
# and dont gets mixed with previous stuff
if (dir.exists(pq_path_edgedata)) {
  unlink(pq_path_edgedata, recursive = TRUE, force = TRUE)
}

dir.create(pq_path_edgedata, recursive = TRUE)

data$bm_edgedata |> 
  group_by(seed) |> 
  write_dataset(path = pq_path_edgedata, format = "parquet")

# 2. List partitioned parquet files
tibble(
  files = list.files(pq_path_edgedata, recursive = TRUE),
  size_MB = file.size(file.path(pq_path_edgedata, files)) / 1024^2
)

# 3. Use dplyr with arrow
edgedata_pq <- open_dataset(pq_path_edgedata)
query <- edgedata_pq |> 
  # Filter only by degraded edges
  filter(
    edge %in% degraded_edges,
    n_agents == congestion_to_analyze,
  ) |> 
  mutate(
    entered = as.numeric(entered)
  ) |> 
  group_by(episode, edge, seed, n_agents, heterogeneous_memory) |> 
  summarise(
    flow_edge = sum(entered)
  ) |> 
  ungroup() 

flow_degraded_edges_df_tmp <- query |> collect()

flow_degraded_edges_df <- flow_degraded_edges_df_tmp |> 
  mutate(
    type_memory_fct = factor(
      heterogeneous_memory,
      levels = representative_types_memory,
      labels = representative_types_memory_labels
    )
  ) |> 
  mutate(type_memory_fct = fct_rev(type_memory_fct)) |> 
  select(-heterogeneous_memory)

flow_degraded_edges_df_summary <- flow_degraded_edges_df |> 
  group_by(episode, edge, n_agents, type_memory_fct) |> 
  summarise(
    median_flow = median(flow_edge),
    q25_flow = quantile(flow_edge, 0.25),
    q75_flow = quantile(flow_edge, 0.75)
  )
```

```{r}
#| label: breaks_plot2

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
breaks <- seq(0,max_episode, by = 75)

# Remove breaks that are within step of the learning start
breaks <- breaks[abs(breaks - exploration_phase_floor) >= (50)]
# Add the learning start break
breaks <- sort(c(breaks, exploration_phase_floor))
labels <- as.character(breaks)
labels[breaks==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```

```{r}
#| label: plot_2

ggplot() +
  geom_line(
    data = flow_degraded_edges_df_summary,
    mapping = aes(x = episode, y = median_flow, color = edge),
  ) + 
  geom_ribbon(
    data = flow_degraded_edges_df_summary,
    mapping = aes(x = episode, ymin = q25_flow, ymax = q75_flow),
    alpha = 0.1
  ) +
  facet_grid(edge ~ type_memory_fct) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = "red",
    alpha = 0.08
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.5
  ) +
  annotate(
    "rect",
    xmin = disruption_start,
    xmax = disruption_end,
    ymin = -Inf,
    ymax = Inf,
    fill = "orange",
    alpha = 0.25,
  ) +
  annotate(
    "text",
    x = (disruption_start + disruption_end) / 2,
    y = Inf,
    label = "Disruption",
    fontface = "bold",
    colour = "firebrick",
    vjust = 2.5,
    size = 4
  ) +
  geom_vline(
  xintercept = disruption_start,
  linetype = 2,
  alpha = 0.7,
  color = "orange",
  linewidth = 0.5
  ) +
  geom_vline(
  xintercept = disruption_end,
  linetype = 2,
  alpha = 0.7,
  color = "orange",
  linewidth = 0.5
  ) +
  scale_x_continuous(breaks = breaks, labels = labels) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1, size = 9),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  labs(
    x = "Episode",
    y = "Edge flow (vehicles)",
    title = "Flow evolution on disrupted edges",
    subtitle = "Flows on the disrupted edges before, during, and after the disruption. High congestion (2000)"
  )

ggsave(here("r","RQ5","figures","plot2_flow_disrupted_edges_evolution.png"), width = 12, height = 6)
```

## Conclusions
For the tested grid-network conditions, the mean memory level of the population appears to be substantially more influential than the heterogeneity of memory levels around that mean.



