---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ5:** Analyzing how introducing heterogeneous memory decay
  parameter values across agents affects the learning dynamics of the model?

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

**The plots to answer this question will be:**

## Learning dynamics in a disrupted network scenario
1. R-gap evolution
2. Flow on disrupted edge
3. Path share evolution

For more details, check the research questions document.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(patchwork)
library(xml2)
library(glue)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
congestion_regime_tiers <- c("High")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```

```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ5","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
parquet_files <- c(
  bm_rgap = "bm_rgap.parquet",
  dua_rgap = "dua_rgap.parquet",
  od_routes = "od_routes.parquet",
  demand_odt = "demand_odt.parquet",
  cost_paths = "cost_paths.parquet",
  bm_results = "bm_results.parquet",
  flow_paths = "flow_paths.parquet"
  # bm_edgedata = "bm_edgedata.parquet",
)

data <- parquet_files |>
  purrr::set_names() |>
  purrr::map(\(f) arrow::read_parquet(file.path(data_dir, f)))
```

```{r}
#| label: representative_seeds_spaghetti_plots

# Randomly select 3 seeds (seeded so the same 3 - and their colors - are
# picked every time this re-renders)
set.seed(123)
representative_seeds <- sample(unique(data$bm_rgap$seed), size = 3)

labels_seeds <- c("S1", "S2", "S3")
```

```{r}
#| label: get_warm_up_period(exploration)

# It's not perfectly precise — the true per-agent start is episode > 24 and all routes visited, so some agents start a few episodes later than 24. But it's the honest guaranteed floor ("no agent has started learning before this point"), it's already computed for every run via config.warm_up, and given the plot already treats each seed's trajectory independently without forcing alignment (per the existing code comment about seeds finishing at very different episode counts), a single shared floor line is consistent with that treatment rather than adding seed-specific precision the rest of the plot doesn't have either.

exploration_phase_floor <- as.numeric(unique(data$bm_rgap$warm_up))
```

```{r}
#| label: check_heterogeneous_memory

# Make sure agents have heteregeneous memory
heterogeneous_memory_agents <- data$bm_results |> 
  filter(heterogeneous_memory == "True") |> 
  group_by(agent_id) |> 
  slice_head(n=1) |> 
  select(agent_id, memory_level, heterogeneous_memory)

empirical_mean <- mean(heterogeneous_memory_agents$memory_level)
theorical_mean <- unique(as.numeric(data$bm_rgap$memory_mean))

diff <- abs(empirical_mean - theorical_mean) < 0.05

if (!diff){
  stop("There is some error related to the memory of the heterogeneous memory agents")
}
```

```{r}
#| label: data_plot1

memory_mean <- unique(data$dua_rgap.parquet$memory_mean)

representative_memory_levels <- c("True","False")
representative_memory_levels_labels <- c(glue("Heteregeneous (Mean = {memory_mean})"), glue("Homogeneous (Mean = {memory_mean})"))

# Not aggregating mean +/- SD across seeds here: seeds finish training at very
# different episode counts (e.g. High regime: some stop by episode ~22, one
# runs to 129), so for a large share of each panel's episode range the "mean"
# would silently collapse to a single seed's raw value (SD undefined). Each
# seed's own trajectory is plotted instead - no cross-seed averaging, so no
# misleading shrinking-N problem.

rgap_df <- data$bm_rgap |>
  mutate(
    congestion_regime = make_congestion_regime(n_agents),
    ) |>
  filter(
    seed %in% representative_seeds
  ) |> 
  mutate(
    seed_fct = factor(seed, levels = representative_seeds, labels = labels_seeds)
  ) |> 
  mutate(
    memory_level_fct = factor(heterogeneous_memory, levels = representative_memory_levels, labels = representative_memory_levels_labels) 
  ) |> 
  mutate(memory_level_fct = fct_rev(memory_level_fct))
```

```{r}
#| label: breaks_plot1

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
breaks <- seq(0,max_episode, by = 75)

# Remove breaks that are within step of the learning start
breaks <- breaks[abs(breaks - exploration_phase_floor) >= (50)]
# Add the learning start break
breaks <- sort(c(breaks, exploration_phase_floor))
labels <- as.character(breaks)
labels[breaks==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```

```{r}
#| label: plot1

fill <- "#E8E5DE"
alpha <- 0.8
color <- "#6F6C67"

upper_limit_y_axis <- quantile(rgap_df$rgap, probs = c(0.95))

ggplot(
  data = rgap_df,
  mapping = aes(x = episode, y = rgap, group = seed_fct)
) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = fill,
    alpha = alpha
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = color,
  linewidth = 0.5
  ) +
  geom_line(alpha = 0.6, linewidth = 0.4) +
  geom_hline(yintercept = 0, alpha = 0.6, linewidth = 0.7, color = "#7D7D7D") +
  geom_hline(
    yintercept = 5,
    colour = "grey40",
    linetype = "longdash",
    linewidth = 0.5
  ) +
  facet_grid(memory_level_fct ~ congestion_regime) +
  labs(
    title = "R-gap evolution across episodes",
    subtitle = "Representative learning trajectories for three memory levels and three random seeds (BM)",
    x = "Episode",
    y = "R-gap (%)"
  ) +
coord_cartesian(ylim=c(-5,upper_limit_y_axis)) +
   scale_x_continuous(breaks = breaks, labels = labels) +
theme(
  axis.title.x = element_text(margin = margin(t = 1)),
  axis.text.x = element_text(angle = 30, hjust = 1),
  strip.text.y = element_text(angle = 0, size = 10)
)

ggsave(here("r","RQ5","figures","plot1_rgap_evolution_mem_seed.png"), width = 12, height = 6)
```

## Behavioral convergence
```{r}
#| label: data_plot_3_route_flow

# Things to take into account:
# 1. OD matrix is different across congestion_regimes (keeping everything else constant)
# 2. OD matrix is different across seeds (keeping everything else constant)
# 3. OD matrix is the same across memory levels (keeping everything else constant)

# When analyzing the impact of memory level on the route flow evolution, we just want to analyze one OD from a FIXED OD MATRIX!!!
###########################
# Hence to choose our representative OD we have to pick:
# #########################
# 1. representative_seed
representative_seed <- representative_seeds[1]
# 2. representative_congestion
representative_congestion <- 2000
representative_congestion_label <- "High congestion (2000)"
# 3. random_memory_level
set.seed(123)
random_memory_level <- "True"

# To choose a representative OD, basically we are going to select an OD with a high number of available routes and a high demand. So we need two things:

# 1. OD matrix (contains the demand on each OD for this scenario)
OD_matrix <- data$demand_odt |> 
  filter(
    heterogeneous_memory == random_memory_level,
    seed == representative_seed, 
    n_agents == representative_congestion
  ) |> 
  select(-c(run_id, warm_up, heterogeneous_memory)) |> 
  group_by(origin, destination, seed, n_agents) |> 
  summarise(demand = sum(count)) |> 
  ungroup() |> 
  rename(
    dest = destination
  )

# 2. OD_routes (available routes for each OD)
OD_routes <- data$od_routes |> 
  filter(
    heterogeneous_memory == random_memory_level,
    seed == representative_seed, 
    n_agents == representative_congestion
  ) |> 
  select(-c(run_id, seed, heterogeneous_memory, n_agents, warm_up)) |> 
  group_by(origin, dest) |> 
  summarise(
    n_routes = n_distinct(route_id)
  ) |> 
  ungroup()

# JOIN: od_matrix + od_routes
OD_info <- OD_matrix |> 
  left_join(OD_routes, by = c("origin", "dest"))

representative_OD <- OD_info |> 
  arrange(desc(n_routes),desc(demand)) |> 
  slice(9) |> 
  select(-c(seed, n_agents))

second_representative_OD <- OD_info |> 
  arrange(desc(n_routes),desc(demand)) |> 
  slice(6) |> 
  select(-c(seed, n_agents))
```

```{r}
#| label: data_plot_3_route_flow_2

# Now that we have the scenario (defined by the seed and congestion_regime) 
# and we also have the representative OD whose paths we want to analyze, 
# we need to do the following:

# 1. Compute the flow on each route of the representative OD per episode
# Efficiently dealing with large dataframes

# 1. Partition parquet file
pq_path_flow_paths <- here("r/RQ5/data/flow_paths_partitioned")

if (dir.exists(pq_path_flow_paths)) {
  unlink(pq_path_flow_paths, recursive = TRUE, force = TRUE)
}

dir.create(pq_path_flow_paths, recursive = TRUE)

data$flow_paths |> 
  group_by(path) |> 
  write_dataset(path = pq_path_flow_paths, format = "parquet")

# 2. List partitioned parquet files
tibble(
  files = list.files(pq_path_flow_paths, recursive = TRUE),
  size_MB = file.size(file.path(pq_path_flow_paths, files)) / 1024^2
)

# 3. Use dplyr with arrow
flow_paths_pq <- open_dataset(pq_path_flow_paths)
query <- flow_paths_pq |> 
  filter(
    # Filter by the particular scenario
    seed == representative_seed, 
    n_agents == representative_congestion,
    # Filter by the particular OD
    origin == representative_OD$origin,
    destination == representative_OD$dest,
  ) |> 
  group_by(episode, path, heterogeneous_memory) |> 
  summarise(
    flow = sum(count)
  ) |> 
  ungroup()

query2 <- flow_paths_pq |> 
  filter(
    # Filter by the particular scenario
    seed == representative_seed, 
    n_agents == representative_congestion,
    # Filter by the particular OD
    origin == second_representative_OD$origin,
    destination == second_representative_OD$dest,
  ) |> 
  group_by(episode, path, heterogeneous_memory) |> 
  summarise(
    flow = sum(count)
  ) |> 
  ungroup()

flow_paths_df_query <- query |> collect()
flow_paths_df_query2 <- query2 |> collect()

# Check: The sum of flows across paths for the particular OD for one episode should be equal to the demand on that OD
check_demand_od_df <- flow_paths_df_query |> 
  group_by(episode, heterogeneous_memory) |> 
  summarise(
    demand = sum(flow)
  ) |> ungroup()

demand_od_sum_paths <- unique(check_demand_od_df$demand)
demand_od <- representative_OD$demand

if (demand_od != demand_od_sum_paths) {
  stop("Something is wrong in the computation of path flows for the particular OD and scenario.")
}

# Check: The sum of flows across paths for the particular OD for one episode should be equal to the demand on that OD
check_demand_od_df_2 <- flow_paths_df_query2 |> 
  group_by(episode, heterogeneous_memory) |> 
  summarise(
    demand = sum(flow)
  ) |> ungroup()

demand_od_sum_paths_2 <- unique(check_demand_od_df_2$demand)
demand_od_2 <- second_representative_OD$demand

if (demand_od_2 != demand_od_sum_paths_2) {
  stop("Something is wrong in the computation of path flows for the particular OD and scenario.")
}
```

```{r}
#| label: data_plot_3_part_3

# Now we still have to manipulate a bit the flow paths on the particular OD and scenario, so that is properly visualized in the plot

# 1. Handling some things to convert path variable to factor
path_levels <- as.character(sort(unique(flow_paths_df_query$path)))
path_levels_labels <- as.character(sort(unique(flow_paths_df_query$path + 1)))

flow_paths_df <- flow_paths_df_query |> 
  mutate(
    # Convert to factors
    congestion_regime = representative_congestion_label,
    memory_level_fct = forcats::fct_rev(factor(
      heterogeneous_memory, 
      levels = representative_memory_levels, 
      labels = representative_memory_levels_labels)),
    path = factor(
      path, 
      levels = path_levels,
      labels = path_levels_labels
    )
  )

# 1. Handling some things to convert path variable to factor
path_levels_2 <- as.character(sort(unique(flow_paths_df_query2$path)))
path_levels_labels_2 <- as.character(sort(unique(flow_paths_df_query2$path + 1)))

flow_paths_df_2 <- flow_paths_df_query2 |> 
  mutate(
    # Convert to factors
    congestion_regime = representative_congestion_label,
    memory_level_fct = forcats::fct_rev(factor(
      heterogeneous_memory, 
      levels = representative_memory_levels, 
      labels = representative_memory_levels_labels)),
    path = factor(
      path, 
      levels = path_levels_2,
      labels = path_levels_labels_2
    )
  )
```

```{r}
#| label: breaks_plot3
#| eval: False

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
step_plot3 <- floor(max_episode / 50) * 10
breaks_plot3 <- seq(0, max_episode, by = step_plot3)

# Remove breaks that are within step of the learning start
breaks_plot3 <- breaks[abs(breaks - exploration_phase_floor) >= step_plot3 / 2]
# Add the learning start break
breaks_plot3 <- sort(c(breaks_plot3, exploration_phase_floor))
labels_plot3 <- as.character(breaks_plot3)
labels_plot3[breaks_plot3==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```

```{r}
#| label: plot3

fill <- "#E8E5DE"
alpha <- 0.8
color <- "#6F6C67"

# This plots refers to a particular OD, for a particular seed and congestion regime, and facet for some representative memory_levels. 

ggplot(
  data = flow_paths_df, 
  mapping = aes(x = episode, y = flow, group = path, color = path)
) +
  facet_grid(memory_level_fct ~ .) +
  labs(
    color = "Route",
    y = "Route flow",
    x = "Episode",
  ) +
  # Shaded region
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = fill,
    alpha = alpha
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = color,
  linewidth = 0.4
  ) +
  annotate(
    "text",
    x = (exploration_phase_floor - 2) / 2,
    y = Inf,
    label = "warm-up \nphase",
    fontface = "bold",
    colour = color,
    vjust = 2.5,
    size = 4
  ) +
  geom_line(linewidth = 0.7) +
  # Change color scale
  scale_color_manual(
    values = c(
      "#000000",
      "#6B6D68",
      "#d2d4c8"
    )
  ) +
  # Tweak miscellaneous stuff
  theme(
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10),
    plot.subtitle = element_text(size = 11),
    strip.text.y = element_text(angle = 0, size = 10),
    panel.grid.major.y = element_line(
      color = "#E7E5DB",
      linewidth = 0.30
    ),
  ) +
  labs(
    title = "Behavioral convergence: Route flow stability",
    subtitle = glue(
      "{representative_congestion_label} | Seed {representative_seed} | OD pair: {representative_OD$origin} → {representative_OD$dest}"
    )
  )

ggsave(here("r","RQ5","figures","plot3_route_flow_stability.png"), width = 12, height = 6)
```

```{r}
#| label: plot3

fill <- "#E8E5DE"
alpha <- 0.8
color <- "#6F6C67"

# This plots refers to a particular OD, for a particular seed and congestion regime, and facet for some representative memory_levels. 

ggplot(
  data = flow_paths_df_2, 
  mapping = aes(x = episode, y = flow, group = path, color = path)
) +
  facet_grid(memory_level_fct ~ .) +
  labs(
    color = "Route",
    y = "Route flow",
    x = "Episode",
  ) +
  # Shaded region
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = fill,
    alpha = alpha
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = color,
  linewidth = 0.4
  ) +
  annotate(
    "text",
    x = (exploration_phase_floor - 2) / 2,
    y = Inf,
    label = "warm-up \nphase",
    fontface = "bold",
    colour = color,
    vjust = 2.5,
    size = 4
  ) +
  geom_line(linewidth = 0.7) +
  # Change color scale
  scale_color_manual(
    values = c(
      "#040700",
      "#4b534b",
      "#677369",
      "#9b9f8a",
      "#d6d6ba"
    )
  ) +
  # Tweak miscellaneous stuff
  theme(
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10),
    plot.subtitle = element_text(size = 11),
    strip.text.y = element_text(angle = 0, size = 10),
    panel.grid.major.y = element_line(
      color = "#E7E5DB",
      linewidth = 0.30
    ),
  ) +
  labs(
    title = "Behavioral convergence: Route flow stability",
    subtitle = glue(
      "{representative_congestion_label} | Seed {representative_seed} | OD pair: {representative_OD$origin} → {representative_OD$dest}"
    )
  )

ggsave(here("r","RQ5","figures","plot3_2_route_flow_stability.png"), width = 12, height = 6)
```



