---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ2:** Analyze whether the learning rate parameter affects
  the convergence speed of the BM model without significantly altering
  the final ability of the system to approach a DUE state.

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

**The plots to answer this question will be:**

1. Aggregated R-gap for different learning rates across episodes (representative learning dynamics)
2. Sensitivity analysis
3. Convergence episode vs Learning rate

For more details, check the research questions document.



```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
# congestion_regime_tiers <- c("Near free-flow", "Moderate", "High")
congestion_regime_tiers <- c("High")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```


```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ3","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
data <- list.files(data_dir, pattern="\\.parquet$", full.names = TRUE) |> 
  (\(x) set_names(x, tools::file_path_sans_ext(basename(x))))() |> 
  map(read_parquet)
```

```{r}
#| label: representative_seeds_spaghetti_plots

# Randomly select 3 seeds (seeded so the same 3 - and their colors - are
# picked every time this re-renders)
set.seed(123)
representative_seeds <- sample(unique(data$bm_rgap$seed), size = 3)

labels_seeds <- c("S1", "S2", "S3")

color_lines <- c(
  "0.9" = "#0072B2",  # blue
  "0.6" = "#E69F00",  # orange
  "0.2" = "#009E73"  # bluish green
) 
```


```{r}
#| label: data_plot1

representative_learning_rate_levels <- c("0.9","0.6","0.2")
representative_learning_rate_labels <- map_chr(representative_learning_rate_levels, \(x) paste0("l = ", x))

# Not aggregating mean +/- SD across seeds here: seeds finish training at very
# different episode counts (e.g. High regime: some stop by episode ~22, one
# runs to 129), so for a large share of each panel's episode range the "mean"
# would silently collapse to a single seed's raw value (SD undefined). Each
# seed's own trajectory is plotted instead - no cross-seed averaging, so no
# misleading shrinking-N problem.

rgap_df <- data$bm_rgap |>
  mutate(
    congestion_regime = make_congestion_regime(n_agents),
    ) |>
  filter(
    seed %in% representative_seeds
  ) |> 
  mutate(
    seed_fct = factor(seed, levels = representative_seeds, labels = labels_seeds)
  ) |> 
  filter(
    learning_rate %in% representative_learning_rate_levels
  ) |> 
  mutate(
    learning_rate_fct = factor(learning_rate, levels = representative_learning_rate_levels, labels = representative_learning_rate_labels) 
  ) |> 
  mutate(learning_rate_fct = fct_rev(learning_rate_fct))
```

```{r}
#| label: plot1

upper_limit_y_axis <- quantile(rgap_df$rgap, probs = c(0.95))

ggplot(
  data = rgap_df,
  mapping = aes(x = episode, y = rgap, group = seed_fct)
) +
  geom_line(alpha = 0.6, linewidth = 0.4) +
  geom_hline(yintercept = 0, alpha = 0.6, linewidth = 0.7, color = "#7D7D7D") +
  facet_grid(learning_rate_fct ~ congestion_regime) +
  labs(
    title = "R-gap evolution across episodes",
    subtitle = "Representative learning trajectories for three learning rates and three random seeds (BM)",
    x = "Episode",
    y = "R-gap (%)"
  ) +
coord_cartesian(ylim=c(-5,upper_limit_y_axis))

ggsave(here("r","RQ3","figures","plot1_rgap_evolution_l_seed.png"), width = 12, height = 6)
```

```{r}
#| label: data_plot2

sensitivity_df <- data$bm_rgap |> 
  mutate(
    learning_rate = as.numeric(learning_rate),
    congestion_regime = make_congestion_regime(n_agents),
  ) |> 
  select(-c(n_agents)) |> 
  group_by(seed, congestion_regime, learning_rate) |> 
  slice_max(episode, n = 1) |> 
  ungroup() 

sensitivity_df_summary <- sensitivity_df |> 
  group_by(learning_rate, congestion_regime) |> 
  summarise(
    median_rgap = median(rgap)
  ) |> 
  ungroup()

```

```{r}
#| label: plot2

ggplot(
  data = sensitivity_df_summary,
  mapping = aes(
    x = learning_rate,
    y = median_rgap
  )
) +
  geom_point() +
  geom_line(linewidth=1.4) +
  geom_line(
    data = sensitivity_df,
    mapping = aes(
      x = learning_rate,
      y = rgap,
      group = seed
    ),
    linewidth = 0.45,
    alpha = 0.6,
    colour = "grey60"
  ) +
  facet_wrap(~congestion_regime) +
  labs(
    title = "Sensitivity of R-gap to learning rate",
    subtitle = "Final-episode R-gap for five random seeds (thin lines) \nand their median (thick line)",
    x = "Learning_rate (l)",
    y = "R-gap (%)"
  ) + 
  scale_x_continuous(
    breaks = unique(sensitivity_df$learning_rate)
  )

ggsave(here("r","RQ3","figures","plot2_sensitivity_analysis_learning.png"), width = 12, height = 6)
```

```{r}
#| label: data_plot3

rgap_threshold <- 5

convergence_speed_df <- data$bm_rgap |> 
  mutate(
    learning_rate = as.numeric(learning_rate),
    congestion_regime = make_congestion_regime(n_agents),
  ) |> 
  select(-c(n_agents)) |>
  mutate(
    below = rgap < rgap_threshold
  ) |> 
  group_by(seed, learning_rate, congestion_regime) |> 
  # TRUE if all remaining episodes are below threshold
  mutate(
    remainder = map_lgl(
      # Returns a vector with all the row numbers
      row_number(),
      \(x) all(below[x:n()])
      )
  ) |> 
  filter(remainder) |> 
  slice_head(n = 1) |> 
  rename(
    convergence_episode = episode
  ) |> 
  ungroup() |> 
  select(-c(below, remainder))

convergence_speed_df_summary <- convergence_speed_df |> 
  group_by(learning_rate, congestion_regime) |> 
  summarise(
    median_convergence_episode = median(convergence_episode)
  ) |> 
  ungroup()
  
  
```

```{r}
#| label: plot3 

ggplot(
  data = convergence_speed_df_summary,
  mapping = aes(
    x = learning_rate,
    y = median_convergence_episode
  )
) +
  geom_point() +
  geom_line(linewidth=1.4) +
  geom_line(
    data = convergence_speed_df,
    mapping = aes(
      x = learning_rate,
      y = convergence_episode,
      group = seed
    ),
    linewidth = 0.45,
    alpha = 0.6,
    colour = "grey60"
  ) +
  facet_wrap(~congestion_regime) +
  labs(
    title = "Sensitivity of R-gap to learning rate",
    subtitle = "Final-episode R-gap for five random seeds (thin lines) \nand their median (thick line)",
    x = "Learning rate (l)",
    y = "R-gap (%)"
  ) + 
  scale_x_continuous(
    breaks = unique(convergence_speed_df$learning_rate)
  )

ggsave(here("r","RQ3","figures","plot3_convergence_speed_l.png"), width = 12, height = 6)


```
