---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ2:** Analyze whether the learning rate parameter affects
  the convergence speed of the BM model without significantly altering
  the final ability of the system to approach a DUE state.

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

**The plots to answer this question will be:**

1. Aggregated R-gap for different learning rates across episodes (representative learning dynamics)
2. Sensitivity analysis
3. Convergence episode vs Learning rate
4. (Appendix) Policy stabilization

For more details, check the research questions document.



```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(scales)
library(glue)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
# congestion_regime_tiers <- c("Near free-flow", "Moderate", "High")
congestion_regime_tiers <- c("Near free-flow", "Moderate", "High")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```


```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ3","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
data <- list.files(data_dir, pattern="\\.parquet$", full.names = TRUE) |> 
  (\(x) set_names(x, tools::file_path_sans_ext(basename(x))))() |> 
  map(read_parquet)
```

```{r}
#| label: representative_seeds_spaghetti_plots

# Randomly select 3 seeds (seeded so the same 3 - and their colors - are
# picked every time this re-renders)
set.seed(123)
representative_seeds <- sample(unique(data$bm_rgap$seed), size = 3)

labels_seeds <- c("S1", "S2", "S3")

color_lines <- c(
  "0.9" = "#bcb8b1",  # blue
  "0.6" = "#4A4844",  # orange
  "0.3" = "#11110F"  # bluish green
) 
```

```{r}
#| label: get_warm_up_period(exploration)

# It's not perfectly precise — the true per-agent start is episode > 24 and all routes visited, so some agents start a few episodes later than 24. But it's the honest guaranteed floor ("no agent has started learning before this point"), it's already computed for every run via config.warm_up, and given the plot already treats each seed's trajectory independently without forcing alignment (per the existing code comment about seeds finishing at very different episode counts), a single shared floor line is consistent with that treatment rather than adding seed-specific precision the rest of the plot doesn't have either.

exploration_phase_floor <- as.numeric(unique(data$bm_rgap$warm_up))
```


```{r}
#| label: data_plot1

representative_learning_rate_levels <- c("0.9","0.6","0.3")
representative_learning_rate_labels <- map_chr(representative_learning_rate_levels, \(x) paste0("LR = ", x))

# Not aggregating mean +/- SD across seeds here: seeds finish training at very
# different episode counts (e.g. High regime: some stop by episode ~22, one
# runs to 129), so for a large share of each panel's episode range the "mean"
# would silently collapse to a single seed's raw value (SD undefined). Each
# seed's own trajectory is plotted instead - no cross-seed averaging, so no
# misleading shrinking-N problem.

rgap_df <- data$bm_rgap |>
  mutate(
    congestion_regime = make_congestion_regime(n_agents),
    ) |>
  filter(
    seed %in% representative_seeds
  ) |> 
  mutate(
    seed_fct = factor(seed, levels = representative_seeds, labels = labels_seeds)
  ) |> 
  filter(
    learning_rate %in% representative_learning_rate_levels
  ) |> 
  mutate(
    learning_rate_fct = factor(learning_rate, levels = representative_learning_rate_levels, labels = representative_learning_rate_labels) 
  ) |> 
  mutate(learning_rate_fct = fct_rev(learning_rate_fct))
```

```{r}
#| label: breaks_plot1
# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
step <- floor(max_episode / 30) * 10
breaks <- seq(0, max_episode, by = step)

# Remove breaks that are within step of the learning start
breaks <- breaks[abs(breaks - exploration_phase_floor) >= (step/2.25)]
# Add the learning start break
breaks <- sort(c(breaks, exploration_phase_floor))
labels <- as.character(breaks)
labels[breaks==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```


```{r}
#| label: plot1

upper_limit_y_axis <- quantile(rgap_df$rgap, probs = c(0.95))

fill <- "#E8E5DE"
alpha <- 0.8
color <- "#6F6C67"

ggplot(
  data = rgap_df,
  mapping = aes(x = episode, y = rgap, group = seed_fct)
) +
  # Shaded region
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = fill,
    alpha = alpha
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = color,
  linewidth = 0.5
  ) +
  facet_grid(learning_rate_fct ~ congestion_regime) +
  labs(
    title = "R-gap evolution across episodes",
    subtitle = "Representative learning trajectories for three learning rates and three random seeds (BM)",
    x = "Episode",
    y = "R-gap (%)"
  ) +
  # Horizontal lines
  geom_hline(yintercept = 0, alpha = 0.6, linewidth = 0.7, color = "#7D7D7D") +
  geom_hline(
    yintercept = 5,
    colour = "grey40",
    linetype = "longdash",
    linewidth = 0.5
  )  +
  # Breaks X-axis
  scale_x_continuous(breaks = breaks, labels = labels) +
coord_cartesian(ylim=c(-5,upper_limit_y_axis)) +
  geom_line(alpha = 0.6, linewidth = 0.4) +
  theme(
    panel.grid.major.y = element_line(
      color = "#E7E5DB",
      linewidth = 0.30
    ),
    panel.grid.minor.y = element_blank(),
    axis.text.x=element_text(size=12)
  )

ggsave(here("r","RQ3","figures","plot1_rgap_evolution_l_seed.png"), width = 12, height = 6)
```

## Interpretation plot 1:
- **Final equilibrium quality**:  Despite the difference learning rates, in all cases it converges to DUE state. Therefore the learning rate has little influence on the final ability of the BM model to approach a DUE state. It also 
- **Behavioral convergence**: Higher learning rates lead to earlier stabilization of agents route choice policies.
- **Stability**: Contrary to the initial expectation, increasing the learning rate did not induce sustained oscillations in the aggregated R-gap trajectories.
- **Robustness across seeds**: All seeds exhibit pretty similar behavior.

```{r}
#| label: data_plot2
#| eval: false

sensitivity_df <- data$bm_rgap |> 
  mutate(
    learning_rate = as.numeric(learning_rate),
    congestion_regime = make_congestion_regime(n_agents),
  ) |> 
  select(-c(n_agents)) |> 
  group_by(seed, congestion_regime, learning_rate) |> 
  slice_max(episode, n = 1) |> 
  ungroup() 

sensitivity_df_summary <- sensitivity_df |> 
  group_by(learning_rate, congestion_regime) |> 
  summarise(
    median_rgap = median(rgap)
  ) |> 
  ungroup()

```

```{r}
#| label: plot2
#| eval: false

ggplot(
  data = sensitivity_df_summary,
  mapping = aes(
    x = learning_rate,
    y = median_rgap
  )
) +
  geom_point() +
  geom_line(linewidth=1.4) +
  geom_line(
    data = sensitivity_df,
    mapping = aes(
      x = learning_rate,
      y = rgap,
      group = seed
    ),
    linewidth = 0.45,
    alpha = 0.6,
    colour = "grey60"
  ) +
  facet_wrap(~congestion_regime) +
  labs(
    title = "Sensitivity of R-gap to learning rate",
    subtitle = "Final-episode R-gap for five random seeds (thin lines) \nand their median (thick line)",
    x = "Learning rate",
    y = "R-gap (%)"
  ) + 
  scale_x_continuous(
    breaks = unique(sensitivity_df_summary$learning_rate),
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9)
  )

ggsave(here("r","RQ3","figures","plot2_sensitivity_analysis_learning.png"), width = 12, height = 6)
```

Across all congestion regimes, the final R-gap remains relatively stable over the tested range of learning rates. Overall, the results indicate that the learning rate has little influence on the final equilibrium quality.

```{r}
#| label: data_plot3
#| eval: false

rgap_threshold <- 5

convergence_speed_df <- data$bm_rgap |> 
  mutate(
    learning_rate = as.numeric(learning_rate),
    congestion_regime = make_congestion_regime(n_agents),
  ) |> 
  select(-c(n_agents)) |>
  mutate(
    below = rgap < rgap_threshold
  ) |> 
  group_by(seed, learning_rate, congestion_regime) |> 
  # TRUE if all remaining episodes are below threshold
  mutate(
    remainder = map_lgl(
      # Returns a vector with all the row numbers
      row_number(),
      \(x) all(below[x:n()])
      )
  ) |> 
  filter(remainder) |> 
  slice_head(n = 1) |> 
  rename(
    convergence_episode = episode
  ) |> 
  ungroup() |> 
  select(-c(below, remainder))

convergence_speed_df_summary <- convergence_speed_df |> 
  group_by(learning_rate, congestion_regime) |> 
  summarise(
    median_convergence_episode = median(convergence_episode)
  ) |> 
  ungroup()
```

```{r}
#| label: plot3 
#| eval: false

range_learning_rate <- range(as.numeric(unique(data$bm_rgap$learning_rate)))

x_location <- ((range_learning_rate[1] + range_learning_rate[2]) / 2)

ggplot(
  data = convergence_speed_df_summary,
  mapping = aes(
    x = learning_rate,
    y = median_convergence_episode
  )
) +
  geom_point() +
  geom_line(linewidth=1.4) +
  geom_line(
    data = convergence_speed_df,
    mapping = aes(
      x = learning_rate,
      y = convergence_episode,
      group = seed
    ),
    linewidth = 0.45,
    alpha = 0.6,
    colour = "grey60"
  ) +
  facet_wrap(~congestion_regime) +
  labs(
    title = "Sensitivity of convergence speed to learning rate",
    subtitle = glue("First episode after which the R-gap remains below {rgap_threshold}% for the remaining of training"),
    x = "Learning rate",
    y = "Episode of convergence"
  ) + 
  scale_x_continuous(
    breaks = unique(convergence_speed_df$learning_rate)
  ) +
  scale_y_continuous(
    limits = c(0,max_episode)
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9)
  ) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = -Inf,
    ymax = exploration_phase_floor,
    fill = "red",
    alpha = 0.08
  ) +
  geom_hline(
  yintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = "red",
  linewidth = 0.5
  ) +
  annotate(
    "text",
    x = x_location,
    y = exploration_phase_floor / 2,
    label = "pre-learning \nphase",
    fontface = "bold",
    colour = "firebrick",
    size = 3.5
  )

# ggsave(here("r","RQ3","figures","plot3_convergence_speed_l.png"), width = 12, height = 6)
```

## Interpretation plot 3

- **Convergence speed (DUE)**: 
  - Higher learning rates generally reduce the convergence episode.
  - The benefit of increasing the learning rate becomes more pronounced as the network becomes more congested. The improvement is much more substantial in the high congestion case (roughly 63 - 35/40) 
  - Most of the gain seems to occur between 0.2 and 0.6. Beyond that, increasing the learning rate further produces relatively small additional reductions in convergence speed.
  
- **Robustness across seeds**: All seeds show similar behavior.

## Behavioral convergence
```{r}

#| label: data_plot_4_route_flow
# Things to take into account:
# 1. OD matrix is different across congestion_regimes (keeping everything else constant)
# 2. OD matrix is different across seeds (keeping everything else constant)
# 3. OD matrix is the same across learning rates (keeping everything else constant)

# When analyzing the impact of learning rate on the route flow evolution, we just want to analyze one OD from a FIXED OD MATRIX!!!
###########################
# Hence to choose our representative OD we have to pick:
# #########################
# 1. representative_seed
representative_seed <- representative_seeds[2]
# 2. representative_congestion
representative_congestion <- 2000
representative_congestion_label <- "High congestion (2000)"
# 3. random_learning_rate
set.seed(123)
random_learning_rate <- sample(unique(data$demand_odt$learning_rate), size = 1)

# To choose a representative OD, basically we are going to select an OD with a high number of available routes and a high demand. So we need two things:

# 1. OD matrix (contains the demand on each OD for this scenario)
OD_matrix <- data$demand_odt |> 
  filter(
    learning_rate == random_learning_rate,
    seed == representative_seed, 
    n_agents == representative_congestion
  ) |> 
  select(-c(run_id, warm_up, learning_rate)) |> 
  group_by(origin, destination, seed, n_agents) |> 
  summarise(demand = sum(count)) |> 
  ungroup() |> 
  rename(
    dest = destination
  )

# 2. OD_routes (available routes for each OD)
OD_routes <- data$od_routes |> 
  filter(
    learning_rate == random_learning_rate,
    seed == representative_seed, 
    n_agents == representative_congestion
  ) |> 
  select(-c(run_id, seed, learning_rate, n_agents, warm_up)) |> 
  group_by(origin, dest) |> 
  summarise(
    n_routes = n_distinct(route_id)
  ) |> 
  ungroup()

# JOIN: od_matrix + od_routes
OD_info <- OD_matrix |> 
  left_join(OD_routes, by = c("origin", "dest"))

representative_OD <- OD_info |> 
  arrange(desc(n_routes),desc(demand)) |> 
  slice(1) |> 
  select(-c(seed, n_agents))
```

```{r}
#| label: data_plot_4_route_flow_2

# Now that we have the scenario (defined by the seed and congestion_regime) 
# and we also have the representative OD whose paths we want to analyze, 
# we need to do the following:

# 1. Compute the flow on each route of the representative OD per episode
# Efficiently dealing with large dataframes

# 1. Partition parquet file
pq_path_flow_paths <- here("r/RQ3/data/flow_paths_partitioned")

# Make sure folder is empty before writing, so we have a clean slate,
# and dont gets mixed with previous stuff
if (dir.exists(pq_path_flow_paths)) {
  unlink(pq_path_flow_paths, recursive = TRUE, force = TRUE)
}

dir.create(pq_path_flow_paths, recursive = TRUE)

data$flow_paths |> 
  group_by(path) |> 
  write_dataset(path = pq_path_flow_paths, format = "parquet")

# 2. List partitioned parquet files
tibble(
  files = list.files(pq_path_flow_paths, recursive = TRUE),
  size_MB = file.size(file.path(pq_path_flow_paths, files)) / 1024^2
)

# 3. Use dplyr with arrow
flow_paths_pq <- open_dataset(pq_path_flow_paths)
query <- flow_paths_pq |> 
  filter(
    # Filter by the particular scenario
    seed == representative_seed, 
    n_agents == representative_congestion,
    learning_rate %in% representative_learning_rate_levels,
    # Filter by the particular OD
    origin == representative_OD$origin,
    destination == representative_OD$dest,
  ) |> 
  group_by(episode, path, learning_rate) |> 
  summarise(
    flow = sum(count)
  ) |> 
  ungroup()

flow_paths_df_query <- query |> collect()

# Check: The sum of flows across paths for the particular OD for one episode should be equal to the demand on that OD
check_demand_od_df <- flow_paths_df_query |> 
  group_by(episode, learning_rate) |> 
  summarise(
    demand = sum(flow)
  ) |> ungroup()

demand_od_sum_paths <- unique(check_demand_od_df$demand)
demand_od <- representative_OD$demand

if (demand_od != demand_od_sum_paths) {
  stop("Something is wrong in the computation of path flows for the particular OD and scenario.")
}
```

```{r}
#| label: data_plot_4_part_3

# Now we still have to manipulate a bit the flow paths on the particular OD and scenario, so that is properly visualized in the plot

# 1. Handling some things to convert path variable to factor
path_levels <- as.character(sort(unique(flow_paths_df_query$path)))
path_levels_labels <- as.character(sort(unique(flow_paths_df_query$path + 1)))

flow_paths_df <- flow_paths_df_query |> 
  mutate(
    # Convert to factors
    congestion_regime = representative_congestion_label,
    learning_rate_fct = forcats::fct_rev(factor(
      learning_rate, 
      levels = representative_learning_rate_levels, 
      labels = representative_learning_rate_labels)),
    path = factor(
      path, 
      levels = path_levels,
      labels = path_levels_labels
    )
  )
```

```{r}
#| label: plot4

fill <- "#E8E5DE"
alpha <- 0.8
color <- "#6F6C67"

# Only routes 1, 3 and 6 are highlighted with their own colour; every other
# route is drawn in the same very light grey so the highlighted ones stand out.
highlight_routes <- c("1", "3", "6")
highlight_colors <- c("1" = "#000000", "3" = "#bcabae", "6" = "#5f7470")
route_levels <- levels(flow_paths_df$path)
route_colors <- setNames(rep("#D9D7CF", length(route_levels)), route_levels)
route_colors[names(highlight_colors)] <- highlight_colors

# This plots refers to a particular OD, for a particular seed and congestion regime, and facet for some representative learning rates
ggplot(
  data = flow_paths_df,
  mapping = aes(x = episode, y = flow, group = path, color = path)
) +
  # Shaded region
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = fill,
    alpha = alpha
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = color,
  linewidth = 0.4
  ) +
  annotate(
    "text",
    x = (exploration_phase_floor - 2) / 2,
    y = Inf,
    label = "warm-up \nphase",
    fontface = "bold",
    colour = color,
    vjust = 2.5,
    size = 3.5
  ) +
  # Draw the non-highlighted routes first, then the highlighted ones on top
  # (slightly thicker) so they are never hidden behind the grey lines.
  geom_line(
    data = ~ dplyr::filter(.x, !path %in% highlight_routes),
    linewidth = 0.5
  ) +
  geom_line(
    data = ~ dplyr::filter(.x, path %in% highlight_routes),
    linewidth = 1
  ) +
  facet_grid(learning_rate_fct ~ .) +
  labs(
    color = "Route",
    y = "Route flow",
    x = "Episode",
    title = "Behavioral convergence: Route flow stability",
    subtitle = glue(
      "{representative_congestion_label} | Seed {representative_seed} | OD pair: {representative_OD$origin} → {representative_OD$dest}"
    )
  ) +
  # Change color scale: highlighted routes keep their colour, the legend only
  # lists them, and every other route collapses to the same light grey.
  scale_color_manual(
    values = route_colors,
    breaks = highlight_routes
  ) +
  # Tweak miscellaneous stuff
  theme(
    plot.subtitle = element_text(size = 11),
    strip.text.y = element_text(angle = 0, size = 10),
    panel.grid.major.y = element_line(
      color = "#E7E5DB",
      linewidth = 0.30
    )
)

ggsave(here("r","RQ3","figures","plot4_route_flow_stability.png"), width = 12, height = 6)
```

## LATEX TABLE

```{r}
#|label: sibedar_latex_table_1

# 1. Choose random seed
set.seed(1)
random_seed <- sample(unique(data$bm_rgap$seed), size = 1)

# 2. Create agents_od_df structure
agents_od_df <- data$agents_od |> 
  mutate(
    learning_rate = as.numeric(learning_rate),
    n_agents = as.numeric(n_agents),
  ) |> 
  filter(
    seed == random_seed,
    learning_rate == 0.3,
    n_agents == 2000
  ) |> 
  select(-c(post_warm_up, run_id, seed, warm_up, learning_rate, n_agents, departure_time)) |> 
  rename(agent_id = id)

# 3. Create agents_representative_od_df structure
agents_representative_od_df <- agents_od_df |> 
  filter(
    origin == representative_OD$origin,
    destination == representative_OD$dest
  )
  
# 4. Create agents_last_episode_info structure
agents_last_episode_info <- data$bm_results |> 
  mutate(
    learning_rate = as.numeric(learning_rate),
    n_agents = as.numeric(n_agents)
  ) |> 
  filter(
    learning_rate == 0.3,
    seed == random_seed,
    n_agents == 2000,
    agent_id %in% agents_representative_od_df$agent_id
  ) |> 
  select(-c(run_id, seed, warm_up, learning_rate, n_agents)) |> 
  group_by(agent_id) |> 
  filter(
    episode == max(episode)
  ) |> 
  select(
    episode, agent_id, route_id, PT
  )

# 5. Handling some things to convert path variable to factor
path_levels <- as.character(sort(unique(flow_paths_df_query$path)))
path_levels_labels <- as.character(sort(unique(flow_paths_df_query$path + 1)))

# 6. Create typical perceptions structure
typical_perceptions <- agents_last_episode_info |> 
  group_by(route_id) |> 
  summarise(
    median_PT = median(PT),
  ) |> 
  mutate(
  path_id = factor(route_id, levels = path_levels, labels = path_levels_labels)
  ) |> 
  select(-route_id) |> 
  ungroup()

# 7.
share_paths_df_last_episode <- flow_paths_df |> 
  group_by(learning_rate) |> 
  filter(
    episode == max(episode)
  ) |> 
  rename(
    path_id = path
  ) |> 
  mutate(
    share = flow / representative_OD$demand
  ) |> 
  select(path_id, share, learning_rate) |> 
  arrange(learning_rate, path_id) |> 
  ungroup() |> 
  mutate(share = round(share, 2))

share_paths_wide <- share_paths_df_last_episode |> 
  pivot_wider(
    names_from = learning_rate,
    values_from = share,
    values_fill = 0
  )

latex_quantitative_table <- typical_perceptions |> 
  left_join(
  share_paths_wide,
  by = "path_id"
) 

write.table(latex_quantitative_table, file=here("r", "RQ3", "figures", "quantitative_table.txt"))
```


