---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "09/aug/2026"
format:
  html:
    self-contained: true
execute:
  warning: false
  message: false
editor_options:
  chunk_output_type: console
---

**RQ7:** Where on the network do Bush-Mosteller's converged route choices
differ from the duaIterate DUE benchmark, in terms of edge usage
(`entered`) and congestion (`density`)?

This is a qualitative spatial case study, not a statistical comparison
across seeds — `experiments/rq7/design.yaml` runs a single production
combination, and the visualization itself is an interactive `sumo-gui`
edge-coloring replay rather than a static plot. Each chunk below launches
one replay: it blocks until the `sumo-gui` window is closed, then the next
chunk runs.

BM and duaIterate are always colored on the *same* scale (computed from
the max across both, see `sumo_edge_analysis.py`), so switching between
the BM replay and the duaIterate replay below is a like-for-like spatial
comparison. `aggregated` colors each edge once for the whole episode;
`interval` recolors every 15 min (900 s).

Things to consider when generating SUMO edge visualization:
1. If aggregated set breakpoint, because is not done in code.
2. Edit Visualization -> Show edge color value (only for selected). Background 
black with alpha = 155, and fontcolor = white. Fontsize 70

Because we are generating 1hour and 5 min. If we divide that in 4 intervals, we got
each interval contains 975s.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(here)
```

```{r}
#| label: config

# run_rq7_edge_viz.py always reads from r/RQ7/data/, populated by
# `python scripts/run_analysis.py RQ7 --prepare-only` (or the full run) —
# see scripts/run_analysis.py::_prepare_rq7_data. Using the downloaded
# copies (not the live local data/ files) keeps this reproducible even
# after later, unrelated simulation runs have overwritten the local files.
python_bin <- "/home/miguel/.venvs/thesis_master/bin/python3"
script <- here("src", "analysis", "run_rq7_edge_viz.py")
```

```{r}
#| label: aggregated_BM

system2(python_bin, c(
  script,
  "--mode", "aggregated",
  "--metric", "entered",
  "--algorithm", "BM"
))
```

```{r}
#| label: aggregated_duaIterate

system2(python_bin, c(
  script,
  "--mode", "aggregated",
  "--metric", "entered",
  "--algorithm", "duaIterate"
))
```

```{r}
#| label: interval_BM

system2(python_bin, c(
  script,
  "--mode", "interval",
  "--period", "975",
  "--metric", "entered",
  "--algorithm", "BM"
))
```

```{r}
#| label: interval_duaIterate

system2(python_bin, c(
  script,
  "--mode", "interval",
  "--period", "975",
  "--metric", "entered",
  "--algorithm", "duaIterate"
))
```

<!--
Swap "--metric density" for "--metric entered" in any of the chunks above
to see edge usage (vehicle counts) instead of congestion.
-->