---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ7:** Qualitative comparison of paths used by both algorithms for a representative OD. In particular this script carries out the OD/path-level comparison.
  
Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

For more details, check the research questions document.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(patchwork)
library(xml2)
library(glue)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
congestion_regime_tiers <- c("High")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```

```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ7","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
parquet_files <- c(
  demand_odt = "demand_odt.parquet",
  od_routes_bm = "od_routes_bm.parquet",
  od_routes_dua = "od_routes_dua.parquet",
  flows_bm = "flows_bm.parquet",
  flows_dua = "flows_dua.parquet"
)

data <- parquet_files |>
  purrr::set_names() |>
  purrr::map(\(f) arrow::read_parquet(file.path(data_dir, f)))
```

```{r}
#| label: representative_OD

# 1. Choose a representative OD pair:
# (i) it has relatively high travel demand
# (ii) it possesses several feasible alternative routes

# Things to take into account:
# 1. OD matrix is different across congestion_regimes (keeping everything else constant)
# 2. OD matrix is different across seeds (keeping everything else constant)

# That is why the simulation for the RQ7 has only been done for a particular seed and congestion regime.

# 1. OD matrix (contains demand on each OD for this particular scenario)
OD_matrix <- data$demand_odt.parquet |> 
  group_by(origin, destination) |> 
  summarise(
    demand = sum(count)
  ) |> 
  ungroup() |> 
  rename(
    dest = destination
  )

# Check to make sure OD matrix is correct
if (!OD_matrix$demand |> sum() == "1847"){
  stop("There is some error related to the OD matrix.")
}

# 2. OD_routes (available routes for each OD)
OD_routes <- data$od_routes_bm.parquet |> 
  group_by(origin, dest) |> 
  summarise(
    n_routes = n_distinct(route_id)
  ) |> 
  ungroup()

representative_OD_df <- OD_matrix |> 
  left_join(OD_routes, by = c("origin", "dest")) |> 
  filter(n_routes > 1) |> 
  arrange(desc(demand), desc(n_routes))

representative_OD <- representative_OD_df |> 
  slice_head(n = 1)
```



```{r}
#| label: route_usage_BM

path_levels_bm <- seq(0,representative_OD$n_routes-1)
path_levels_bm_labels <- seq(1,representative_OD$n_routes)
  
flow_paths_bm_df <- data$flows_bm.parquet |> 
  filter(
    origin == representative_OD$origin,
    destination == representative_OD$dest,
    episode == max(episode)
  ) |> 
  mutate(
    path = factor(
      path, 
      levels = path_levels_bm,
      labels = path_levels_bm_labels
    ) 
  ) |> 
  group_by(path) |> 
  summarise(
    flow = sum(count)
  )
```

```{r}
#| label: route_usage_duaiterate

paths_dua <- data$flows_dua.parquet |> 
  filter(
    origin == representative_OD$origin,
    destination == representative_OD$dest
  ) |> 
  summarise(
    n_paths = n_distinct(path)
  ) |> 
  pull(n_paths)

path_levels_dua <- seq(0,paths_dua-1)
path_levels_dua_labels <- seq(1,paths_dua)

# Check duaiterate data structure is right
recorded_episodes_dua <- data$flows_dua.parquet$episode |> n_distinct()
if (recorded_episodes_dua > 1){
  stop("There is some error related to 'data$flows_dua.parquet' file.")
}

flow_paths_dua_df <- data$flows_dua.parquet |> 
  filter(
    origin == representative_OD$origin,
    destination == representative_OD$dest
  ) |> 
  mutate(
    path = factor(
      path, 
      levels = path_levels_dua,
      labels = path_levels_dua_labels
    ) 
  ) |> 
  group_by(path) |>
  summarise(
    flow = sum(count)
  )
```

```{r}
#| label: network_edge_shapes

# Non-internal <edge> elements each have exactly one <lane> child with a
# "shape" attribute (a straight polyline "x1,y1 x2,y2 ..."), which is the
# edge's physical geometry — reused both to draw route trajectories and as
# the light-gray full-network backdrop that gives them spatial context.

# Read XML of the network used for this run
network_stem <- read_lines(here("r", "RQ7", "data", "network_stem.txt"))
network_filename <- paste0(network_stem, ".xml")
all_networks <- list.files(
  here("sumo", "net"),
  pattern = "\\.net\\.xml$",
  recursive = TRUE,
  full.names = TRUE
)
network_matches <- all_networks[basename(all_networks) == network_filename]
if (length(network_matches) != 1) {
  stop(glue(
    "Expected exactly one network file named '{network_filename}' under sumo/net/, ",
    "found {length(network_matches)}."
  ))
}
network_path <- network_matches

# Function to parse the network XML edge shapes attributes.
# In particular:
# 1. SUMO shape attribute string has two points "x,y" separated by an space. 
# Example: 
# "107.20,-301.60 192.80,-301.60"
#    x1     y1      x2      y2
# 2. We wanna convert it to a df:
# | x      | y       |
# |--------|---------|
# | 107.20 | -301.60 |
# | 192.80 | -301.60 |

########
# Steps:
########
# 1. str_split(shape_str, " ")[[1]]: Splits on spaces. The [[1]] is because this function can accept more than one input string, and it returns a list with as much elements as input strings. So, with [[1]] we are saying that we want the results of separating by spaces the first (and only) string we passed as input.
# Result: c("107.20,-301.60", "192.80,-301.60")

# 2. str_split(","): Separate by comma
# Result: list(c("107.20", "-301.60"), c("192.80", "-301.60"))

# 3. map_chr(points,1): Walks the list and pulls out element 1 (x-string) from every point.
# Result: c("107.20", "192.80")
parse_shape <- function(shape_str) {
  points <- str_split(shape_str, " ")[[1]] |> str_split(",")
  tibble(
    x = as.numeric(map_chr(points, 1)),
    y = as.numeric(map_chr(points, 2))
  )
}

# Find all relevant edges elements of the network
network_edges <- xml_find_all(read_xml(network_path), "//edge[not(@function='internal')]")

# We create a tibble with two columns:
# 1. edge: (vectorized) For every edge it pulls out the id attribute giving one character vector — e.g. c("E10_11_EB", "E10_11_WB", ...)
# 2. shape_str: 
#   xml_find_first(): Find only first match. Basically for every edge, it pulls out the first lane child element. Because input were n edges, output is n lane elements as well.
# Then it pulls out the shape of the lane.
# So we end up with a tibble that contains the edges and its shapes,
# and its well aligned (since both columns are build from the same set of edges "network_edges" on the same order, is guarantee they are correctly aligned)
# Example:
# ┌───────────┬─────────────────────────────────┐
# │   edge    │            shape_str            │
# ├───────────┼─────────────────────────────────┤
# │ E10_11_EB │ "107.20,-301.60 192.80,-301.60" │
# └───────────┴─────────────────────────────────┘

# |> mutate(): Parses the shape of the edge as we saw before. Each cell is an entire small tibble containing the two points (x1,y1) and (x2,y2) for that edge
# Example:
# ┌───────────┬───────────────────────────────┬────────────────────────────────────────────────┐
# │   edge    │           shape_str           │                     shape                      │
# ├───────────┼───────────────────────────────┼────────────────────────────────────────────────┤
# │ E10_11_EB │ "107.20,-301.60               │ tibble(x=c(107.20,192.80),                     │
# │           │ 192.80,-301.60"               │ y=c(-301.60,-301.60))                          │
# └───────────┴───────────────────────────────┴────────────────────────────────────────────────┘


edge_shapes <- tibble(
  edge = xml_attr(network_edges, "id"),
  shape_str = xml_attr(xml_find_first(network_edges, "./lane[1]"), "shape")
) |>
  mutate(shape = map(shape_str, parse_shape)) |>
  select(-shape_str)

# We convert to long format
# Example one edge-row with a 2-point shape becomes 2 rows
# ┌───────────┬────────┬─────────┐
# │   edge    │   x    │    y    │
# ├───────────┼────────┼─────────┤
# │ E10_11_EB │ 107.20 │ -301.60 │
# ├───────────┼────────┼─────────┤
# │ E10_11_EB │ 192.80 │ -301.60 │
# └───────────┴────────┴─────────┘
network_backdrop <- edge_shapes |>
  unnest(shape)
```

```{r}
#| label: route_geometry

# Builds one ordered polyline per route by walking od_routes' edges in
# step order and concatenating each edge's shape points; inner_join with
# the flow table both attaches the linewidth aesthetic and drops candidate
# routes nobody actually used (routes present in od_routes but not in the
# converged flow table).
build_route_geometry <- function(od_routes_df, flow_df) {
  usage_df <- flow_df |>
    mutate(route_id = as.integer(path) - 1) |>
    select(route_id, flow)

  od_routes_df |>
    filter(
      origin == representative_OD$origin,
      dest == representative_OD$dest
    ) |>
    inner_join(usage_df, by = "route_id") |>
    arrange(route_id, step) |>
    left_join(edge_shapes, by = "edge") |>
    unnest(shape) |>
    select(route_id, flow, x, y)
}

route_geometry <- bind_rows(
  BM = build_route_geometry(data$od_routes_bm.parquet, flow_paths_bm_df),
  duaIterate = build_route_geometry(data$od_routes_dua.parquet, flow_paths_dua_df),
  .id = "algorithm"
) |>
  mutate(
    algorithm = factor(algorithm, levels = c("BM", "duaIterate")),
    route_id = factor(route_id)
  )
```

```{r}
#| label: plot_route_composition

ggplot() +
  geom_path(
    data = network_backdrop,
    mapping = aes(x = x, y = y, group = edge),
    color = "grey85",
    linewidth = 0.3
  ) +
  geom_path(
    data = route_geometry,
    mapping = aes(
      x = x, y = y,
      group = interaction(algorithm, route_id),
      color = route_id,
      linewidth = flow
    ),
    lineend = "round"
  ) +
  facet_wrap(~algorithm) +
  coord_equal() +
  scale_color_brewer(palette = "Dark2") +
  scale_linewidth(range = c(0.4, 3)) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  ) +
  labs(
    color = "Route",
    linewidth = "Flow (veh)",
    title = "Path-level route composition for a representative OD",
    subtitle = glue(
      "OD pair: {representative_OD$origin} → {representative_OD$dest} | Demand: {representative_OD$demand}"
    )
  )

ggsave(here("r","RQ7","figures","plot_route_composition.png"), width = 12, height = 6)
```
