---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ7:** Qualitative comparison of paths used by both algorithms for a representative OD. In particular this script carries out the OD/path-level comparison.
  
Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

For more details, check the research questions document.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(patchwork)
library(xml2)
library(glue)
library(gt)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
congestion_regime_tiers <- c("High")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```

```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ7","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
parquet_files <- c(
  demand_odt = "demand_odt.parquet",
  od_routes_bm = "od_routes_bm.parquet",
  od_routes_dua = "od_routes_dua.parquet",
  flows_bm = "flows_bm.parquet",
  flows_dua = "flows_dua.parquet"
)

data <- parquet_files |>
  purrr::set_names() |>
  purrr::map(\(f) arrow::read_parquet(file.path(data_dir, f)))
```

```{r}
#| label: representative_OD

# 1. Choose a representative OD pair:
# (i) it has relatively high travel demand
# (ii) it possesses several feasible alternative routes

# Things to take into account:
# 1. OD matrix is different across congestion_regimes (keeping everything else constant)
# 2. OD matrix is different across seeds (keeping everything else constant)

# That is why the simulation for the RQ7 has only been done for a particular seed and congestion regime.

# 1. OD matrix (contains demand on each OD for this particular scenario)
OD_matrix <- data$demand_odt.parquet |> 
  group_by(origin, destination) |> 
  summarise(
    demand = sum(count)
  ) |> 
  ungroup() |> 
  rename(
    dest = destination
  )

# Check to make sure OD matrix is correct
if (!OD_matrix$demand |> sum() == "1847"){
  stop("There is some error related to the OD matrix.")
}

# 2. OD_routes (available routes for each OD)
OD_routes <- data$od_routes_bm.parquet |> 
  group_by(origin, dest) |> 
  summarise(
    n_routes = n_distinct(route_id)
  ) |> 
  ungroup()

representative_OD_df <- OD_matrix |> 
  left_join(OD_routes, by = c("origin", "dest")) |> 
  filter(n_routes > 1) |> 
  arrange(desc(demand), desc(n_routes))

representative_OD <- representative_OD_df |> 
  slice_head(n = 1)
```



```{r}
#| label: route_usage_BM

path_levels_bm <- seq(0,representative_OD$n_routes-1)
path_levels_bm_labels <- seq(1,representative_OD$n_routes)
  
flow_paths_bm_df <- data$flows_bm.parquet |> 
  filter(
    origin == representative_OD$origin,
    destination == representative_OD$dest,
    episode == max(episode)
  ) |> 
  mutate(
    path = factor(
      path, 
      levels = path_levels_bm,
      labels = path_levels_bm_labels
    ) 
  ) |> 
  group_by(path) |> 
  summarise(
    flow = sum(count)
  )
```

```{r}
#| label: route_usage_duaiterate

paths_dua <- data$flows_dua.parquet |> 
  filter(
    origin == representative_OD$origin,
    destination == representative_OD$dest
  ) |> 
  summarise(
    n_paths = n_distinct(path)
  ) |> 
  pull(n_paths)

path_levels_dua <- seq(0,paths_dua-1)
path_levels_dua_labels <- seq(1,paths_dua)

# Check duaiterate data structure is right
recorded_episodes_dua <- data$flows_dua.parquet$episode |> n_distinct()
if (recorded_episodes_dua > 1){
  stop("There is some error related to 'data$flows_dua.parquet' file.")
}

flow_paths_dua_df <- data$flows_dua.parquet |> 
  filter(
    origin == representative_OD$origin,
    destination == representative_OD$dest
  ) |> 
  mutate(
    path = factor(
      path, 
      levels = path_levels_dua,
      labels = path_levels_dua_labels
    ) 
  ) |> 
  group_by(path) |>
  summarise(
    flow = sum(count)
  )
```

```{r}
#| label: network_edge_shapes

# Non-internal <edge> elements each have exactly one <lane> child with a
# "shape" attribute (a straight polyline "x1,y1 x2,y2 ..."), which is the
# edge's physical geometry — reused both to draw route trajectories and as
# the light-gray full-network backdrop that gives them spatial context.

# Read XML of the network used for this run
network_stem <- read_lines(here("r", "RQ7", "data", "network_stem.txt"))
network_filename <- paste0(network_stem, ".xml")
all_networks <- list.files(
  here("sumo", "net"),
  pattern = "\\.net\\.xml$",
  recursive = TRUE,
  full.names = TRUE
)
network_matches <- all_networks[basename(all_networks) == network_filename]
if (length(network_matches) != 1) {
  stop(glue(
    "Expected exactly one network file named '{network_filename}' under sumo/net/, ",
    "found {length(network_matches)}."
  ))
}
network_path <- network_matches

# Function to parse the network XML edge shapes attributes.
# In particular:
# 1. SUMO shape attribute string has two points "x,y" separated by an space. 
# Example: 
# "107.20,-301.60 192.80,-301.60"
#    x1     y1      x2      y2
# 2. We wanna convert it to a df:
# | x      | y       |
# |--------|---------|
# | 107.20 | -301.60 |
# | 192.80 | -301.60 |

########
# Steps:
########
# 1. str_split(shape_str, " ")[[1]]: Splits on spaces. The [[1]] is because this function can accept more than one input string, and it returns a list with as much elements as input strings. So, with [[1]] we are saying that we want the results of separating by spaces the first (and only) string we passed as input.
# Result: c("107.20,-301.60", "192.80,-301.60")

# 2. str_split(","): Separate by comma
# Result: list(c("107.20", "-301.60"), c("192.80", "-301.60"))

# 3. map_chr(points,1): Walks the list and pulls out element 1 (x-string) from every point.
# Result: c("107.20", "192.80")
parse_shape <- function(shape_str) {
  points <- str_split(shape_str, " ")[[1]] |> str_split(",")
  tibble(
    x = as.numeric(map_chr(points, 1)),
    y = as.numeric(map_chr(points, 2))
  )
}

# Find all relevant edges elements of the network
network_edges <- xml_find_all(read_xml(network_path), "//edge[not(@function='internal')]")

# We create a tibble with two columns:
# 1. edge: (vectorized) For every edge it pulls out the id attribute giving one character vector — e.g. c("E10_11_EB", "E10_11_WB", ...)
# 2. shape_str: 
#   xml_find_first(): Find only first match. Basically for every edge, it pulls out the first lane child element. Because input were n edges, output is n lane elements as well.
# Then it pulls out the shape of the lane.
# So we end up with a tibble that contains the edges and its shapes,
# and its well aligned (since both columns are build from the same set of edges "network_edges" on the same order, is guarantee they are correctly aligned)
# Example:
# ┌───────────┬─────────────────────────────────┐
# │   edge    │            shape_str            │
# ├───────────┼─────────────────────────────────┤
# │ E10_11_EB │ "107.20,-301.60 192.80,-301.60" │
# └───────────┴─────────────────────────────────┘

# |> mutate(): Parses the shape of the edge as we saw before. Each cell is an entire small tibble containing the two points (x1,y1) and (x2,y2) for that edge
# Example:
# ┌───────────┬───────────────────────────────┬────────────────────────────────────────────────┐
# │   edge    │           shape_str           │                     shape                      │
# ├───────────┼───────────────────────────────┼────────────────────────────────────────────────┤
# │ E10_11_EB │ "107.20,-301.60               │ tibble(x=c(107.20,192.80),                     │
# │           │ 192.80,-301.60"               │ y=c(-301.60,-301.60))                          │
# └───────────┴───────────────────────────────┴────────────────────────────────────────────────┘


edge_shapes <- tibble(
  edge = xml_attr(network_edges, "id"),
  shape_str = xml_attr(xml_find_first(network_edges, "./lane[1]"), "shape")
) |>
  mutate(shape = map(shape_str, parse_shape)) |>
  select(-shape_str)

# We convert to long format
# Example one edge-row with a 2-point shape becomes 2 rows
# ┌───────────┬────────┬─────────┐
# │   edge    │   x    │    y    │
# ├───────────┼────────┼─────────┤
# │ E10_11_EB │ 107.20 │ -301.60 │
# ├───────────┼────────┼─────────┤
# │ E10_11_EB │ 192.80 │ -301.60 │
# └───────────┴────────┴─────────┘
network_backdrop <- edge_shapes |>
  unnest(shape)
```

```{r}
#| label: create_path_signature

# route_id is assigned independently by each algorithm's route-generation
# process so "route_id == 0" in onealgorithm has no relationship to 
# "route_id == 0" in the other.
# Build a common identifier path_id from each route's actual ordered edge sequence (its
# "signature") so the same physical path gets the same color in both
# facets, regardless of what each algorithm happened to number it.
bm_routes <- data$od_routes_bm.parquet |>
  filter(
    origin == representative_OD$origin,
    dest == representative_OD$dest
    ) |>
  arrange(route_id, step) |>
  group_by(route_id) |>
  summarise(
    path_signature = paste(edge, collapse = " -> ")
  ) |>
  ungroup()

dua_routes <- data$od_routes_dua.parquet |>
  filter(
    origin == representative_OD$origin,
    dest == representative_OD$dest
    ) |>
  arrange(route_id, step) |>
  group_by(route_id) |>
  summarise(
    path_signature = paste(edge, collapse = " -> ")
  ) |>
  ungroup()

# Common identifier
all_paths <- bind_rows(
  bm_routes,
  dua_routes
) |>
  distinct(path_signature) |>
  arrange(path_signature) |>
  mutate(
    path_id = paste0("P", row_number())
  ) |>
  ungroup()

# Attach the shared path_id back onto each algorithm's (route_id -> signature)
# lookup, so build_route_geometry() can map its own route_id to it.
bm_routes  <- bm_routes  |> left_join(all_paths, by = "path_signature")
dua_routes <- dua_routes |> left_join(all_paths, by = "path_signature")
```

```{r}
#| label: route_geometry

# Builds one ordered polyline per route by walking od_routes' edges in
# step order and concatenating each edge's shape points; inner_join with
# the flow table both attaches the linewidth aesthetic and drops candidate
# routes nobody actually used (routes present in od_routes but not in the
# converged flow table). path_id_map supplies the algorithm's (route_id ->
# path_id) lookup from create_path_signature, so the returned rows are
# keyed by the cross-algorithm-comparable path_id, not the algorithm-local
# route_id.
build_route_geometry <- function(od_routes_df, flow_df, path_id_map) {
  usage_df <- flow_df |>
    mutate(route_id = as.integer(path) - 1) |>
    select(route_id, flow)

  od_routes_df |>
    filter(
      origin == representative_OD$origin,
      dest == representative_OD$dest
    ) |>
    inner_join(usage_df, by = "route_id") |>
    left_join(path_id_map |> select(route_id, path_id), by = "route_id") |>
    arrange(route_id, step) |>
    left_join(edge_shapes, by = "edge") |>
    unnest(shape) |>
    select(path_id, flow, edge, x, y)
}

route_geometry <- bind_rows(
  BM = build_route_geometry(data$od_routes_bm.parquet, flow_paths_bm_df, bm_routes),
  duaIterate = build_route_geometry(data$od_routes_dua.parquet, flow_paths_dua_df, dua_routes),
  .id = "algorithm"
) |>
  mutate(
    algorithm = factor(algorithm, levels = c("BM", "duaIterate")),
    path_id = factor(path_id, levels = all_paths$path_id)
  )

relevant_routes_geometry <- route_geometry |>
  filter(
    flow > representative_OD$demand / 10
  )

```

```{r}
#| label: od_markers

# Origin/destination labels are placed at the midpoint of the origin/dest
# edge's shape (mean of its shape points — exact midpoint for the typical
# 2-point straight edge) rather than at its start/end vertex, so the label
# sits centered along the edge instead of right on top of a corner. Not
# faceted by algorithm (same OD for both), so ggplot repeats it on both panels.
roles_levels <- c("Origin", "Destination")
roles_labels <- c("Origin edge", "Destination edge")

od_markers <- bind_rows(
  Origin = edge_shapes |>
    filter(edge == representative_OD$origin) |>
    unnest(shape) |>
    summarise(x = mean(x), y = mean(y)),
  Destination = edge_shapes |>
    filter(edge == representative_OD$dest) |>
    unnest(shape) |>
    summarise(x = mean(x), y = mean(y)),
  .id = "role"
) |>
  mutate(role = factor(role, levels = roles_levels, labels = roles_labels))
```

```{r}
#| label: route_segments

# geom_path can't show direction of travel, and without it a route that
# doubles back (e.g. starts on a westbound edge before turning towards a
# destination to the east) reads as an arbitrary detour. Turning each
# route into (x, y) -> (xend, yend) segments lets us attach an arrowhead
# to specific edges instead of the whole line.
route_segments <- route_geometry |>
  group_by(algorithm, path_id, edge) |>
  mutate(xend = lead(x), yend = lead(y)) |>
  ungroup() |>
  filter(!is.na(xend))

# Only the origin/destination edges get an arrowhead — enough to show
# initial/final direction of travel (e.g. why a route detours west first)
# without cluttering every intermediate edge along the route. Each row's
# `edge` is the edge its *start* point belongs to, so filtering on it picks
# out exactly that edge's own segment
route_endpoint_segments <- route_segments |>
  filter(edge %in% c(representative_OD$origin)) |> 
  group_by(edge, algorithm) |> 
  slice_head(n = 1)
```

```{r}
#| label: path_share_table

flows_bm_table <- data$flows_bm.parquet |> 
  filter(
    origin == representative_OD$origin,
    destination == representative_OD$dest,
    episode == max(episode)
  ) |> 
  group_by(path) |>
  summarise(
    flow = sum(count)
  ) |> 
  rename(route_id = path) |> 
  left_join(bm_routes, by = c("route_id")) |> 
  select(path_id, flow)

flows_dua_table <- data$flows_dua.parquet |> 
  filter(
    origin == representative_OD$origin,
    destination == representative_OD$dest,
  ) |> 
  group_by(path) |>
  summarise(
    flow = sum(count)
  ) |> 
  rename(route_id = path) |> 
  left_join(dua_routes, by = c("route_id")) |> 
  select(path_id, flow)

comparison_table <- full_join(
  flows_bm_table,
  flows_dua_table,
  by = "path_id",
  suffix = c("_BM", "_DUE")
) %>%
  mutate(
    flow_BM = coalesce(flow_BM, 0),
    flow_DUE = coalesce(flow_DUE, 0),
    share_BM = flow_BM / representative_OD$demand,
    share_DUE = flow_DUE / representative_OD$demand
  )

comparison_final_table <- comparison_table |> 
  select(path_id, share_BM, share_DUE) |> 
  rename(
    Path = path_id,
    "BM share (%)" = share_BM,
    "dueIterate share (%)" = share_DUE
  )
```

```{r}
#| label: table

p_table_path_share <- comparison_final_table |>
  gt() |>
  tab_header(
    title = "Path shares by algorithm"
  ) |>
  fmt_percent(
    columns = c(`BM share (%)`, `dueIterate share (%)`),
    decimals = 1
  ) |>
  tab_options(
    table.width = pct(100),
    data_row.padding = px(4),
    heading.padding = px(4)
  )

gtsave(
  p_table_path_share,
  here("r","RQ7","figures","path_share_table.png")
)
```

```{r}
#| label: plot_route_composition

ggplot() +
  geom_path(
    data = network_backdrop,
    mapping = aes(x = x, y = y, group = edge),
    color = "grey85",
    linewidth = 0.3
  ) +
  geom_path(
    data = route_geometry,
    mapping = aes(
      x = x, y = y,
      group = interaction(algorithm, path_id),
      color = path_id,
      linewidth = flow
    ),
    lineend = "round"
  ) +
  geom_segment(
    data = route_endpoint_segments,
    mapping = aes(
      x = x, y = y, xend = xend, yend = yend,
    ),
    lineend = "round",
    arrow = arrow(length = unit(0.20, "cm"), type = "closed")
  ) +
  ggrepel::geom_label_repel(
    data = od_markers,
    mapping = aes(x = x, y = y, label = role),
    
    # label appearance
    size = 3,
    fontface = "bold",
    label.size = NA,
    fill = alpha("white", 0.8),
    
    # repel behavior
    box.padding = 1,
    nudge_x = 0.4,

      # connector
    min.segment.length = 0,
    segment.size = 0.3,
    segment.color = "grey40",
    
    # reproducibility
    seed = 1,
  )  +
  facet_wrap(~algorithm) +
  # Preserve network aspect ratio
  coord_equal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
  ) +
  scale_color_brewer(palette = "Dark2", name = "Path") +
  scale_linewidth(
    guide = 'none',
    range = c(0.5, 2.5)
  ) +
  labs(
    color = "Route",
    linewidth = "Flow (veh)",
    title = "Path-level route composition for a representative OD",
    subtitle = glue(
      "OD pair: {representative_OD$origin} → {representative_OD$dest} | Demand: {representative_OD$demand} veh"
    )
  )

ggsave(here("r","RQ7","figures","plot_route_composition.png"), width = 12, height = 6)
```


