---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ7:** Quantitative comparison of paths used by both algorithms.
  
Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

For more details, check the research questions document.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(patchwork)
library(xml2)
library(glue)
library(gt)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())

# Builds the congestion_regime factor from whatever n_agents values are
# present in the data, instead of hardcoding levels/labels per plot.
congestion_regime_tiers <- c("High")

make_congestion_regime <- function(n_agents) {
  levels_sorted <- sort(unique(n_agents))
  
  if (length(congestion_regime_tiers) != length(levels_sorted)) {
    stop("The number of congestion regime tiers must equal the number of unique agent counts.")
  }
  
  factor(
    n_agents, 
    levels = levels_sorted,
    labels = paste0(congestion_regime_tiers, " (", levels_sorted, ")"))
}
```

```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ7","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
parquet_files <- c(
  demand_odt = "demand_odt.parquet",
  od_routes_bm = "od_routes_bm.parquet",
  od_routes_dua = "od_routes_dua.parquet",
  flows_bm = "flows_bm.parquet",
  flows_dua = "flows_dua.parquet"
)

data <- parquet_files |>
  purrr::set_names() |>
  purrr::map(\(f) arrow::read_parquet(file.path(data_dir, f)))
```

```{r}
#| label: Link_Flow_RMSE_part1_bm

# route_id is only unique within an OD pair (routes are numbered 0..k-1 per
# OD), so origin/dest must be carried through the join alongside route_id -
# joining on route_id alone pools flow across unrelated OD pairs that happen
# to share the same route number.
bm_edges_routes <- data$od_routes_bm.parquet |>
  distinct(origin, dest, route_id, edge)

bm_flow_routes <- data$flows_bm.parquet |>
  filter(
    episode == max(episode)
  ) |>
  group_by(origin, destination, path) |>
  summarise(
    flow = sum(count),
    .groups = "drop"
  ) |>
  rename(
    dest = destination,
    route_id = path
  )
bm_edges_flows <- bm_edges_routes |>
  left_join(bm_flow_routes, by = c("origin", "dest", "route_id")) |>
  mutate(
    flow = coalesce(flow, 0)
  ) |>
  group_by(edge) |>
  summarise(
    flow = sum(flow)
  )
```

```{r}
#| label: Link_Flow_RMSE_part2_dua

dua_edges_routes <- data$od_routes_dua.parquet |>
  distinct(origin, dest, route_id, edge)

dua_flow_routes <- data$flows_dua.parquet |>
  group_by(origin, destination, path) |>
  summarise(
    flow = sum(count),
    .groups = "drop"
  ) |>
  rename(
    dest = destination,
    route_id = path
  )
dua_edges_flows <- dua_edges_routes |>
  left_join(dua_flow_routes, by = c("origin", "dest", "route_id")) |>
  mutate(
    flow = coalesce(flow, 0)
  ) |>
  group_by(edge) |>
  summarise(
    flow = sum(flow)
  )
```

```{r}
#| label: Link_flow_RMSE_joined_table

edge_flows_table <- full_join(
  bm_edges_flows,
  dua_edges_flows,
  by = "edge",
  suffix = c("_BM", "_DUE")
) |> 
  mutate(
    flow_BM = coalesce(flow_BM, 0),
    flow_DUE = coalesce(flow_DUE, 0)
  )

M <- edge_flows_table$edge |> n_distinct()
```

```{r}
#| label: Link_flow_RMSE_computation

sum_squared_diffs <- edge_flows_table |> 
  mutate(
    diff = flow_BM - flow_DUE
  ) |> 
  mutate(
    squared_diffs = diff ** 2
  ) |> 
  summarise(
    sum_squared_diffs = sum(squared_diffs)
  )

numerator <- sqrt(sum_squared_diffs / M)
denominator <-(edge_flows_table$flow_DUE |> sum()) / M
link_flow_rmse <- numerator / denominator
```

```{r}
#| label: Effective_n_routes
```




