---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

**RQ12:** Question: Analyze how incorporating
waiting time into the generalized perceived route cost affects route-choice
behavior, learning dynamics, and convergence properties of the BM
model.

Convergence is measured via three R-gap variants computed by the `DUE_convergence`
package: aggregated R-gap (overall convergence level), refined R-gap by departure
interval (temporal heterogeneity), and OD-specific R-gap (spatial heterogeneity).
BM is compared against `duaIterate` — SUMO's built-in DUE solver — as a sanity-check
baseline.

On this research question only the aggregated R-gap will be considered.

**The plots to answer this question will be:**

1. Risk scatter (mechanism plot)
2. R-gap evolution across $\phi$
3. Flow evolution over episodes + quantitative table

This research question plots have the following structure. The first plot is a mechanism figure, illustrating the implications of the proposed mathematical formulation for route choice. It focuses on a single representative agent and shows how the introduction of the new behavioral component changes the perceived cost of alternative routes. The representative agent is selected so that its route-perceptions are aligned with the population-level aggregated metrics. In this sense, the agent is intended to approximate a typical individual within the population (so that when looking at the third figure, it is more easy to interpret the results).
The second plot is a convergence figure.
The third component consists of a qualitative route-share figure accompanied by a quantitative table. The qualitative figure shows the evolution of route shares over episodes for a particular OD pair, where route shares are aggregated across all agents using that OD. It therefore represents population-level route-choice behavior. The accompanying table provides quantitative information on the route characteristics and the resulting redistribution of route shares under the different model formulations.

It is important to distinguish the interpretation of the first and third figures. The behavior of the representative individual agent in the mechanism figure does not have why to coincide exactly with the population-level behavior shown in the route-share figure. That's why we choose an individual whose perceptions are aligned with the population, to ease the interpretation and narrative of the plots.

For more details, check the research questions document.

```{r}
#| label: setup

# To make sure r virtual environment is loaded when executing this script from python
library(renv)
renv::load()
library(tidyverse)
library(arrow)
library(ggh4x)
library(xml2)
library(glue)
library(ggtext)
library(patchwork)
```

```{r}
#| label: load_custom_functions

library(here)

# Load custom theme
source(file = here("r","shared","theme.R"))

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())
```


```{r}
#| label: data

# This function helps us to create reliable file paths
data_dir <- here("r","RQ12","data")

# Dynamic data import
# 1) list.files(): returns a vector with the file paths
# 2) set_names(): transforms it into a named vector, with names equal to the filename without extension. This functions are vectorized, thats why it works
# 3) map(): maps the function to every parquet file. It returns a list, where each element is a df
data <- list.files(data_dir, pattern="\\.parquet$", full.names = TRUE) |> 
  (\(x) set_names(x, tools::file_path_sans_ext(basename(x))))() |> 
  map(read_parquet)
```

```{r}
#| label: random_seed
# Randomly select 1 seed 
set.seed(23)
random_seed <- sample(unique(data$bm_rgap$seed), size = 1)
```

```{r}
#| label: get_warm_up_period(exploration)

# It's not perfectly precise — the true per-agent start is episode > 24 and all routes visited, so some agents start a few episodes later than 24. But it's the honest guaranteed floor ("no agent has started learning before this point"), it's already computed for every run via config.warm_up, and given the plot already treats each seed's trajectory independently without forcing alignment (per the existing code comment about seeds finishing at very different episode counts), a single shared floor line is consistent with that treatment rather than adding seed-specific precision the rest of the plot doesn't have either.

exploration_phase_floor <- as.numeric(unique(data$bm_rgap$warm_up))
```

```{r}
#| label: representative_OD

# Choose an OD whose shortest route is not the shortest in waitingTime (hence, there exist alternative routes that are longer but with less waitingTime).
# Is gonna be chosen taking into account equally weighted means (without memory mechanism) and with phi=0 (under normal drivers scenario). 
# We do it this way because we merely wanna inspect the network/demand structure to choose our OD and isolate it from the effect of other variables

# We want that the shortest route is at least 5 seconds faster than the most stable route
# (this way we could see a difference in the r-gap when implemented risk sensitivity model)
min_diff <- 5

rewards_df <- data$rewards |> 
  mutate(
    waiting_time_sensitivity = as.numeric(waiting_time_sensitivity)
  ) |> 
  filter(
    seed == random_seed,
    waiting_time_sensitivity == 0
  ) |> 
  select(-c(run_id, seed, warm_up, waiting_time_sensitivity))

actions_df <- data$actions |> 
  mutate(
    waiting_time_sensitivity = as.numeric(waiting_time_sensitivity)
  ) |> 
  filter(
    waiting_time_sensitivity == 0,
    seed == random_seed
  ) |> 
  select(-c(run_id, seed, warm_up, waiting_time_sensitivity))

agents_od_df <- data$agents_od |> 
  mutate(
    waiting_time_sensitivity = as.numeric(waiting_time_sensitivity)
  ) |> 
  filter(
    seed == random_seed,
    waiting_time_sensitivity == 0
  ) |> 
  select(-c(post_warm_up, run_id, seed, warm_up, waiting_time_sensitivity, departure_time)) |> 
  rename(agent_id = id)

trips_info_df <- data$trips_info |> 
  mutate(
    waiting_time_sensitivity = as.numeric(waiting_time_sensitivity)
  ) |> 
  filter(
    seed == random_seed,
    waiting_time_sensitivity == 0
  ) |> 
  select(c(episode, vehicle_id, waiting_time)) |> 
  rename(
    agent_id = vehicle_id
  ) |> 
  mutate(
    waiting_time = as.numeric(waiting_time)
  )

representative_od_df <- rewards_df |> 
  left_join(actions_df, by = c("episode", "agent_id")) |> 
  left_join(trips_info_df, by = c("episode", "agent_id")) |> 
  left_join(agents_od_df, by = "agent_id")

representative_od_df_summary <- representative_od_df |> 
  group_by(origin, destination, action) |> 
  summarise(
    mean_tt = mean(reward),
    mean_wt = mean(waiting_time),
    count = n()
  ) |> 
  ungroup() |> 
  filter(
    count > 15
  )

representative_od_decision_df <- representative_od_df_summary |> 
  group_by(
    origin, destination
  ) |> 
  mutate(
    is_shortest = mean_tt == min(mean_tt),
    is_wt_shortest = mean_wt == min(mean_wt),
    candidate_od = is_shortest != is_wt_shortest
  ) |> 
  filter(
    candidate_od
  ) |> 
  ungroup()

representative_od <- representative_od_decision_df |> 
  group_by(
    origin, destination
  ) |> 
  summarise(
    diff_mean_tt = abs(mean_tt[1] - mean_tt[2]),
    diff_mean_wt = abs(mean_wt[1] - mean_wt[2])
  ) |> 
  filter(
    diff_mean_tt > min_diff & diff_mean_wt > min_diff
  ) |> 
  arrange(desc(diff_mean_tt)) |> 
  slice_head(n = 1) |> 
  select(origin, destination) |> 
  ungroup()

representative_od_routes <- representative_od_df_summary |> 
  filter(
    origin == representative_od$origin,
    destination == representative_od$destination
  ) |> 
  arrange(mean_tt)
```


```{r}
#| label: data_plot1_points
agents_representative_od_df <- agents_od_df |> 
  filter(
    origin == representative_od$origin,
    destination == representative_od$destination
  ) 

# The points in the plot 1, will represent the beliefs of an agent about 
# the mean_tt and waitingTime of its routes. 
# We just need one agent, because that way we can easily illustrate 
# the mechanisms and how an agent perception about routes would change by 
# considering different phi we just need to use one agent.
# It is an agent belonging to the representative OD that we are gonna talk
# about in the following plots
agents_last_episode_info <- data$bm_results |> 
  mutate(
    waiting_time_sensitivity = as.numeric(waiting_time_sensitivity)
  ) |> 
  filter(
    waiting_time_sensitivity == 0,
    seed == random_seed,
    agent_id %in% agents_representative_od_df$agent_id
  ) |> 
  select(-c(run_id, seed, warm_up, waiting_time_sensitivity)) |> 
  group_by(agent_id) |> 
  filter(
    episode == max(episode)
  ) |> 
  select(
    episode, agent_id, route_id, PT, WT
  )

# The representative agent has been selected by inspecting at agents_last_episode_info and looking for an agent whose perceptions coincide with the typical perceptions, so that it is aligned with what we see in plot 3 qualitative. Is not any trick, is just that on plot 1 we show individual level, and plot 3 it shows aggregate level, and individual level and aggregate doesnt have to coincide, but to make plots easier to understand, if they can coincide it makes easier to tell the story
representative_agent <- "agent_169"
agent_last_episode_info <- agents_last_episode_info |> 
  filter(
    agent_id == representative_agent
  )
```

```{r}
#| label: phi_values

phi_values <- as.numeric(data$demand_odt$waiting_time_sensitivity |> unique())
```

```{r}
#| label: data_plot1_lines

# Intuition about the plot:
# 1) I wanna see which route is the best for each phi value.
# 2) Draw the contour lines for each phi value. In particular, draw the
# contour corresponding to the minimum (best) route cost. This shows which
# other combinations of values result in the same cost as the best route.

# map_dfr: runs the function once for each value in phi_values, and stacks the three results into one dataframe

iso_cost_lines_tmp <- map_dfr(phi_values, function(phi){
  # Choose route with the minimum generalized cost
  best <- agent_last_episode_info |> 
    mutate(
      cost = PT + phi * WT
    ) |> 
    slice_min(cost, n = 1)
  
  # Line when we do not take into account the variability.
  # Basically the contour line is vertical because variability does not matter for the generalized cost
  if (phi == 0){
    tibble(phi, type = "vertical", xintercept = best$PT, route_id = best$route_id, cost = best$cost)
  } else{
    # (WT_r = -PT_r/phi + cost/phi, i.e. slope -1/phi, intercept cost/phi)
    tibble(phi, type = "line", slope = -1 / phi, intercept = best$cost / phi, route_id = best$route_id, cost = best$cost)
  }
})

iso_cost_lines <- iso_cost_lines_tmp |> mutate(
  phi = factor(phi)
)
```

```{r}
#| label: Tweaks_to_improve_visualization

# 1. Improve route_id labels
path_levels <- agent_last_episode_info$route_id |> unique()
path_levels_labels <- map_chr(path_levels, \(x) paste0("Path ", x+1))
agent_last_episode_info_plot <- agent_last_episode_info |> 
  mutate(
    path_id = factor(route_id, levels = path_levels, labels = path_levels_labels)
  ) |> 
  select(-route_id)

# 2. Choose only iso cost lines that have different minimum cost routes
iso_cost_lines_plot_tmp <- iso_cost_lines |> 
  mutate(
    phi_num = as.numeric(as.character(phi))
  ) |> 
  group_by(route_id) |> 
  mutate(
    plot = if (any(type == "vertical", na.rm = TRUE)) {
      type == "vertical"
    } else {
      phi_num == max(phi_num, na.rm = TRUE)
    }
  ) |> 
  ungroup() |> 
  filter(plot) |> 
  select(-c(plot, phi_num))

# 3. Highlight min_cost routes
min_cost_routes <- iso_cost_lines_plot_tmp$route_id 
min_cost_routes_df <- agent_last_episode_info |> 
  filter(
    route_id %in% min_cost_routes
  )

upper_y_axis <- quantile(agent_last_episode_info_plot$WT, 0.80)

# 4. Label the iso cost lines
iso_cost_lines_plot <- iso_cost_lines_plot_tmp |> 
  left_join(min_cost_routes_df, by = "route_id") |> 
  mutate(
    phi_num = as.numeric(phi)
  ) |> 
  mutate(
    x = ifelse(type == "vertical", xintercept, (cost - phi_num * WT)),
    y = ifelse(
      type == "vertical", 
      max(agent_last_episode_info_plot |> filter(agent_last_episode_info_plot$WT < upper_y_axis) |> pull(WT)), 
      0
    ),
    cost = round(cost, 0),
    label = glue("\u03c6 = {phi}, C = {cost} s")
  )
```

```{r}
#| label: plot_1
ggplot(
  data = agent_last_episode_info_plot, 
  mapping = aes(x = PT, y = WT)
) +
  geom_point() +
  geom_point(
    data = min_cost_routes_df,
    mapping = aes(x = PT, y = WT),
    size = 4, 
    shape = "circle open", 
  ) +
  geom_vline(
    data = filter(iso_cost_lines_plot, type == "vertical"),
    mapping = aes(
      xintercept = xintercept, 
      ),
      linetype = "dashed",
      linewidth = 0.6
  ) +
  geom_abline(
    data = filter(iso_cost_lines_plot, type == "line"),
    mapping = aes(
      slope = slope, 
      intercept = intercept, 
      ), 
    linetype = "dashed",
    linewidth = 0.6
  ) +
  ggrepel::geom_text_repel(
    mapping = aes(label = path_id),
    size = 3.5,
    box.padding = 0.5,
    point.padding = 0.3,
    force = 1,
    max.overlaps = Inf,
    seed = 1,
    nudge_x = -1,
    nudge_y = 1
  ) +
  ggrepel::geom_text_repel(
    data = iso_cost_lines_plot,
    mapping = aes(
      x = x,
      y = y,
      label = label
    ),
    size = 3.5,
    box.padding = 1,
    point.padding = 0.3,
    force = 1,
    max.overlaps = Inf,
    seed = 123,
    nudge_x = 2,
    fontface = "bold",
    segment.alpha = 0.5,
    segment.size = 0.3,
  ) +
  labs(
    title = "Waiting-time sensitivity shifts the minimum-cost route",
    subtitle = "One representative agent; highlighted points show \nthe minimum-cost route. Dashed lines indicate equal perceived cost.",
  x = expression("Perceived travel time " * PT[r] * " (s)"),
  y = expression("Perceived waiting time " * WT[r] * " (s)"),
  ) +
  coord_cartesian(ylim = c(0, upper_y_axis))

ggsave(here("r","RQ12","figures","plot1_waitingTime_sensitivity_mechanism.png"), width = 12, height = 6)
```

```{r}
#| label: representative_seeds
representative_seeds <- sample(unique(data$bm_rgap$seed), size = 3)
labels_seeds <- c("S1", "S2", "S3")
```

```{r}
#| label: data_plot2

representative_waiting_time_sensitivity_levels <- c(0, 5, 10)
representative_waiting_time_sensitivity_levels_labels <- map_chr(
  representative_waiting_time_sensitivity_levels,
  \(x) paste0("\u03c6 = ", x)
)

# Not aggregating mean +/- SD across seeds here: seeds finish training at very
# different episode counts (e.g. High regime: some stop by episode ~22, one
# runs to 129), so for a large share of each panel's episode range the "mean"
# would silently collapse to a single seed's raw value (SD undefined). Each
# seed's own trajectory is plotted instead - no cross-seed averaging, so no
# misleading shrinking-N problem.

rgap_df <- data$bm_rgap |> 
  mutate(
    waiting_time_sensitivity = as.numeric(waiting_time_sensitivity)
  ) |> 
  filter(
    seed %in% representative_seeds,
    waiting_time_sensitivity %in% representative_waiting_time_sensitivity_levels
  ) |> 
  mutate(
    seed_fct = factor(seed, levels = representative_seeds, labels = labels_seeds)
  ) |> 
  select(-c(run_id, warm_up, seed)) |> 
  mutate(
    waiting_time_sensitivity_fct = 
      factor(waiting_time_sensitivity,
             levels = representative_waiting_time_sensitivity_levels,
             labels = representative_waiting_time_sensitivity_levels_labels
      )
  ) 
```

```{r}
#| label: breaks_plot2

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
step <- floor(max_episode / 30) * 10
breaks <- seq(0, max_episode, by = step)

# Remove breaks that are within step of the learning start
breaks <- breaks[abs(breaks - exploration_phase_floor) >= (step/2.25)]
# Add the learning start break
breaks <- sort(c(breaks, exploration_phase_floor))
labels <- as.character(breaks)
labels[breaks==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```

```{r}
#| label: plot2

fill <- "#E8E5DE"
alpha <- 0.8
color <- "#6F6C67"

upper_limit_y_axis <- quantile(rgap_df$rgap, probs = c(0.99))

ggplot(
  data = rgap_df,
  mapping = aes(x = episode, y = rgap, group = seed_fct)
) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = fill,
    alpha = alpha
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = color,
  linewidth = 0.5
  ) +
  geom_line(alpha = 0.6, linewidth = 0.4) +
  geom_hline(yintercept = 0, alpha = 0.6, linewidth = 0.7, color = "#7D7D7D") +
  geom_hline(
    yintercept = 5,
    colour = "grey25",
    linetype = "longdash",
    linewidth = 0.6,
  ) +
  facet_grid(waiting_time_sensitivity_fct ~ .) +
  labs(
    title = "Waiting-time sensitivity affects R-gap convergence",
    subtitle = glue("Representative learning trajectories for various waiting-time sensitivities across different seeds"),
    x = "Episode",
    y = "R-gap (%)"
  ) +
  coord_cartesian(ylim=c(-5,upper_limit_y_axis)) +
  theme(
    axis.title.x = element_text(margin = margin(t = 1)),
    axis.text.y = element_text(size = 11)
  ) +
  scale_x_continuous(breaks = breaks, labels = labels) +
  scale_y_continuous(
  breaks = function(x) {
    c(scales::breaks_pretty()(x), 5)
  }
)

ggsave(here("r","RQ12","figures","plot2_convergence_plot.png"), width = 12, height = 6)
```

```{r}
#| label: data_plot3_qualitative

# Things to take into account:
# 1. OD matrix is different across seeds (keeping everything else constant)
# 2. OD matrix is the same across waiting-time sensitivities (keeping everything else constant)

# When analyzing the impact of waiting-time sensitivity on the route flow evolution, we just want to analyze one OD from a FIXED OD MATRIX!!!
###########################
# We already choose our representative OD before
# #########################
# 1. representative_seed
representative_seed <- random_seed
# 2. representative_waiting_time_sensitivity
representative_waiting_time_sensitivity_levels
# 3. representative_od
representative_OD <- data$demand_odt |> filter(
  seed == representative_seed,
  origin == representative_od$origin,
  destination == representative_od$destination
) |> 
  mutate(
    waiting_time_sensitivity = as.numeric(waiting_time_sensitivity)
  ) |> 
  filter(
    waiting_time_sensitivity == min(waiting_time_sensitivity)
  ) |> 
  select(-c(run_id, seed, warm_up, waiting_time_sensitivity)) |> 
  group_by(origin, destination) |> 
  summarise(
    demand = sum(count)
  ) |> 
  ungroup() |> 
  rename(
    dest = destination
  )
```

```{r}
#| label: data2_plot3_qualitative

# Now that we have the scenario (defined by the seed) 
# and we also have the representative OD whose paths we want to analyze, 
# we need to do the following:

# 1. Compute the flow on each route of the representative OD per episode
# Efficiently dealing with large dataframes

# 1. Partition parquet file
pq_path_flow_paths <- here("r/RQ12/data/flow_paths_partitioned")

if (dir.exists(pq_path_flow_paths)) {
  unlink(pq_path_flow_paths, recursive = TRUE, force = TRUE)
}

dir.create(pq_path_flow_paths, recursive = TRUE)

data$flow_paths |> 
  group_by(path) |> 
  write_dataset(path = pq_path_flow_paths, format = "parquet")

# 2. List partitioned parquet files
tibble(
  files = list.files(pq_path_flow_paths, recursive = TRUE),
  size_MB = file.size(file.path(pq_path_flow_paths, files)) / 1024^2
)

# 3. Use dplyr with arrow
flow_paths_pq <- open_dataset(pq_path_flow_paths)
query <- flow_paths_pq |> 
  mutate(
    waiting_time_sensitivity = as.numeric(waiting_time_sensitivity)
  ) |> 
  filter(
    # Filter by the particular scenario
    seed == representative_seed, 
    waiting_time_sensitivity %in% representative_waiting_time_sensitivity_levels,
    # Filter by the particular OD
    origin == representative_OD$origin,
    destination == representative_OD$dest,
  ) |> 
  group_by(episode, path, waiting_time_sensitivity) |> 
  summarise(
    flow = sum(count)
  ) |> 
  ungroup()

flow_paths_df_query <- query |> collect()

# Check: The sum of flows across paths for the particular OD for one episode should be equal to the demand on that OD
check_demand_od_df <- flow_paths_df_query |> 
  group_by(episode, waiting_time_sensitivity) |> 
  summarise(
    demand = sum(flow)
  ) |> ungroup()

demand_od_sum_paths <- unique(check_demand_od_df$demand)
demand_od <- representative_OD$demand

if (demand_od != demand_od_sum_paths) {
  stop("Something is wrong in the computation of path flows for the particular OD and scenario.")
}
```

```{r}
#| label: data_plot_3_part_3

# Now we still have to manipulate a bit the flow paths on the particular OD and scenario, so that is properly visualized in the plot

# 1. Handling some things to convert path variable to factor
path_levels <- as.character(sort(unique(flow_paths_df_query$path)))
path_levels_labels <- as.character(sort(unique(flow_paths_df_query$path + 1)))

# 2. Facet labels
facet_levels <- representative_waiting_time_sensitivity_levels
facet_labels = map_chr(
  facet_levels, 
  \(phi) ifelse(
    phi == 0, 
    glue("Waiting-time insensitive (\u03c6 = {phi})"), 
    glue("Waiting-time sensitive (\u03c6 = {phi})"))
)

flow_paths_df <- flow_paths_df_query |> 
  mutate(
    # Convert to factors
    waiting_time_sensitivity_fct = factor(
      waiting_time_sensitivity, 
      levels = facet_levels, 
      labels = facet_labels),
    path = factor(
      path, 
      levels = path_levels,
      labels = path_levels_labels
    ),
    share = flow / representative_OD$demand
  ) 
```

```{r}
#| label: breaks_plot3
#| eval: False

# Choose x-axis ticks
max_episode <- max(data$bm_rgap$episode)
step_plot4 <- floor(max_episode / 50) * 10
breaks_plot4 <- seq(0, max_episode, by = step_plot4)

# Remove breaks that are within step of the learning start
breaks_plot4 <- breaks[abs(breaks - exploration_phase_floor) >= step_plot4 / 2]
# Add the learning start break
breaks_plot4 <- sort(c(breaks_plot4, exploration_phase_floor))
labels_plot4 <- as.character(breaks_plot4)
labels_plot4[breaks_plot4==exploration_phase_floor] <- paste0("Learning\nstarts (", exploration_phase_floor, ")")
```

```{r}
#|label: sibedar_latex_table

# Based on median (more robust to outliers than mean)
typical_perceptions <- agents_last_episode_info |> 
  group_by(route_id) |> 
  summarise(
    median_PT = median(PT),
    median_WT = median(WT)
  ) |> 
  mutate(
  path_id = factor(route_id, levels = path_levels, labels = path_levels_labels)
  ) |> 
  select(-route_id) |> 
  ungroup()

share_paths_df_last_episode <- flow_paths_df |> 
  group_by(waiting_time_sensitivity) |> 
  filter(
    episode == max(episode)
  ) |> 
  rename(
    path_id = path
  ) |> 
  select(path_id, share, waiting_time_sensitivity) |> 
  arrange(waiting_time_sensitivity, path_id) |> 
  ungroup() |> 
  mutate(share = round(share, 2))

share_paths_wide <- share_paths_df_last_episode |> 
  pivot_wider(
    names_from = waiting_time_sensitivity,
    values_from = share,
    values_fill = 0
  )

latex_quantitative_table <- typical_perceptions |> 
  left_join(
  share_paths_wide,
  by = "path_id"
) 

write.table(latex_quantitative_table, file=here("r", "RQ12", "figures", "quantitative_table.txt"))
```

```{r}
#|label: Highlight_relevant_routes
relevant_routes <- latex_quantitative_table |> 
  mutate(
    is_shortest = median_PT == min(median_PT),
    is_wt_smallest = median_WT == min(median_WT),
    relevant = is_shortest != is_wt_smallest
  ) |> 
  filter(relevant) |> 
  pull(path_id) |> 
  as.numeric()

flow_paths_df_plot <- flow_paths_df |> 
  mutate(
    relevant = path %in% relevant_routes
  )

relevant_labels <- flow_paths_df_plot |>
  filter(relevant) |>
  group_by(waiting_time_sensitivity_fct, path) |>
  slice_max(episode, n = 1, with_ties = FALSE) |>
  ungroup()

# Colour assignment: black + #696969 for the two relevant routes, the
# remaining greyscale shades for the non-relevant background routes.
plot4_palette <- c(
  "#000000", "#111111", "#232323", "#343434",
  "#464646", "#575757", "#696969", "#7a7a7a"
)

all_path_labels <- levels(flow_paths_df_plot$path)
relevant_path_labels <- as.character(sort(relevant_routes))
nonrelevant_path_labels <- setdiff(all_path_labels, relevant_path_labels)

route_colors <- setNames(character(length(all_path_labels)), all_path_labels)
route_colors[relevant_path_labels[1]] <- "#000000"
route_colors[relevant_path_labels[2]] <- "#696969"
remaining_colors <- setdiff(plot4_palette, c("#000000", "#696969"))
route_colors[nonrelevant_path_labels] <- remaining_colors[seq_along(nonrelevant_path_labels)]
```

```{r}
#| label: plot3

# This plots refers to a particular OD, for a particular seed, and facet for some representative waiting-time sensitivities
ggplot(
  data = flow_paths_df_plot, 
  mapping = aes(x = episode, y = share, group = path, color = path)
) +
  # Shaded region
  annotate(
    "rect",
    xmin = -Inf,
    xmax = exploration_phase_floor,
    ymin = -Inf,
    ymax = Inf,
    fill = fill,
    alpha = alpha
  ) +
  geom_vline(
  xintercept = exploration_phase_floor,
  linetype = 2,
  alpha = 0.4,
  color = color,
  linewidth = 0.4
  ) +
  annotate(
    "text",
    x = (exploration_phase_floor - 2) / 2,
    y = Inf,
    label = "warm-up \nphase",
    fontface = "bold",
    colour = color,
    vjust = 2.5,
    size = 3.5
  ) +
  geom_line(
    mapping = aes(linewidth = relevant, alpha = relevant)
  ) +
   scale_linewidth_manual(
    values = c(
      "TRUE" = 1.3,
      "FALSE" = 0.4
    )
  ) + 
  scale_alpha_manual(
    values = c(
      "TRUE" = 1,
      "FALSE" = 0.4
    )
  ) + 
  facet_grid(waiting_time_sensitivity_fct ~ .) +
  labs(
    color = "Route",
    y = "Route flow",
    x = "Episode",
  ) +
  # Change color scale
  scale_color_manual(values = route_colors, name = "Path") +
  # Tweak miscellaneous stuff
  theme(
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10),
    legend.key.size = unit(0.6, "lines"),
    plot.subtitle = element_text(size = 11),
    strip.text.y = element_text(angle = 0, size = 10),
    panel.grid.major.y = element_line(
      color = "#E7E5DB",
      linewidth = 0.30
    ),
    panel.grid.minor.y = element_blank()
) +
  scale_y_continuous(labels = scales::percent_format()) +
  guides(alpha = "none", linewidth = "none", color = "none") +
  labs(
    y = "Route share (%)"
  ) +
  ggrepel::geom_text_repel(
  data = relevant_labels,
  aes(
    x = episode,
    y = share,
    label = paste0("Path ", path),
    color = path
  ),
  inherit.aes = FALSE,
  direction = "y",
  hjust = 0,
  nudge_x = 2,
  size = 3.8,
  fontface = "bold",
  box.padding = 0.6,
  point.padding = 0.2,
  segment.color = NA,
  min.segment.length = Inf,
  seed = 123
) +
  labs(
    title = "Waiting-time sensitivity changes route-choice patterns",
    subtitle = "Route-share trajectories for a representative OD under two levels of waiting-time sensitivity."
  )

ggsave(here("r","RQ12","figures","plot3_qualitative_route_flow_evolution.png"), width = 12, height = 6)
```


# PLOT RELEVANT ROUTES SHAPES

```{r}
#| label: network_edge_shapes

# Non-internal <edge> elements each have exactly one <lane> child with a
# "shape" attribute (a straight polyline "x1,y1 x2,y2 ..."), which is the
# edge's physical geometry — reused both to draw route trajectories and as
# the light-gray full-network backdrop that gives them spatial context.

# Read XML of the network used for this run
network_stem <- data$agents_od$network |> unique()
network_filename <- paste0(network_stem, ".xml")
all_networks <- list.files(
  here("sumo", "net"),
  pattern = "\\.net\\.xml$",
  recursive = TRUE,
  full.names = TRUE
)
network_matches <- all_networks[basename(all_networks) == network_filename]
if (length(network_matches) != 1) {
  stop(glue(
    "Expected exactly one network file named '{network_filename}' under sumo/net/, ",
    "found {length(network_matches)}."
  ))
}
network_path <- network_matches

# Function to parse the network XML edge shapes attributes.
# In particular:
# 1. SUMO shape attribute string has two points "x,y" separated by an space. 
# Example: 
# "107.20,-301.60 192.80,-301.60"
#    x1     y1      x2      y2
# 2. We wanna convert it to a df:
# | x      | y       |
# |--------|---------|
# | 107.20 | -301.60 |
# | 192.80 | -301.60 |

########
# Steps:
########
# 1. str_split(shape_str, " ")[[1]]: Splits on spaces. The [[1]] is because this function can accept more than one input string, and it returns a list with as much elements as input strings. So, with [[1]] we are saying that we want the results of separating by spaces the first (and only) string we passed as input.
# Result: c("107.20,-301.60", "192.80,-301.60")

# 2. str_split(","): Separate by comma
# Result: list(c("107.20", "-301.60"), c("192.80", "-301.60"))

# 3. map_chr(points,1): Walks the list and pulls out element 1 (x-string) from every point.
# Result: c("107.20", "192.80")
parse_shape <- function(shape_str) {
  points <- str_split(shape_str, " ")[[1]] |> str_split(",")
  tibble(
    x = as.numeric(map_chr(points, 1)),
    y = as.numeric(map_chr(points, 2))
  )
}

# Find all relevant edges elements of the network
network_edges <- xml_find_all(read_xml(network_path), "//edge[not(@function='internal')]")

# We create a tibble with two columns:
# 1. edge: (vectorized) For every edge it pulls out the id attribute giving one character vector — e.g. c("E10_11_EB", "E10_11_WB", ...)
# 2. shape_str: 
#   xml_find_first(): Find only first match. Basically for every edge, it pulls out the first lane child element. Because input were n edges, output is n lane elements as well.
# Then it pulls out the shape of the lane.
# So we end up with a tibble that contains the edges and its shapes,
# and its well aligned (since both columns are build from the same set of edges "network_edges" on the same order, is guarantee they are correctly aligned)
# Example:
# ┌───────────┬─────────────────────────────────┐
# │   edge    │            shape_str            │
# ├───────────┼─────────────────────────────────┤
# │ E10_11_EB │ "107.20,-301.60 192.80,-301.60" │
# └───────────┴─────────────────────────────────┘

# |> mutate(): Parses the shape of the edge as we saw before. Each cell is an entire small tibble containing the two points (x1,y1) and (x2,y2) for that edge
# Example:
# ┌───────────┬───────────────────────────────┬────────────────────────────────────────────────┐
# │   edge    │           shape_str           │                     shape                      │
# ├───────────┼───────────────────────────────┼────────────────────────────────────────────────┤
# │ E10_11_EB │ "107.20,-301.60               │ tibble(x=c(107.20,192.80),                     │
# │           │ 192.80,-301.60"               │ y=c(-301.60,-301.60))                          │
# └───────────┴───────────────────────────────┴────────────────────────────────────────────────┘


edge_shapes <- tibble(
  edge = xml_attr(network_edges, "id"),
  shape_str = xml_attr(xml_find_first(network_edges, "./lane[1]"), "shape")
) |>
  mutate(shape = map(shape_str, parse_shape)) |>
  select(-shape_str)

# We convert to long format
# Example one edge-row with a 2-point shape becomes 2 rows
# ┌───────────┬────────┬─────────┐
# │   edge    │   x    │    y    │
# ├───────────┼────────┼─────────┤
# │ E10_11_EB │ 107.20 │ -301.60 │
# ├───────────┼────────┼─────────┤
# │ E10_11_EB │ 192.80 │ -301.60 │
# └───────────┴────────┴─────────┘
network_backdrop <- edge_shapes |>
  unnest(shape)
```

```{r}
#| label: OD_routes

filter_relevant_routes <- relevant_routes-1

routes_agent <- data$od_routes |> 
  mutate(
    waiting_time_sensitivity = as.numeric(waiting_time_sensitivity)
  ) |> 
  filter(
    origin == representative_od$origin,
    dest == representative_od$destination,
    waiting_time_sensitivity == min(waiting_time_sensitivity),
    route_id %in% filter_relevant_routes
  ) |> 
  select(-c(warm_up, network, run_id, waiting_time_sensitivity))
```

```{r}
#| label: paths_factor
path_levels <- routes_agent$route_id |> unique()
path_levels_labels <- map_chr(path_levels, \(x) paste0("Path ", x+1))
```

```{r}
#| label: route_geometry

# Builds one ordered polyline per route by walking od_routes' edges in
# step order and concatenating each edge's shape points
build_route_geometry <- function(od_routes_df) {
  od_routes_df |>
    filter(
      origin == representative_od$origin,
      dest == representative_od$destination
    ) |> 
    arrange(route_id, step) |>
    left_join(edge_shapes, by = "edge") |>
    unnest(shape) |>
    select(route_id, edge, x, y)
}

route_geometry <- build_route_geometry(
  data$od_routes |> 
  mutate(
    waiting_time_sensitivity = as.numeric(waiting_time_sensitivity)
  ) |> 
  filter(
    waiting_time_sensitivity == min(waiting_time_sensitivity),
    route_id %in% filter_relevant_routes
  ) |> 
  select(-c(run_id, waiting_time_sensitivity, network, warm_up))
  ) |> 
  mutate(
    route_id = factor(route_id, levels = path_levels, labels = path_levels_labels)
  )
```

```{r}
#\ label: ManuallyModifyP2_shape
route_geometry <- route_geometry |>
  mutate(
    y = if_else(
      route_id == "Path 2" & row_number() > n() - 4,
      y + 3,
      y
    )
  )
```

```{r}
#| label: od_markers

# Origin/destination labels are placed at the midpoint of the origin/dest
# edge's shape (mean of its shape points — exact midpoint for the typical
# 2-point straight edge) rather than at its start/end vertex, so the label
# sits centered along the edge instead of right on top of a corner. Not
# faceted by algorithm (same OD for both), so ggplot repeats it on both panels.
roles_levels <- c("Origin", "Destination")
roles_labels <- c("Origin edge", "Destination edge")

od_markers <- bind_rows(
  Origin = edge_shapes |>
    filter(edge == representative_od$origin) |>
    unnest(shape) |>
    summarise(x = mean(x), y = mean(y)),
  Destination = edge_shapes |>
    filter(edge == representative_od$destination) |>
    unnest(shape) |>
    summarise(x = mean(x), y = mean(y)),
  .id = "role"
) |>
  mutate(role = factor(role, levels = roles_levels, labels = roles_labels))
```

```{r}
#| label: route_segments

# geom_path can't show direction of travel, and without it a route that
# doubles back (e.g. starts on a westbound edge before turning towards a
# destination to the east) reads as an arbitrary detour. Turning each
# route into (x, y) -> (xend, yend) segments lets us attach an arrowhead
# to specific edges instead of the whole line.
route_segments <- route_geometry |>
  group_by(route_id, edge) |>
  mutate(xend = lead(x), yend = lead(y)) |>
  ungroup() |>
  filter(!is.na(xend))

# Only the origin/destination edges get an arrowhead — enough to show
# initial/final direction of travel (e.g. why a route detours west first)
# without cluttering every intermediate edge along the route. Each row's
# `edge` is the edge its *start* point belongs to, so filtering on it picks
# out exactly that edge's own segment
route_endpoint_segments <- route_segments |>
  filter(edge %in% c(representative_od$origin)) |> 
  group_by(edge) |> 
  slice_head(n = 1) |> 
  select(-route_id)
```

```{r}
#| label: routes_plot

# Sidebar plot showing all 3 candidate routes since plot_agent_learning()
# only draws the chosen route per episode panel.
ggplot() +
    geom_path(
      data = network_backdrop,
      mapping = aes(x = x, y = y, group = edge),
      color = "grey85",
      linewidth = 0.3
    ) +
    geom_path(
      data = route_geometry,
      mapping = aes(
        x = x, y = y, color = route_id
      ),
      lineend = "round"
    ) +
    scale_colour_manual(
      values = c(
        `Path 1` = "#000000",
        `Path 2` = "#696969"
      )
    ) +
    # Preserve network aspect ratio
    coord_equal() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      panel.grid = element_blank(),
      legend.position = "none",
      panel.border = element_blank(),
    ) +
    geom_segment(
      data = route_endpoint_segments,
      mapping = aes(
        x = x, y = y, xend = xend, yend = yend,
      ),
      lineend = "round",
      arrow = arrow(length = unit(0.20, "cm"), type = "closed")
    ) +
    facet_wrap(~route_id, ncol = 2) +
    ggrepel::geom_label_repel(
      data = od_markers,
      mapping = aes(x = x, y = y, label = role),

      # label appearance
      size = 3,
      fontface = "bold",
      label.size = NA,
      fill = alpha("white", 0.8),

      # repel behavior
      box.padding = 0.5,
      nudge_y = 1.5,
      nudge_x = -0.5,

        # connector
      min.segment.length = 0,
      segment.size = 0.3,
      segment.color = "grey20",

      # reproducibility
      seed = 1,
    ) +
    labs(title = "Relevant routes for the considered OD pair")

ggsave(
  here("r","RQ12","figures","relevant_routes.png"),
  width = 10.7,
  height = 9.12,
  units = "in",
  dpi = 300
  )
```


