---
title: "Report"
author: "Miguel Angel Donado Fernandez"
date: "20/jun/2026"
format:
  html:
    self-contained: true
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

This R script answers the next research question:
"Question (RQ1): Analyze whether a Multi-Agent reinforcement
learning approach, like Bush-Mosteller, is capable of converging
towards a Dynamic User Equilibrium (DUE) state within a
microscopic traffic simulation environmnet"

**The plots to answer this question will be:**

1. Aggregate R-gap evolution (main convergence evidence).
2. Refined R-gap by departure interval in final episode (temporal heterogeneity).
3. OD-specific R-gap boxplot in final episode (spatial heterogeneity).
4. Maximum OD-specific R-gap evolution (robustness of convergence).

```{r}
#| label: setup
library(tidyverse)
library(arrow)
```

```{r}
#| label: load_and_set_custom_theme_

# Load custom theme
source(file = "r/theme.R")

# Make it global theme (every plot automatically uses it)
theme_set(theme_thesis())
```


```{r}
#| label: paths

base_dir <- "/home/miguel/6.Projects/Thesis/data/DUE"

algorithms <- c("BM", "duaIterate")

paths <- map(algorithms, \(alg) {
  list(
    rgap = file.path(base_dir, alg, "R-gap", "rgap.parquet"),
    rgap_od = file.path(base_dir, alg, "R-gap", "rgap_by_od.parquet"),
    refined_rgap = file.path(base_dir, alg, "R-gap", "refined_rgap.parquet"),
    refined_rgap_od = file.path(base_dir, alg, "R-gap", "refined_rgap_by_od.parquet")
  )}
)

names(paths) <- algorithms
```

```{r}
#| label: data

data <- map(paths, \(x) map(x, read_parquet))
```


```{r}
#| label: data_plot1

# Plot1: R-gap evolution across episode for the learning algorithms (general rgap)
# map: For each element of the list, extract the rgap df
# imap: For each dataframe, get the df and the name of the algorithm
# Outcome_df = episode | rgap | algorithm
# list_rbind: Combine dataframes of the different algorithms

rgap_df <- data |> 
  map("rgap") |> 
  imap(\(x,name) mutate(x, algorithm = name)) |> 
  list_rbind()

rgap_df <- rgap_df |> 
  filter(algorithm!="duaIterate")
```

```{r}
#| label: plot1

# scales = "free_y": Each facet get its own y-axis

ggplot(
  data = rgap_df,
  mapping = aes(
    x = episode,
    y = rgap,
  )
) +
  geom_line() + 
  facet_wrap(
    ~algorithm, scales = "free_y"
  ) +
  labs(
    title = "R-gap evolution across episodes",
    x = "Episode",
    y = "R-gap"
  )
```

```{r}
#| label: data_plot2

# Plot2: Refined r-gap (by interval) of the last episode for the learning algorithms

refined_rgap_df <- data |> 
  map("refined_rgap") |> 
  imap(\(x,name) mutate(x, algorithm = name)) |> 
  list_rbind()

refined_rgap_df_last_ep <- refined_rgap_df |> 
  group_by(algorithm) |> 
  filter(episode == max(episode)) |> 
  ungroup()
```

```{r}
#| label: plot2

### OPTION A

ggplot(
  data = refined_rgap_df_last_ep,
  mapping = aes(
    x = time_interval,
    y = refined_rgap,
    color = algorithm
  )
) +
  geom_line() +
  geom_point() +
  labs(
    title = "Refined R-gap by Departure Interval",
    subtitle = "Comparison between algorithms in the final episode",
    x = "Departure Time Interval",
    y = "Refined R-gap"
  )

### OPTION B

ggplot(
  refined_rgap_df_last_ep,
  aes(
    x = factor(time_interval),
    y = refined_rgap,
    fill = algorithm
  )
) +
  geom_col(position = "dodge") +
  labs(
    title = "Refined R-gap by Departure Interval",
    subtitle = "Comparison between algorithms in the final episode",
    x = "Departure Time Interval",
    y = "Refined R-gap"
  )
```

```{r}
#| label: data_plot3

# Analyze OD-specific R-gap (distribution)

rgap_od_df <- data |> 
  map("rgap_od") |> 
  imap(\(x,name) mutate(x, algorithm = name)) |> 
  list_rbind()

rgap_od_last_ep_df <- rgap_od_df |> 
  group_by(algorithm) |> 
  filter(episode == max(episode))
```

```{r}
#| label: plot3

# Option A (histogram)
ggplot(
  data = rgap_od_last_ep_df,
  mapping = aes(
    x = rgap,
    fill = algorithm
  )
) +
  geom_histogram() +
  labs(
    title = "Distribution of OD-Specific R-gaps",
    subtitle = "Comparison between algorithms in the final episode",
    x = "OD-Specific R-gap",
    y = "Number of OD Pairs"
  ) +
  scale_y_continuous(
    breaks = \(x) seq(0,max(x),by = 1)
  ) +
  scale_x_continuous(
    breaks = scales::breaks_pretty(n = 10)
  )

# Option B (density)
ggplot(
  data = rgap_od_last_ep_df,
  mapping = aes(
    x = rgap,
    fill = algorithm
  )
) +
  geom_density(alpha = 0.3) +
  labs(
    title = "Distribution of OD-Specific R-gaps",
    subtitle = "Comparison between algorithms in the final episode",
    x = "r-gap"
  )

# Option C
ggplot(
  data = rgap_od_last_ep_df,
  mapping = aes(
    x = algorithm,
    y = rgap,
    fill = algorithm
  ) 
) +
  geom_boxplot()
```

```{r}
#| label: data_plot4

# Visualize the maximum od-specific r-gap evolution

rgap_od_df_max <- 
  rgap_od_df |> 
  group_by(algorithm, episode) |> 
  summarise(max_rgap = max(abs(rgap))) |> filter(algorithm != "duaIterate")
```

```{r}
#| label: plot4

ggplot(
  data = rgap_od_df_max,
  mapping = aes(x = episode, y = max_rgap, color = algorithm)
) + 
  geom_line() +
  labs(
    title = "Evolution of maximum OD-Specific R-gap",
    subtitle = "Comparison between algorithms in the final episode",
    x = "Episode",
    y = "Max. od-specific rgap" 
  )
```

The file refined_rgap_by_od is not needed to be analyzed. A plot using this file would not add anything new.


